---
title: "Sample AKAP79 with ABC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample AKAP79 with ABC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

We use `icpm-kth/SBtabVFGEN` to read the model files and `icpm-kth/rgsl` as an
interface to `gsl_odeiv2` solvers (for initial value problems) to simulate the model:

```{r setup}
require(SBtabVFGEN)
require(rgsl)
library(uqsa)
require(parallel)
```

Next, we import the model's data tables from the SBtab files:

```{r read-model, eval=TRUE}
modelName <- checkModel("AKAP79",uqsa_example("AKAP79",pat="_gvf[.]c$"))
modelFiles <- uqsa_example(modelName,pat="[.]tsv$")
SBtab <- SBtabVFGEN::sbtab_from_tsv(modelFiles)
experiments <- SBtabVFGEN::sbtab.data(SBtab)
```

The function `checkModel` checks that the model files exist and
compiles the model to a shared library using a c compiler, if the
second argument is a c file.

```{r defaults, eval=TRUE}
parVal <- SBtabVFGEN::sbtab_quantity(SBtab$Parameter)
```

We have also defined a function that will transform the sampling
variables `parABC` used by the ABC method into something that the
model will accept as parameters. The plan is to sample in logarithmic
space: the ABC algorithm will sample the logarithms of the model
parameters. We limit the sampling to ranges, as appropriate for each
parameter:

```{r bounds, eval=TRUE}
ll <- log10(SBtab$Parameter[["!Min"]])
ul <- log10(SBtab$Parameter[["!Max"]])

set.seed(7619201)
```

The sampling procedure takes data in the list of `experiments` and
compares the data points to the solution that `gsl_odeiv2` returns. We
define a list of experiments, subdividing them into smaller groups and
processing the groups in sequence. Between the rounds of sampling the
posterior of every result is used as the prior distribution of the
next round. This mimics the arrival of data sets in sequence (from the
lab). The intermediate distributions will be modelled using
`VineCopula`.

```{r schedule, eval=TRUE}
chunks <- list(c(3, 12,18, 9), c(2, 11, 17, 8), c(1, 10, 16, 7))
```

Problem Size and core distribution:

```{r sample-size, eval=TRUE}
ns <- 300 # Size of the sub-sample from each chain
npc <- 70000 # pre-calibration sample size
```

ABC related settings:

```{r ABC, eval=TRUE}
delta <- 5

set.seed(7619201)

distanceMeasure <- function(funcSim, dataExpr=Inf, dataErr=Inf){
  if (all(is.finite(funcSim))){
    distance <- mean(((funcSim-t(dataExpr))/t(dataErr))^2, na.rm=TRUE)
  } else {
    distance <- Inf
  }
  return(distance)
}
```

Build a random number generator, and density function for the intial prior:
```{r initial-prior, eval=TRUE}
rprior <- rUniformPrior(ll, ul)
dprior <- dUniformPrior(ll, ul)
```

The sampling loop:

```{r sample, eval=TRUE}
nChains <- 4

start_time = Sys.time()
for (i in 1:length(chunks)){
	expInd <- chunks[[i]]
	simulate <- simulator.c(experiments,modelName,log10ParMap)
	objectiveFunction <- makeObjective(experiments[expInd], modelName, distanceMeasure, log10ParMap, simulate)
	options(mc.cores = length(expInd))
	pC <- preCalibration(objectiveFunction, npc, rprior,num=nChains)
	options(mc.cores = max(1,parallel::detectCores() %/% nChains))
	cl <- makeForkCluster(nChains)
	clusterExport(cl, c("objectiveFunction", "ns", "delta", "dprior"))
	out <- parLapply(cl, 1:nChains, function(j) ABCMCMC(objectiveFunction, pC$startPar[,j], ns, pC$Sigma, delta, dprior))
	stopCluster(cl)
	ABCMCMCoutput <- do.call(Map, c(rbind,out))
	if(i == 1){
		C <- fitCopula(ABCMCMCoutput$draws)
	}
	rprior <- rCopulaPrior(C)
	dprior <- dCopulaPrior(C)
}
end_time = Sys.time()
time_ = end_time - start_time

cat("Total time:\n",time_)
```

The sampling loop ran for ~24.5 minutes on a machine with the following specs:

| MacBook Pro | (13-inch, M1, 2020) |
|------------:|:--------------------|
|       Chip: | Apple M1            |
|     Memory: | 16 GB               |

The result is a collection of intermediate samples and a final posterior sample.

This article was build on this CPU:

```{sh}
grep "model name" /proc/cpuinfo | head -n 1
head -n 1 /proc/meminfo
```

Let's display the difference between the posterior and prior distribution:

```{r plot}
posterior <- ABCMCMCoutput$draws[,seq(6)]
prior <- rprior(NROW(posterior))[,seq(6)]
uqsa::showPosterior(posterior,prior)
```

Alternatively:

```{r hexbinplot}
if (require(hexbin)){
    hexbin::hexplom(posterior)
} else {
    pairs(posterior)
}
```
