R---
title: "Sample AKAP79 with ABC"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(parallel)
options(mc.cores=detectCores())
set.seed(7619201)
```

We import the model's data tables from the TSV files (one of the included examples):

```{r read-model}
m <- model_from_tsv(uqsa_example("AKAP79"))
cme <- as_cme(m)
c_path(cme) <- write_c_code(generate_code(cme,LV=5e7)) # a very small volume
so_path(cme) <- shlib(cme)
ex <- experiments(m)[seq(18)]
```

The function `shlib` compiles the model to a _shared library_ using
`R CMD SHLIB`. On a fairly normal system you can also make a shared
library using any c compiler with the options `cc -shared -fPIC ... `.

```{r defaults}
stopifnot(all(m$Parameter$scale == "log10"))
parABC <- values(m$Parameter) # initial value
```

The parameters are given in logarithmic space (log10). So, the natural
approach is to sample in logarithmic space: the ABC algorithm will
sample the logarithms of the model parameters. This way, the actual
parameters `10^parABC` are guaranteed to be positive.

The sampling procedure takes data in the list of experiments `ex` and
compares the data points to the solution that `gsl_odeiv2` returns.

```{r sample-size}
N <- 1000                    # sample size
```

Build a random number generator, and density function for the intial prior:

```{r initial-prior}
rprior <- rNormalPrior(values(m$Parameter), 0.5*m$Parameter$stdv)
dprior <- dNormalPrior(values(m$Parameter), 0.5*m$Parameter$stdv)
```

# ABC SMC for the CME model

We obtain a sample from tthe posterior:

```{r ABCSMC}
start_time = Sys.time()
X <- rprior(N) # sample from the prior

s <- simulator.c(ex,cme,log10ParMap)
Obj <- makeObjective(ex,s)
ABC_posterior <- ABCSMC(
	Obj,
	t(X),
	Sigma=cov(X),
	dprior=dprior,
	delta=c(1,30), # either a range or c(initial, final)
	messages=TRUE
)

Z <- ABC_posterior$draws

end_time = Sys.time()
time_span = end_time - start_time

cat("Total time:\n",time_span)
```

With this method the sample is practically uncorrelated:

```{r, out.width="100%"}
A <- acf(ABC_posterior$scores, plot=TRUE)
```

This article was build on this CPU:

```{sh}
grep "model name" /proc/cpuinfo | head -n 1
head -n 1 /proc/meminfo
```

Let's display the difference between the posterior and prior
distribution:

```{r plot}
posterior <- Z
prior <- rprior(NROW(posterior))
uqsa::showPosterior(posterior[,seq(6)],prior[,seq(6)])
```

Alternatively:

```{r hexbinplot}
if (require(hexbin)){
	hexbin::hexplom(posterior[,seq(6)])
} else {
	pairs(posterior[,seq(6)])
}
```

# Copula

We can estimate the posterior density function using copulas:

```{r copula}
C <- fitCopula(ABC_posterior$draws)
rprior <- rCopulaPrior(C)
dprior <- dCopulaPrior(C)

print(dprior(values(m$Parameter)))
```
