---
title: "Sample deterministic AKAR4 with ABC"
output:
  html_document:
    mathjax: "default"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(uqsa)
library(parallel)
```

We load the model from a collection of TSV files (`inst/extdata/AKAR4`
on github), the function below locates the tsv files after package
installation:

```{r, setup}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m)
C <- generate_code(o)
```

The model can exist in R and C form, in different files:
- `/path/to/AKAR4.R`
- `/path/to/AKAR4.c` or `AKAR4_gvf.c`

By default `generate_code` creates C code, it has a `language` option.

The C file contains functions used for simulations with GSL solvers --
odeiv2 module in the GNU Scientific Library.  The uqsa package has
interface functions that call GSL functions, but we don't include the
GSL library itself, ot has to be installed in the OS (system).

The functions in `.R` can be used more freely in R, and are
functionally identical (they describe the same model), but are
intended for use with deSolve, rather than our GSL interface
functions.

The C code needs to be written to a file and compiled to a shared
library. On a fairly normal system, you can make the shared library
manually, on the command line (if you want this).

In most cases, it is however better to let R make the shared library:

```{r, label="make shared library"}
c_path(o) <- write_c_code(C) # or cat
so_path(o) <- shlib(o)       # R CMD SHLIB
print(o)
```

The `o` variable (a list of ODE specific vectors), stores the path to
the different files. The C file can be inspected and modified before
making th eshared library. The code can also be written with
`cat(C,sep="\n",...)` to any file.

We want to sample in logarithmic space, so we set up a mapping
function `parMap`.

We distinguish between the Markov chain (sampling) variable, often
written as $\theta$, and the model's natural parameters. Any
relationship between the two can be encoded as a function:

$$
\text{parMap}: \theta -> \rho
$$

where $\rho$ are the parameters the ODE model expects.

The sampler will call this function on the sampling variable `parABC`
before it simulates the model (where `ABC` indicates the sampling
method we intend to use).

```{r, label="describe parameter space"}
parMap <- function (parABC=0) {
	return(10^parABC)
}
```

The ODE model receives a parameter vector `parMap(parABC)` for each
experiment, this way we can sample in logarithmic sapce. It consists
of the biological model parameters (which are the same for all
experiments), and input parameters (which together with initial
conditions distinguish the experimental setups). This model doesn't
have input parameters, so the biological and ODE parameters are the
same. Otherwise, two different kinds of parameters are concatenated
into one big numeric vector inside of the solver. The model function
`AKAR4_default()` returns the default values for `p`.

Next, we load the list of experiments from the same list of `data.frames` (SBtab content):
```{r, "load-experiments"}
ex <- experiments(m,o)
stopifnot(all(m$Parameter$scale == "linear"))
parVal <- values(m$Parameter)
print(parVal)
```

Define lower and upper limits for log-uniform prior distribution for
the parameters:

```{r limits}
ll <- log10(parVal) - 2
ul <- log10(parVal) + 2
```
We define a function that measures the distance between experiment and simulation:

```{r distance}
distanceMeasure <- function(funcSim, dataExpr, dataErr = NA){
  stopifnot(is.matrix(funcSim))
  stopifnot(is.matrix(dataExpr))
  j <- match("AKAR4pOUT",rownames(dataExpr))
  stopifnot(is.finite(j))
  stopifnot(1 <= j && j <= NROW(dataExpr))
  distance <- mean(((funcSim[j,]-dataExpr[j,])/max(dataExpr[j,],na.rm=TRUE))^2,na.rm=TRUE)
  return(distance)
}
```

If the data is very informative, then the difference between the prior
and posterior are large. In such cases, it may be difficult to find
the posterior through sampling. Dividing the data into chunks and
processing them sequentially can help. Our model of the posterior of
previous chunks is based on Copulas and the `VineCopula` package.

We divide the workload into chunks and loop over the chunks, iterating
the Bayesian problem via density estimation (VineCopula):

```{r ABC, collapse=FALSE}
## work chunks
chunks <- list(
	first=c(1,2), # then "fitCopula()"
	second=3
)

dprior <- dUniformPrior(ll, ul)
rprior <- rUniformPrior(ll, ul)

options(mc.cores=detectCores())

start_time = Sys.time()
for (i in seq(length(chunks))){
	simulate <- simulator.c(ex[chunks[[i]]],o,parMap=parMap,omit=3)
	Obj <- makeObjective(ex[chunks[[i]]],simulate,distance=distanceMeasure)
	prior <- rprior(10000)
	Sigma <- cov(prior)
	posterior <- ABCSMC(Obj,t(prior),dprior=dprior,Sigma=Sigma,delta=c(0.06,2))
	if (i < length(chunks)){
		COP <- fitCopula(posterior$draws)
		rprior <- rCopulaPrior(COP)
		dprior <- dCopulaPrior(COP)
	}
}
end_time = Sys.time()
time_ = end_time - start_time
print(time_)
```

We plot the sample as a two dimensional histogram plot-matrix using
the hexbin package:

```{r sample-density, out.width="100%", fig.width=12, fig.height=12, dpi=200}
colnames(posterior$draws)<-names(parVal)
#
if (require(hadron)){
    par(mfrow=c(3,1))
    ac <- hadron::uwerr(data=posterior$scores,pl=TRUE)
    tau <- ceiling(ac$tauint+ac$dtauint)
    i <- seq(1,NROW(posterior$draws),by=tau)
} else {
    A <- acf(posterior$scores,lag.max=100)
    l <- which(A$acf>0.2)
    tau <- sum(A$acf[l])
    i <- seq(1,NROW(posterior$draws),by=tau)
}

if (require(hexbin)){
    hexbin::hexplom(posterior$draws[i,],xbins=10)
} else {
    pairs(posterior$draws[i,])
}
```

A sensitivity plot using the results of the above loop:

```{r simulate}
y<-simulate(t(posterior$draws))
dim(y[[1]]$func)
```

where `y[[1]]` refers to the simulation corresponding to the first
experiment: `experiment[[1]]`, and `$func` refers to the output
function values (rather than state variables). The outout functions
are defined in the table `SBtab$Output`.


```{r plot, out.width="100%"}
f<-aperm(y[[1]]$func[1,,]) # aperm makes the sample-index (3rd) the first index of f, default permutation
S<-gsa_binning(posterior$draws,f)
S[1,]<-0 # the first index of S is time, and initially sensitivity is 0
cuS<-t(apply(S,1,cumsum))
plot.new()
tm<-ex[[3]]$outputTimes
plot(tm,cuS[,3],type="l",xlab="time",ylab="S",main="sensitivity (cumulative)")
## this section makes a little sensitivity plot:
for (si in dim(S)[2]:1){
	polygon(c(tm,rev(tm)),c(cuS[,si],numeric(length(tm))),col=si+1)
}
```
