---
title: "AKAR4"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(uqsa)
library(parallel)
library(SBtabVFGEN)
library(rgsl)
```

We load the model from a collection of TSV files (`inst/extdata/AKAR4`
on github), the function below locates the tsv files after package
installation:

```{r, setup}
modelFiles <- uqsa_example("AKAR4",full.names=TRUE)
```
Then we import the contents as a list of `data.frames`

```{r, import}
SBtab <- sbtab_from_tsv(modelFiles) # SBtabVFGEN
```
We load the R functions of the
model (e.g. the jacobian of the model):

- `AKAR4_vf`, the vector field,
- `AKAR4_jac`, the jacobian,
- `AKAR4_default`, the default parameters

The R file we source additionally includes a variable called `model`, which serves the purpose of having generic names for all model constituents, to make it easier to write generic scripts:

- `model$vf()` corresponds to `AKAR4_vf()`
    + but does nt use AKAR4 in the name
- `model$jac()` corresponds to `AKAR4_jac()`
- `model$name` is a string, `"AKAR4"` in this case

You can investigate the sourced file to find out more.

Generally, `model` is a list of functions, one of which is called
`model$par()` and returns the default parameters of the model in
linear space:

```{r, label="source the model files"}
source(uqsa_example("AKAR4",pat="^AKAR4[.]R$")) # loads "model"
model$name   # AKAR4
model$par()  # default values for parameters
model$init() # default initial values for ODE state variables
```

The model exists in R and C form:
- `AKAR4.R`
- `AKAR4_gvf.c`

The file ending in `_gvf.c` contains functions used for simulations
with `rgsl::r_gsl_odeiv2_outer()` (`_gvf.c` needs to be compiled). The
functions in `.R` can be used more freely in R, and are functionally
identical (they describe the same model, but solutions are slower in
R).

The next function call checks that the model `[...]_gvf.c` file exist,
and builds a shared library `.so` from these sources. It attaches a
comment to its return value that indicates how the shared library for
this model is named:

```{r, label="make shared library"}
modelName <- checkModel("AKAR4",uqsa_example("AKAR4",pat="_gvf[.]c$")) # SBtabVFGEN
comment(modelName) # should be AKAR4.so
```

The comment will be used to find the `.so` file (if you move that
file, adjust the comment on the `modelName`).

We want to sample in logarithmic space, so we set up a mapping
function `parMap`. The sampler will call this function on the sampling
variable `parABC` before it simulates the model.

```{r, label="describe parameter space"}
parMap <- function (parABC=0) {
	return(10^parABC)
}
```

The ODE model receives a parameter vector `p` for each experiment. It
consists of the biological model parameters (which are the same for
all experiments), and input parameters (which together with initial
conditions distinguish the experimental setups). This model deosn't
have input parameters, so the biological and ODE parameters are the
same. Otherwise, two different kinds of parameters are concatenated into
one big numeric vector inside of the solver. The model function
`AKAR4_default()` returns the default values for `p`.

Next, we load the list of experiments from the same list of `data.frames` (SBtab content):
```{r, "load-experiments"}
experiments <- sbtab.data(SBtab)
parVal <- AKAR4_default()
print(parVal)
```

Define lower and upper limits for log-uniform prior distribution for
the parameters:

```{r limits}
ll <- log10(parVal) - 3
ul <- log10(parVal) + 3
```

Define Number of Samples for the pre-calibration `npc` and each
ABC-MCMC chain `ns`. During ABC, we save every 100-th point, so `work(nChains*ns*100) ≈ work(npc)`:

```{r sample-size}
ns <- 800    # ABC-MCMC sample size
npc <- 50000 # pre-calibration size
```

```{r ABC-settings}
delta <- 0.02
set.seed(2022)
nCores <- parallel::detectCores()
options(mc.cores=nCores)
```

We define a function that measures the distance between experiment and simulation:

```{r distance}
distanceMeasure <- function(funcSim, dataExpr, dataErr = 1.0){
  distance <- mean(((funcSim-dataExpr$AKAR4pOUT)/max(dataExpr$AKAR4pOUT))^2,na.rm=TRUE)
  return(distance)
}
```

We divide the workload into chunks and loop over the chunks:

```{r ABC}
chunks <- list(c(1,2),3)
priorPDF <- dUniformPrior(ll, ul)
rprior <- rUniformPrior(ll, ul)

start_time = Sys.time()
for (i in seq(length(chunks))){
    expInd <- chunks[[i]]
    simulate <- simulator.c(experiments[expInd],modelName,parMap,noise=TRUE)
    Obj <- makeObjective(experiments[expInd],modelName,distanceMeasure,parMap,simulate)
    time_pC <- Sys.time()

    pC <- preCalibration(Obj, npc, rprior, rep=3)
    M <- getMCMCPar(pC$prePar, pC$preDelta, delta, num=1)
    time_pC <- Sys.time() - time_pC
    cat(sprintf("\t - time spent on precalibration: %g s\n",time_pC))

    time_ABC <- Sys.time()
    mcmc <- ABCMCMC(Obj, as.numeric(M$startPar), ns, M$Sigma, delta, priorPDF)
    time_ABC <- Sys.time() - time_ABC
    cat(sprintf("\t - time spent on ABC-MCMC: %g s\n",time_ABC))

    if (i>1){
        simulate <- simulator.c(experiments[chunks[[1]]],modelName,parMap)
        Obj <- makeObjective(experiments[chunks[[1]]],modelName,distanceMeasure,parMap,simulate)
        mcmc$draws <- checkFitWithPreviousExperiments(mcmc$draws, Obj, delta)
    }

    C <- fitCopula(mcmc$draws)
    priorPDF <- dCopulaPrior(C)
    rprior <- rCopulaPrior(C)
}
end_time = Sys.time()
time_ = end_time - start_time
print(time_)
```

We plot the sample as a two dimensional histogram plot-matrix using
the hexbin package:

```{r sample-density}
colnames(mcmc$draws)<-names(parVal)
hexbin::hexplom(mcmc$draws)
```

A sensitivity plot using the results of the above loop:

```{r simulate}
y<-simulate(t(mcmc$draws))
dim(y[[1]]$func)
```

where `y[[1]]` refers to the simulation corresponding to the first
experiment: `experiment[[1]]`, and `$func` refers to the output
function values (rather than state variables). The outout functions
are defined in the table `SBtab$Output`.


```{r plot}
f<-aperm(y[[1]]$func[1,,]) # aperm makes the sample-index (3rd) the first index of f, default permutation
S<-sensitivity(mcmc$draws,f)
S[1,]<-0 # the first index of S is time, and initially sensitivity is 0
cuS<-t(apply(S,1,cumsum))
plot.new()
tm<-experiments[[3]]$outputTimes
plot(tm,cuS[,3],type="l")
## this section makes a little sensitivity plot:
for (si in dim(S)[2]:1){
	polygon(c(tm,rev(tm)),c(cuS[,si],numeric(length(tm))),col=si+1)
}
```
