---
title: "Importing Data into R"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

In the previous article [Importing Models](./models.html) we have seen
the command `experiments(m,o)`. This commands reads the data from `m`,
but not only this, it also attempts to find simulation instructions
for `o`, to reproduce the sam effect using the model:

```{r, echo=FALSE, eval=TRUE}
library(uqsa)
library(errors)
```

```{r data}
library(uqsa)
library(errors)

## Read the TSV files, create an ODE model
m <- model_from_tsv(uqsa_example("AKAR4"))
o <- as_ode(m)                                 # the ODE

C <- generate_code(o)                          # list of strings
c_path(o) <- write_c_code(C)
so_path(o) <- shlib(o)
ex <- experiments(m,o)
```


The variable `ex` is a list of simulation experiments. If there are
only time-series experiments available in the files, the length of
`ex` will be the same as the number of rows in the table _Experiment_
(Experiment.tsv).

For convenience, we have implemented parsing for dose-response
experiments: experiments in which some condition (initial value, or
input) is varied between the data points, but for each data-point, a
fresh simulation is required. Experiments with `type` _dose response_
will be split into time-series simulation instructions, one line at a
time.

An Experiment Table can look like this (as a TSV):

```
id	t0	cAMP	type
X1	100	1.5	time-series
X2	100	3.0	time-series
X3	100	4.5	time-series
```

In this table, we update the default initial value of `cAMP`,
different for each experiment. the `ex` variable will have these
values in it:

```{r ex, eval=FALSE}
print(ex$X2$initialState['cAMP'])
# 3.0
```

Any reacting compound and input can be updated here. But, parameters
can not be updated: they are not considered experimental
conditions. We are concerned with inferring their values from data,
and thus shouldn't assign values to them during an experiment.

If you wish to omit a parameter from _uncertainty quantification_ or
inference, then move it into the list of inputs even if they aren't
set differently between the experiments.

## Normal Case

Our assumption is that the model has a defined list of possible
observables. They are described in the `Output.tsv` file. Each line
defines some mathematical expression of the model's state and any
other internal variable, e.g.: `A + AB + AC + ABC` for an observable that
measures all compounds that contain `A`, bound or otherwise.

In addition to this, we assume that the measurements, are related to
these output functions.

Let us assume that we have an output called `totalAmountOfA` (defined
like above). And the lab has measured this, but in arbitrary
units. This means that the measured value will not numerically agree
with the calculation above, the measurement is more like this:
`f*(A + AB + AC + ABC)`, with an unknown `f`. This is fine.

We only require that the measurement has something to do with the
Outputs. The `experiments()` function will find the measurement column
named `totalAmountOfA` and parse the values in that column to create a
data-matrix. The comparison between data and measurement happens in
the likelihood function. It is left to the user to write a likelihood
function that deals with the problem of unknown factors like `f`.

We only require a one-to-one relationship between data and output.

## Special Cases

If the situation is very special, and the data-sets have a different
dimensionality to the output, then a different treatment is required,
and `experiments` shouldn't be used probably.

Perhaps th eraw data can be processed enough so that it begins to
reflect some output function. But, if the user insists on having data
like: `Channel1_Y` and `Channel1_B`, but output functions like `PKA`,
then there is no way we can automatically find out which values in the
data belong to which of the outputs.

# Data Tables

The data must be given together with the uncertainty. Precisely like
the `errors` package does this:

```{r errors, eval=TRUE}
x <- seq(12)
errors(x) <- runif(12)
print(as.data.frame(x))
```

If you really don't want to use this notation, our parser also allows
this: `12±0.1` (same as `12.0(1)`), and the `±` symbol can also be
replaced with a semicolon `;` (and some other characters as well). The
function `uqsa::parse_concise()` parses these data entries, the
[experiments] function uses this parser:

```{r parse_concise, eval=TRUE}
z <- data.frame(
	data=parse_concise(
		c("12.0,0.1","12;0.1","12±0.1","12.0(1)")
	)
)
print(as.data.frame(z))
```

These are all equivalent, but only the last entry uses concise
parenthesised error notation that this function was written
for. Parsing a delimited entry with `;`, `±`, or `,` is much easier,
and happens a as a fall-back). *Note*: `+-` cannot be used instead of
`±`.

In all of this, it is only important that a value and its uncertainty
always travel together.


## Note on Aesthetics

The notation that the errors package uses is the best, shortest way to
write numbers with standard errors. The mass of the electron:

- $0.51099895069(16)\,\text{MeV}$ or
- $9.1093837139(28)\times 10^{−31}\,\text{kg}$

would look like this with the normal notation:
$(9.1093837139 \pm 0.0000000028) \times 10^{−31}\,\text{kg}$.  So, the
more precise your measurements are, the more appealing this notation
becomes.


### Aesthetics in Spreadsheets

It may look good to write this as a data table: value, separator, and
standard error in three different columns. The separator would go
entirely unparsed and is there for us human readers. This is what we
do for the reaction products and reactants. This table looks quite
clean:

  id time   AKAR4p\_m pm AKAR4p\_se
  -- ---- ----------- -- -------------
  t0  -10         200 +- 12
  t1   -5         234 +- 14
  ..  ...         ... .. ..
  tf  100        1031 +- 51

This table is also easy to plot with directly in
numbers/calc/gnumeric/google-sheets, as the values are separate from
the error-bars. So, we see the appeal.

But, once this table provides data for 5 different quantities, we need to find
the column of standard errors for each of the measured values according to
*some* rule. In addition, several different syntax rules for _column names_ may sound intuitive:

- `length_m`: length is in _meters_
- `length_m`: _mean_ value of length (the estimated value)
- `area_se`: _standard error_ of `area`
- `volume_sd`: _standard deviation_ of `volume`
- `ATP_u`: _uncertainty_ of `ATP`
- `ATP_v`: _value_ of `ATP`

Prefix/Suffix-Rules like that are used somewhere, I'm sure. And we
would like to avoid such rules.

Furthermore, several groupings of values and their uncertainties make
sense to us (see below).  So, we opted for not having special rules
for how to read the column names in this format, and keep the
measurement uncertainty very close to the values.

#### Error is next to value

  id A\_m A\_u B\_m B_u
  -- ---- ---- ---- ---
  D1    1 0.1     2 0.2
  D2    2 0.1     6 0.2
  D3    3 0.1     8 0.2

#### Values form a Block, Errors form a second Block

  id A\_m B\_m A\_u B_u
  -- ---- ---- ---- ---
  D1    1    2  0.1 0.2
  D2    2    6  0.1 0.2
  D3    3    8  0.1 0.2

## SBtab

In the [SBtabVGEN](https://github.com/icpm-kth/SBtabVFGEN) we actually
have a custom rule for this: data is marked with `>` and uncertainty
is marked with `~`, e.g.: `>AKAR4p` and `~AKAR4p`. The official SBtab
documentation doesn't commetn on this issue.

SBtab is an alternative format that we can use. The SBtabVFGEN package
provides the translation layer to import from that model type.
