---
title: "Transformation Events"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SBtabVFGEN)
library(rgsl)
library(uqsa)
```

The solver we use in most of our examples is
`rgsl::r_gsl_odeiv2_outer` which has the built-in ability to perform
sudden interventions. Here we show an example of how this can be used to trigger sudden spikes (action potentials)

```{r load model}
f <- uqsa_example('Spike')
sb <- sbtab_from_tsv(f)
ex <- sbtab.data(sb)
```

The transformations are still present in `sb`:

```{r}
sb$Transformation
```

But, also built into the model and simulation instructions `ex`

```{r}
ex[[1]]$events
```


# Simulation

```{r more-time-points}
cfile <- uqsa_example("Spike",pat='_gvf[.]c$')
modelName <- checkModel("Spike",cfile)
t <- seq(-4,1000)

ex[[1]]$outputTimes <- t
SIM <- simcf(ex,modelName)
file.exists(comment(modelName))
par <- sb$Parameter[["!DefaultValue"]]
print(par)
res <- SIM(par)
y <- as.numeric(res[[1]]$func)

plot(t,y,bty="n",type='l',xlab="t",ylab="Ca")
```


Maybe, we should also simulate how one Ca spike would look like, and also try reducing the dose a bit:

```{r just-one-spike}
ex[[1]]$event$time <- 0.0
ex[[1]]$event$label <- 0
ex[[1]]$event$dose <- 1.0

ex[[2]] <- ex[[1]]
ex[[2]]$event$dose <- 0.6 

SIM <- simcf(ex,modelName)
par <- sb$Parameter[["!DefaultValue"]]
print(par)
res <- SIM(par)
y1 <- as.numeric(res[[1]]$func)
y2 <- as.numeric(res[[2]]$func)

plot(t,y1,bty="n",type='l',xlab="t",ylab="Ca")
lines(t,y2,lty=2)
```
