---
title: "Transformation Events"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

The solvers we use in most of our examples have the built-in ability
to perform sudden interventions. Here we show an example of how this
can be used to trigger sudden spikes (action potentials) in a
neuroscience context.

```{r load-model}
f <- uqsa_example('Spike')
m <- model_from_tsv(f)
o <- as_ode(m,cla=FALSE) # no conservation law analysis
ex <- experiments(m,o)

C <- generate_code(o)
c_path(o) <- write_c_code(C)
so_path(o) <- shlib(o)
```

The transformations are still present in `m`:

```{r}
m$Transformation
```

But, also built into the model and simulation instructions `ex`

```{r}
ex[[1]]$events
```

Here, an event has a _label_, which is just the c-style offset into
the Transformation table. Offset `0` specifies the first
transformation. Each line of the transformation table is a code block
inside of a C switch statement. So, the label also selects the case of
that switch statement, see further below.

Each event has a scalar `dose` variable, currently this lower-case
`dose` variable cannot be renamed. It specifies the intensity of the
transformation. If the transformation requires more than one parameter
they have to be defined explicitly as part of the model's state, or
input (`Input` table).

Because events can change Input values, we can also define an input
that becomes a sort of storage unit that can remember the value of a
variable at a given time. So, we could define an output function that
uses this time-spcificity of events to define more intricate return
values.

Events can change state variables, but also the
inputs. They can also depend on the time `t`.


# Simulation

```{r more-time-points}
tm <- seq(-4,1000)

ex[[1]]$outputTimes <- tm
SIM <- simulator.c(ex,o)

par <- values(m$Parameter)

print(par)
res <- SIM(par)
y <- as.numeric(res[[1]]$func)

plot(tm,y,bty="n",type='l',xlab="t",ylab="Ca")
```

# One Calcium Spike

Maybe, we should also simulate how one Ca spike would look like, and also try reducing the dose a bit:

```{r just-one-spike}
ex[[1]]$event$time <- 0.0
ex[[1]]$event$label <- 0
ex[[1]]$event$dose <- 1.0

ex[[2]] <- ex[[1]]
ex[[2]]$event$dose <- 0.6

SIM <- simulator.c(ex,o)
res <- SIM(par)
y1 <- as.numeric(res[[1]]$func)
y2 <- as.numeric(res[[2]]$func)

plot(tm,y1,bty="n",type='l',xlab="t",ylab="Ca")
lines(tm,y2,lty=2)
```

# Details

In this section, we provide details about the implementation of
scheduled events. This information could be useful in writing complex
simulation instructions. Some concepts from _systems biology_ cannot
be expressed in the model files as they have a very plain set of
rules. But with some knowledge of the event system, we can express
fairly intricate simulation instructions.

## Switch Statement

This specific model has only one available transformation, the only
valid label is `0` and corresponds to the `enum` value `APCa`.

```{c, eval=FALSE}
/* ...                                 */
enum eventLabel { APCa, numEvents };
/* ...                                 */
/* skipped functions                   */
/* ...                                 */
int Spike_event(double t, double *y_, double *p_, int EventLabel, double dose){
	if (!y_ || EventLabel<0) return numEvents;
	double k1 = p_[_k1];                                                /* [  0] */
	double k2 = p_[_k2];                                                /* [  1] */
	double A = p_[_A];                                                  /* [  2] */
	double CaBase = p_[_CaBase];                                        /* [  3] */
	double Ca = y_[_Ca];                                                /* [  0] */
	double Buffer = y_[_Buffer];                                        /* [  1] */
	double rf1 = Buffer;
	double rf2 = -k1*k2*Ca-(k1+k2)*Buffer;
	switch(EventLabel){
	case APCa:
		y_[_Buffer] = A*(k2-k1)*dose;
	break;
	}
	return GSL_SUCCESS;
}
```

The C cource defines enums for the event labels to make the code more
readable and identify the transformations (`case` labels) by name.

The transformation expression `A*(k2-k1)*dose` has access to all constants, parameters
(input or otherwise), and expressions such as reaction fluxes.

This function is called when an event is scheduled to occur. The
integrator is then reset as if a new integration were started from the
changed model state (this does not reset the input values). This is
different from triggered events, which happen when a condition is
met. Triggered events are not implemented but exist in other
packages/toolboxes/languages such as COPASI.

## Inputs as Read/Write Buffers

Systems Biology often requires fairly difficult to express
experimental setups. Relative Data (Experiment and Control) are
phrased in such a way that two simulations are performed for each data
(pair of) meaningful data. If the experiment measures $A$ and $\prime
A$ where $\prime A$ is the _control_, and their ratio $a$ is
meaningful, we should compare that ratio after simulating the
scenarios for `A` and then for `A'`.

This could be either two experiments in the Experiments table and a
custom likelihood function that knows to take the ratio. Or, a
simulation that receives to different parameter vectors (as a
two-column-matrix). And defines a transformation event that stores a
value in an input (which retains its value) for the next parameter
column.

The experiment list `ex` will not have its values changed of
course. This retention happens in an internal buffer that is
read/write accessible to the ODE solver. The final state of the
input-vector it uses is not returned in any way.

Each experiment in the list is assigned its own input buffer, which is
re-used when the next parameter vector is processed, this is actually
true for the input parameters and the model's kinetic parameters as
the ODE solver doesn't distinguish these two. The solver works with a
matrix of size $N\times P$ where $N$ is the number of experiments and
$P$ is the number of overall parameters. The parameters are modified
by enum label, so there is no need to know where in the array they are
stored.
