---
title: "Examples overview"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Example Models

A simple example using the AKAR4 model is available in the article [AKAR4](AKAR4.html). You can run it locally using

```{r, eval=FALSE}
library(uqsa)
source(uqsa_example("AKAR4",pat="^runABCMCMC_AKAR4[.]R$"))
```

A few larger (computationally expensive) examples: *AKAP79*, *CaMKII*, and the stochastic version of AKAR4 are available in the form of R scripts and can be run locally or (preferably) on a computer cluster. These are the examples that are included in the package

* [AKAR4](AKAR4.html)
* [AKAP79 ODE model](./uqsa-AKAP79.html)
* [CaMKII ODE model](https://github.com/icpm-kth/uqsa/blob/master/inst/extdata/CaMKII/runABCMCMC_CaMKII.R)
* [AKAR4 stochastic model](https://github.com/icpm-kth/uqsa/blob/master/inst/extdata/AKAR4/runABCMCMC_AKAR4_withStochasticStimulation.R)


For the ODE models we added measurement noise or used a acceptance probability in order to allow the use of ABC ([Alahmadi et al 2020](http://dx.doi.org/10.1098/rsos.191315)).
The demo R scrips can be found in the folder [`inst/extdata`](https://github.com/icpm-kth/uqsa/tree/master/inst/extdata).

The `uqsa::uqsa_example` function can locate the example folders after package installation.
