---
title: "Generate C Code outside of R"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=FALSE, cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(uqsa)
```

The github repository of
[SBtabVFGEN](https://www.github.com/icpm-kth/SBtabVFGEN) (systems
biology table vector field generator) also includes an Rscript called
`sbtab_to_vfgen`. This exists for convenience and ease of automation.

The model probably exists as a git repository
(or another version control system). Perhaps it uses github-actions or
something similar. 

When the model changes, it is convenient to also immediately change
the C source code.

This can be done directly from the command line, like this:

```{sh, label="calling sbtab_to_vfgen", eval=FALSE}
sbtab_to_vfgen *.tsv        # no   conservation law analysis
sbtab_to_vfgen --cla *.tsv  # with conservation law analysis
sbtab_to_vfgen *.ods *.xlsx # for an ods file, or xlsx file
```

Here, we didn't have to launch R explicitly; the commands still run in
R. If github actions are used, R continues to be a dependency in the build
environment.

See also [PATH variable](path.html) explanations, about how to call
scripts from any location.

This script creates the `.vf`, `.xml`, and
[`.mod`](https://www.neuron.yale.edu/neuron/static/py_doc/modelspec/programmatic/mechanisms/nmodl2.html)
files, as well as the optional `ConservationLaws.{h5,RDS}` (an hdf5
file, and an RDS file), with the conservation laws; just like from within R.

To generate the _c source code_ we use the `sh/ode.sh` script from
[RPN-derivative](icpm-kth/RPN-derivative):

```{sh, label="create c code", eval=FALSE}
ode --maxima -C myModel.vf > myModel_gvf.c
```

Compile:

```
gcc -shared -fPIC -O2 -o myModel.so myModel_gvf.c
```

The `_gvf` suffix is not mandatory, this is how `vfgen gsl` calls such files.

---

R code can be created as well (pick some combination of options):

```{sh, label="create R code", eval=FALSE}
ode [--yacas|--maxima] -R myModel.{zip,vf,tar.gz} > myModel.R
```

This does not need to be compiled, it is plain R code for [deSolve](https://cran.r-project.org/package=deSolve).

# Generated Code

The files generated in this manner only need to be re-created if you change the model
structurally. A change in initial conditions, or parameter values will
result in largely the same code, except for the function that returns
the default initial conditions and default parameter values.

Once you have `myModel_gvf.c` you can either convert it to a shared
library yourself, or use the `checkModel` function included in `uqsa`:

```{sh, label="shared-object, manually made", eval=FALSE}
gcc -shared -fPIC -O2 -o myModel.so myModel_gvf.c
```

The simulator will guess the name `./myModel.so` if you supply
`modelName="myModel"`, for other locations, attach the path to
`modelName` as a comment.

# GNU make

All of these can be automated further with a `Makefile`:

```{make, eval=FALSE}
.PHONY: all

model = myModel

all: $(model).so $(model).R


%.R: %.tar.gz
	ode -R --maxima $^ > $@

%_gvf.c: %.tar.gz
	ode -C --maxima $^ > $@

%.vf %.tar.gz %.zip: *.tsv
	sbtab_to_vfgen $^

%.so: %_gvf.c
	gcc -shared -fPIC -O2 -o $@ $^
```

Calling `make` with no arguments in the project directory will then re
build the model if anything has changed in the tsv files.

## for Open Document Spreadsheets

A very similar file, for models stored as `.ods` spreadsheets, and using the prefix function of GNU make:

```{make, eval=FALSE}
.PHONY: all

all: $(addprefix myModel, .so .R)


%.R: %.tar.gz
	ode -R --maxima $^ > $@

%_gvf.c: %.tar.gz
	ode -C --maxima $^ > $@

%.vf %.tar.gz: %.ods
	sbtab_to_vfgen $^

%.so: %_gvf.c
	gcc -shared -fPIC -O2 -o $@ $^
```

