---
title: "Generate Model Code"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

It is possible to generate C code for the model in R. The currently
available computer algebra system for this task is the
[Ryacas](https://cran.r-project.org/package=Ryacas) package that
interfaces between R and [yacas](http://www.yacas.org/) -- both need
to be installed.

The functions use the same files as the standalone POSIX shell implementation in RPN-derivative.

```{r, label="load the file", eval=TRUE}
 f <- uqsa::uqsa_example("AKAP79")    # paths to the TSV files
sb <- SBtabVFGEN::sbtab_from_tsv(f)   # loads the TSV files
```

```{r, include=FALSE}
cl <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

Create the ODE files:
```{r, eval=FALSE}
cl <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

Load the content of the ODE files into R
```{r, eval=TRUE}
ode <- uqsa::loadODE("AKAP79.tar.gz")
```

The first three steps above are unrelated to code generation. They just find, and load the biological model, then convert it to an ordinary differential equation (and create intermediate files, various formats).

The function `loadODE` merely performs `read.delim` and creates a list
of all ODE components. The `ode` variable is a list of `data.frames`.

```{r, eval=TRUE}
C <- uqsa::generateCode(ode)
```

This returns the generated code as a character vector, it can be viewed (and written to a file using the `cat` function):

```{r, eval=FALSE}
cat(C,sep="\n")
```

This command will print the code, use the `file=` argument of `cat` to print it into a file.
See Appendix for the printout.

# Appendix

```
#include <stdlib.h>
#include <math.h>
#include <string.h>
#include <gsl/gsl_errno.h>
#include <gsl/gsl_odeiv2.h>
#include <gsl/gsl_math.h.h>

/* Enums will be used for indexing purposes.   */
enum stateVariable { _RiiP, _RiiP_cAMP, _RiiP_C, _RiiP_C_cAMP, _C, _Rii_cAMP, _Rii_C_cAMP, _RiiP_CaN, _RiiP_cAMP_CaN, _AKAR4_C, _AKAR4p, numStateVar };
enum param { _kf_Rii_C__RiiP_C, _kf_RiiP_CxcAMP__RiiP_C_cAMP, _kf_RiiP_cAMPxC__RiiP_C_cAMP, _kb_RiiP_cAMPxC__RiiP_C_cAMP, _kb_RiiPXcAMP__RiiP_cAMP, _kf_RiiPXcAMP__RiiP_cAMP, _kf_RiiPxC__RiiP_C, _kb_RiiPxC__RiiP_C, _kf_cAMPxRii__Rii_cAMP, _kb_cAMPxRii__Rii_cAMP, _kf_Rii_CxcAMP__Rii_C_cAMP, _kb_Rii_CxcAMP__Rii_C_cAMP, _kf_RiixC__Rii_C, _kf_Rii_cAMPxC__Rii_C_cAMP, _kb_Rii_cAMPxC__Rii_C_cAMP, _kf_Rii_C_cAMP__RiiP_C_cAMP, _kb_RiixC__Rii_C, _AKAPoff_1, _AKAPoff_3, _AKAPon_1, _AKAPon_3, _kf_C_AKAR4, _kb_C_AKAR4, _kcat_AKARp, _kmOFF, _kmON, _KD_T, _b_AKAP, _AKAR4_ConservedConst, _CaN_ConservedConst, _Rii_C_ConservedConst, _cAMP_ConservedConst, _Rii_ConservedConst, numParam };
enum func { _AKAR4pOUT, numFunc };

/* The error codes indicate how many values a function returns.                             */
/* Each function expects the output buffer to be allocated with at least that many values   */
/* ODE vector field: y' = f(t,y;p)   */
int AKAP79_vf(double t, const double y_[], double *f_, void *par){
	double *p_=par;
	if (!y_ || !f__) return 11;
	double kf_Rii_C__RiiP_C = p_[_kf_Rii_C__RiiP_C];                    /* [  0] */
	double kf_RiiP_CxcAMP__RiiP_C_cAMP = p_[_kf_RiiP_CxcAMP__RiiP_C_cAMP];       /* [  1] */
	double kf_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kf_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  2] */
	double kb_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kb_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  3] */
	double kb_RiiPXcAMP__RiiP_cAMP = p_[_kb_RiiPXcAMP__RiiP_cAMP];       /* [  4] */
	double kf_RiiPXcAMP__RiiP_cAMP = p_[_kf_RiiPXcAMP__RiiP_cAMP];       /* [  5] */
	double kf_RiiPxC__RiiP_C = p_[_kf_RiiPxC__RiiP_C];                  /* [  6] */
	double kb_RiiPxC__RiiP_C = p_[_kb_RiiPxC__RiiP_C];                  /* [  7] */
	double kf_cAMPxRii__Rii_cAMP = p_[_kf_cAMPxRii__Rii_cAMP];          /* [  8] */
	double kb_cAMPxRii__Rii_cAMP = p_[_kb_cAMPxRii__Rii_cAMP];          /* [  9] */
	double kf_Rii_CxcAMP__Rii_C_cAMP = p_[_kf_Rii_CxcAMP__Rii_C_cAMP];       /* [ 10] */
	double kb_Rii_CxcAMP__Rii_C_cAMP = p_[_kb_Rii_CxcAMP__Rii_C_cAMP];       /* [ 11] */
	double kf_RiixC__Rii_C = p_[_kf_RiixC__Rii_C];                      /* [ 12] */
	double kf_Rii_cAMPxC__Rii_C_cAMP = p_[_kf_Rii_cAMPxC__Rii_C_cAMP];       /* [ 13] */
	double kb_Rii_cAMPxC__Rii_C_cAMP = p_[_kb_Rii_cAMPxC__Rii_C_cAMP];       /* [ 14] */
	double kf_Rii_C_cAMP__RiiP_C_cAMP = p_[_kf_Rii_C_cAMP__RiiP_C_cAMP];       /* [ 15] */
	double kb_RiixC__Rii_C = p_[_kb_RiixC__Rii_C];                      /* [ 16] */
	double AKAPoff_1 = p_[_AKAPoff_1];                                  /* [ 17] */
	double AKAPoff_3 = p_[_AKAPoff_3];                                  /* [ 18] */
	double AKAPon_1 = p_[_AKAPon_1];                                    /* [ 19] */
	double AKAPon_3 = p_[_AKAPon_3];                                    /* [ 20] */
	double kf_C_AKAR4 = p_[_kf_C_AKAR4];                                /* [ 21] */
	double kb_C_AKAR4 = p_[_kb_C_AKAR4];                                /* [ 22] */
	double kcat_AKARp = p_[_kcat_AKARp];                                /* [ 23] */
	double kmOFF = p_[_kmOFF];                                          /* [ 24] */
	double kmON = p_[_kmON];                                            /* [ 25] */
	double KD_T = p_[_KD_T];                                            /* [ 26] */
	double b_AKAP = p_[_b_AKAP];                                        /* [ 27] */
	double AKAR4_ConservedConst = p_[_AKAR4_ConservedConst];            /* [ 28] */
	double CaN_ConservedConst = p_[_CaN_ConservedConst];                /* [ 29] */
	double Rii_C_ConservedConst = p_[_Rii_C_ConservedConst];            /* [ 30] */
	double cAMP_ConservedConst = p_[_cAMP_ConservedConst];              /* [ 31] */
	double Rii_ConservedConst = p_[_Rii_ConservedConst];                /* [ 32] */
	double RiiP = y_[_RiiP];                                            /* [  0] */
	double RiiP_cAMP = y_[_RiiP_cAMP];                                  /* [  1] */
	double RiiP_C = y_[_RiiP_C];                                        /* [  2] */
	double RiiP_C_cAMP = y_[_RiiP_C_cAMP];                              /* [  3] */
	double C = y_[_C];                                                  /* [  4] */
	double Rii_cAMP = y_[_Rii_cAMP];                                    /* [  5] */
	double Rii_C_cAMP = y_[_Rii_C_cAMP];                                /* [  6] */
	double RiiP_CaN = y_[_RiiP_CaN];                                    /* [  7] */
	double RiiP_cAMP_CaN = y_[_RiiP_cAMP_CaN];                          /* [  8] */
	double AKAR4_C = y_[_AKAR4_C];                                      /* [  9] */
	double AKAR4p = y_[_AKAR4p];                                        /* [ 10] */
	double kf_RiiP_cAMP_CaN__CaNXRii_cAMP = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_RiiPxCaN__RiiP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiP_CaN__RiixCaN = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiPxCaN__RiiP_CaN = b_AKAP * ((kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP)/kmON ) + (1 - b_AKAP) * (kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP) / kmOFF;
	double kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP * ((kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmON) + (1 - b_AKAP) * (kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmOFF;
	double kb_RiiP_CxcAMP__RiiP_C_cAMP = kf_RiiP_CxcAMP__RiiP_C_cAMP * KD_T;
	double AKAR4 = (AKAR4_ConservedConst - (AKAR4_C+AKAR4p));
	double CaN = (CaN_ConservedConst - (RiiP_CaN+RiiP_cAMP_CaN));
	double Rii_C = (Rii_C_ConservedConst - (RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	double cAMP = (cAMP_ConservedConst - (RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	double Rii = (Rii_ConservedConst - (RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	double reaction_51 = kf_Rii_C__RiiP_C*Rii_C;
	double reaction_14 = kf_RiiPxC__RiiP_C*RiiP*C - kb_RiiPxC__RiiP_C*RiiP_C;
	double reaction_12 = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C*cAMP - kb_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_43 = kf_RiiPXcAMP__RiiP_cAMP*cAMP*RiiP - kb_RiiPXcAMP__RiiP_cAMP*RiiP_cAMP;
	double reaction_23 = kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP*C - kb_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_78 = kf_cAMPxRii__Rii_cAMP*cAMP*Rii - kb_cAMPxRii__Rii_cAMP*Rii_cAMP;
	double reaction_56 = kf_Rii_CxcAMP__Rii_C_cAMP*Rii_C*cAMP - kb_Rii_CxcAMP__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_76 = kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP*C - kb_Rii_cAMPxC__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_62 = kf_Rii_C_cAMP__RiiP_C_cAMP*Rii_C_cAMP;
	double reaction_58 = kf_RiixC__Rii_C*Rii*C - kb_RiixC__Rii_C*Rii_C;
	double reaction_44 = kf_RiiPxCaN__RiiP_CaN*RiiP*CaN - kb_RiiPxCaN__RiiP_CaN*RiiP_CaN;
	double reaction_33 = kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN*CaN*RiiP_cAMP - kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN*RiiP_cAMP_CaN;
	double reaction_48 = kf_RiiP_CaN__RiixCaN*RiiP_CaN;
	double reaction_37 = kf_RiiP_cAMP_CaN__CaNXRii_cAMP*RiiP_cAMP_CaN;
	double reaction_1 = kf_C_AKAR4*C*AKAR4 - kb_C_AKAR4*AKAR4_C;
	double reaction_2 = kcat_AKARp*AKAR4_C;
	memset(f_,0,sizeof(double)*11); 
	f_[_RiiP] = -reaction_14-reaction_43-reaction_44;
	f_[_RiiP_cAMP] = +reaction_43-reaction_23-reaction_33;
	f_[_RiiP_C] = +reaction_51+reaction_14-reaction_12;
	f_[_RiiP_C_cAMP] = +reaction_12+reaction_23+reaction_62;
	f_[_C] = -reaction_14-reaction_23-reaction_76-reaction_58-reaction_1+reaction_2;
	f_[_Rii_cAMP] = +reaction_78-reaction_76+reaction_37;
	f_[_Rii_C_cAMP] = +reaction_56+reaction_76-reaction_62;
	f_[_RiiP_CaN] = +reaction_44-reaction_48;
	f_[_RiiP_cAMP_CaN] = +reaction_33-reaction_37;
	f_[_AKAR4_C] = +reaction_1-reaction_2;
	f_[_AKAR4p] = +reaction_2;
	return GSL_SUCCESS;
}

/* ODE Jacobian: df(t,y;p)/dy   */
int AKAP79_jac(double t, const double y_[], double *jac_, double *dfdt_, void *par){
	double *p_=par;
	if (!y_ || !jac__) return 121;
	double kf_Rii_C__RiiP_C = p_[_kf_Rii_C__RiiP_C];                    /* [  0] */
	double kf_RiiP_CxcAMP__RiiP_C_cAMP = p_[_kf_RiiP_CxcAMP__RiiP_C_cAMP];       /* [  1] */
	double kf_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kf_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  2] */
	double kb_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kb_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  3] */
	double kb_RiiPXcAMP__RiiP_cAMP = p_[_kb_RiiPXcAMP__RiiP_cAMP];       /* [  4] */
	double kf_RiiPXcAMP__RiiP_cAMP = p_[_kf_RiiPXcAMP__RiiP_cAMP];       /* [  5] */
	double kf_RiiPxC__RiiP_C = p_[_kf_RiiPxC__RiiP_C];                  /* [  6] */
	double kb_RiiPxC__RiiP_C = p_[_kb_RiiPxC__RiiP_C];                  /* [  7] */
	double kf_cAMPxRii__Rii_cAMP = p_[_kf_cAMPxRii__Rii_cAMP];          /* [  8] */
	double kb_cAMPxRii__Rii_cAMP = p_[_kb_cAMPxRii__Rii_cAMP];          /* [  9] */
	double kf_Rii_CxcAMP__Rii_C_cAMP = p_[_kf_Rii_CxcAMP__Rii_C_cAMP];       /* [ 10] */
	double kb_Rii_CxcAMP__Rii_C_cAMP = p_[_kb_Rii_CxcAMP__Rii_C_cAMP];       /* [ 11] */
	double kf_RiixC__Rii_C = p_[_kf_RiixC__Rii_C];                      /* [ 12] */
	double kf_Rii_cAMPxC__Rii_C_cAMP = p_[_kf_Rii_cAMPxC__Rii_C_cAMP];       /* [ 13] */
	double kb_Rii_cAMPxC__Rii_C_cAMP = p_[_kb_Rii_cAMPxC__Rii_C_cAMP];       /* [ 14] */
	double kf_Rii_C_cAMP__RiiP_C_cAMP = p_[_kf_Rii_C_cAMP__RiiP_C_cAMP];       /* [ 15] */
	double kb_RiixC__Rii_C = p_[_kb_RiixC__Rii_C];                      /* [ 16] */
	double AKAPoff_1 = p_[_AKAPoff_1];                                  /* [ 17] */
	double AKAPoff_3 = p_[_AKAPoff_3];                                  /* [ 18] */
	double AKAPon_1 = p_[_AKAPon_1];                                    /* [ 19] */
	double AKAPon_3 = p_[_AKAPon_3];                                    /* [ 20] */
	double kf_C_AKAR4 = p_[_kf_C_AKAR4];                                /* [ 21] */
	double kb_C_AKAR4 = p_[_kb_C_AKAR4];                                /* [ 22] */
	double kcat_AKARp = p_[_kcat_AKARp];                                /* [ 23] */
	double kmOFF = p_[_kmOFF];                                          /* [ 24] */
	double kmON = p_[_kmON];                                            /* [ 25] */
	double KD_T = p_[_KD_T];                                            /* [ 26] */
	double b_AKAP = p_[_b_AKAP];                                        /* [ 27] */
	double AKAR4_ConservedConst = p_[_AKAR4_ConservedConst];            /* [ 28] */
	double CaN_ConservedConst = p_[_CaN_ConservedConst];                /* [ 29] */
	double Rii_C_ConservedConst = p_[_Rii_C_ConservedConst];            /* [ 30] */
	double cAMP_ConservedConst = p_[_cAMP_ConservedConst];              /* [ 31] */
	double Rii_ConservedConst = p_[_Rii_ConservedConst];                /* [ 32] */
	double RiiP = y_[_RiiP];                                            /* [  0] */
	double RiiP_cAMP = y_[_RiiP_cAMP];                                  /* [  1] */
	double RiiP_C = y_[_RiiP_C];                                        /* [  2] */
	double RiiP_C_cAMP = y_[_RiiP_C_cAMP];                              /* [  3] */
	double C = y_[_C];                                                  /* [  4] */
	double Rii_cAMP = y_[_Rii_cAMP];                                    /* [  5] */
	double Rii_C_cAMP = y_[_Rii_C_cAMP];                                /* [  6] */
	double RiiP_CaN = y_[_RiiP_CaN];                                    /* [  7] */
	double RiiP_cAMP_CaN = y_[_RiiP_cAMP_CaN];                          /* [  8] */
	double AKAR4_C = y_[_AKAR4_C];                                      /* [  9] */
	double AKAR4p = y_[_AKAR4p];                                        /* [ 10] */
	double kf_RiiP_cAMP_CaN__CaNXRii_cAMP = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_RiiPxCaN__RiiP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiP_CaN__RiixCaN = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiPxCaN__RiiP_CaN = b_AKAP * ((kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP)/kmON ) + (1 - b_AKAP) * (kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP) / kmOFF;
	double kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP * ((kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmON) + (1 - b_AKAP) * (kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmOFF;
	double kb_RiiP_CxcAMP__RiiP_C_cAMP = kf_RiiP_CxcAMP__RiiP_C_cAMP * KD_T;
	double AKAR4 = (AKAR4_ConservedConst - (AKAR4_C+AKAR4p));
	double CaN = (CaN_ConservedConst - (RiiP_CaN+RiiP_cAMP_CaN));
	double Rii_C = (Rii_C_ConservedConst - (RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	double cAMP = (cAMP_ConservedConst - (RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	double Rii = (Rii_ConservedConst - (RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	double reaction_51 = kf_Rii_C__RiiP_C*Rii_C;
	double reaction_14 = kf_RiiPxC__RiiP_C*RiiP*C - kb_RiiPxC__RiiP_C*RiiP_C;
	double reaction_12 = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C*cAMP - kb_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_43 = kf_RiiPXcAMP__RiiP_cAMP*cAMP*RiiP - kb_RiiPXcAMP__RiiP_cAMP*RiiP_cAMP;
	double reaction_23 = kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP*C - kb_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_78 = kf_cAMPxRii__Rii_cAMP*cAMP*Rii - kb_cAMPxRii__Rii_cAMP*Rii_cAMP;
	double reaction_56 = kf_Rii_CxcAMP__Rii_C_cAMP*Rii_C*cAMP - kb_Rii_CxcAMP__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_76 = kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP*C - kb_Rii_cAMPxC__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_62 = kf_Rii_C_cAMP__RiiP_C_cAMP*Rii_C_cAMP;
	double reaction_58 = kf_RiixC__Rii_C*Rii*C - kb_RiixC__Rii_C*Rii_C;
	double reaction_44 = kf_RiiPxCaN__RiiP_CaN*RiiP*CaN - kb_RiiPxCaN__RiiP_CaN*RiiP_CaN;
	double reaction_33 = kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN*CaN*RiiP_cAMP - kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN*RiiP_cAMP_CaN;
	double reaction_48 = kf_RiiP_CaN__RiixCaN*RiiP_CaN;
	double reaction_37 = kf_RiiP_cAMP_CaN__CaNXRii_cAMP*RiiP_cAMP_CaN;
	double reaction_1 = kf_C_AKAR4*C*AKAR4 - kb_C_AKAR4*AKAR4_C;
	double reaction_2 = kcat_AKARp*AKAR4_C;
	memset(jac_,0,sizeof(double)*121); 
	jac_[0] = -(kf_RiiPxC__RiiP_C*C+kf_RiiPXcAMP__RiiP_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))+((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN)));
	jac_[1] = kf_RiiPXcAMP__RiiP_cAMP*RiiP+kb_RiiPXcAMP__RiiP_cAMP;
	jac_[2] = kb_RiiPxC__RiiP_C;
	jac_[3] = kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[4] = -kf_RiiPxC__RiiP_C*RiiP;
	jac_[5] = kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[6] = kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[7] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3;
	jac_[8] = kf_RiiPXcAMP__RiiP_cAMP*RiiP+((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP;
	jac_[11] = kf_RiiPXcAMP__RiiP_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[12] = -(kf_RiiPXcAMP__RiiP_cAMP*RiiP+kb_RiiPXcAMP__RiiP_cAMP+kf_RiiP_cAMPxC__RiiP_C_cAMP*C+((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN)));
	jac_[14] = kb_RiiP_cAMPxC__RiiP_C_cAMP-kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[15] = -kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP;
	jac_[16] = -kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[17] = -kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[18] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP;
	jac_[19] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3-kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jac_[22] = kf_RiiPxC__RiiP_C*C;
	jac_[23] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[24] = -(kb_RiiPxC__RiiP_C+kf_Rii_C__RiiP_C+kf_RiiP_CxcAMP__RiiP_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN)));
	jac_[25] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C+kf_RiiP_CxcAMP__RiiP_C_cAMP*KD_T-kf_Rii_C__RiiP_C;
	jac_[26] = kf_RiiPxC__RiiP_C*RiiP-kf_Rii_C__RiiP_C;
	jac_[27] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[28] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C-kf_Rii_C__RiiP_C;
	jac_[30] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[31] = -kf_Rii_C__RiiP_C;
	jac_[34] = kf_RiiP_cAMPxC__RiiP_C_cAMP*C-kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[35] = kf_RiiP_CxcAMP__RiiP_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[36] = -(kb_RiiP_cAMPxC__RiiP_C_cAMP+kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C+kf_RiiP_CxcAMP__RiiP_C_cAMP*KD_T);
	jac_[37] = kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP;
	jac_[38] = -kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[39] = kf_Rii_C_cAMP__RiiP_C_cAMP-kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[41] = -kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jac_[44] = kf_RiixC__Rii_C*C-kf_RiiPxC__RiiP_C*C;
	jac_[45] = kf_RiixC__Rii_C*C-kf_RiiP_cAMPxC__RiiP_C_cAMP*C;
	jac_[46] = kb_RiiPxC__RiiP_C-kb_RiixC__Rii_C;
	jac_[47] = kb_RiiP_cAMPxC__RiiP_C_cAMP-kb_RiixC__Rii_C;
	jac_[48] = -(kf_RiiPxC__RiiP_C*RiiP+kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP+kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP+kf_RiixC__Rii_C*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C))+kf_RiixC__Rii_C*C+kb_RiixC__Rii_C+kf_C_AKAR4*(AKAR4_ConservedConst-(AKAR4_C+AKAR4p)));
	jac_[49] = kf_RiixC__Rii_C*C-kf_Rii_cAMPxC__Rii_C_cAMP*C;
	jac_[50] = kb_Rii_cAMPxC__Rii_C_cAMP-kb_RiixC__Rii_C;
	jac_[51] = kf_RiixC__Rii_C*C;
	jac_[52] = kf_RiixC__Rii_C*C;
	jac_[53] = kf_C_AKAR4*C+kb_C_AKAR4-(kf_RiixC__Rii_C*C+kb_RiixC__Rii_C)+kcat_AKARp;
	jac_[54] = kf_C_AKAR4*C;
	jac_[55] = -kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[56] = -(kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C))+kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN)));
	jac_[58] = -kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	jac_[59] = kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))-kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP;
	jac_[60] = -(kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C))+kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))+kb_cAMPxRii__Rii_cAMP+kf_Rii_cAMPxC__Rii_C_cAMP*C);
	jac_[61] = kb_Rii_cAMPxC__Rii_C_cAMP-kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	jac_[62] = -kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[63] = b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1-(kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C))+kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN)));
	jac_[64] = kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[67] = -kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	jac_[68] = -kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[69] = -(kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))+kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C)));
	jac_[70] = kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP-kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[71] = kf_Rii_cAMPxC__Rii_C_cAMP*C-kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	jac_[72] = -(kb_Rii_cAMPxC__Rii_C_cAMP+kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))+kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C))+kb_Rii_CxcAMP__Rii_C_cAMP+kf_Rii_C_cAMP__RiiP_C_cAMP);
	jac_[74] = -kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	jac_[75] = -kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jac_[77] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jac_[84] = -(((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1);
	jac_[85] = -((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP;
	jac_[89] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jac_[95] = -((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP;
	jac_[96] = -(((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1);
	jac_[103] = kf_C_AKAR4*(AKAR4_ConservedConst-(AKAR4_C+AKAR4p));
	jac_[108] = -(kf_C_AKAR4*C+kb_C_AKAR4+kcat_AKARp);
	jac_[109] = -kf_C_AKAR4*C;
	jac_[119] = kcat_AKARp;
	return GSL_SUCCESS;
}

/* ODE parameter Jacobian: df(t,y;p)/dp   */
int AKAP79_jacp(double t, const double y_[], double *jacp_, double *dfdt_, void *par){
	double *p_=par;
	if (!y_ || !jacp__) return 363;
	double kf_Rii_C__RiiP_C = p_[_kf_Rii_C__RiiP_C];                    /* [  0] */
	double kf_RiiP_CxcAMP__RiiP_C_cAMP = p_[_kf_RiiP_CxcAMP__RiiP_C_cAMP];       /* [  1] */
	double kf_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kf_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  2] */
	double kb_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kb_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  3] */
	double kb_RiiPXcAMP__RiiP_cAMP = p_[_kb_RiiPXcAMP__RiiP_cAMP];       /* [  4] */
	double kf_RiiPXcAMP__RiiP_cAMP = p_[_kf_RiiPXcAMP__RiiP_cAMP];       /* [  5] */
	double kf_RiiPxC__RiiP_C = p_[_kf_RiiPxC__RiiP_C];                  /* [  6] */
	double kb_RiiPxC__RiiP_C = p_[_kb_RiiPxC__RiiP_C];                  /* [  7] */
	double kf_cAMPxRii__Rii_cAMP = p_[_kf_cAMPxRii__Rii_cAMP];          /* [  8] */
	double kb_cAMPxRii__Rii_cAMP = p_[_kb_cAMPxRii__Rii_cAMP];          /* [  9] */
	double kf_Rii_CxcAMP__Rii_C_cAMP = p_[_kf_Rii_CxcAMP__Rii_C_cAMP];       /* [ 10] */
	double kb_Rii_CxcAMP__Rii_C_cAMP = p_[_kb_Rii_CxcAMP__Rii_C_cAMP];       /* [ 11] */
	double kf_RiixC__Rii_C = p_[_kf_RiixC__Rii_C];                      /* [ 12] */
	double kf_Rii_cAMPxC__Rii_C_cAMP = p_[_kf_Rii_cAMPxC__Rii_C_cAMP];       /* [ 13] */
	double kb_Rii_cAMPxC__Rii_C_cAMP = p_[_kb_Rii_cAMPxC__Rii_C_cAMP];       /* [ 14] */
	double kf_Rii_C_cAMP__RiiP_C_cAMP = p_[_kf_Rii_C_cAMP__RiiP_C_cAMP];       /* [ 15] */
	double kb_RiixC__Rii_C = p_[_kb_RiixC__Rii_C];                      /* [ 16] */
	double AKAPoff_1 = p_[_AKAPoff_1];                                  /* [ 17] */
	double AKAPoff_3 = p_[_AKAPoff_3];                                  /* [ 18] */
	double AKAPon_1 = p_[_AKAPon_1];                                    /* [ 19] */
	double AKAPon_3 = p_[_AKAPon_3];                                    /* [ 20] */
	double kf_C_AKAR4 = p_[_kf_C_AKAR4];                                /* [ 21] */
	double kb_C_AKAR4 = p_[_kb_C_AKAR4];                                /* [ 22] */
	double kcat_AKARp = p_[_kcat_AKARp];                                /* [ 23] */
	double kmOFF = p_[_kmOFF];                                          /* [ 24] */
	double kmON = p_[_kmON];                                            /* [ 25] */
	double KD_T = p_[_KD_T];                                            /* [ 26] */
	double b_AKAP = p_[_b_AKAP];                                        /* [ 27] */
	double AKAR4_ConservedConst = p_[_AKAR4_ConservedConst];            /* [ 28] */
	double CaN_ConservedConst = p_[_CaN_ConservedConst];                /* [ 29] */
	double Rii_C_ConservedConst = p_[_Rii_C_ConservedConst];            /* [ 30] */
	double cAMP_ConservedConst = p_[_cAMP_ConservedConst];              /* [ 31] */
	double Rii_ConservedConst = p_[_Rii_ConservedConst];                /* [ 32] */
	double RiiP = y_[_RiiP];                                            /* [  0] */
	double RiiP_cAMP = y_[_RiiP_cAMP];                                  /* [  1] */
	double RiiP_C = y_[_RiiP_C];                                        /* [  2] */
	double RiiP_C_cAMP = y_[_RiiP_C_cAMP];                              /* [  3] */
	double C = y_[_C];                                                  /* [  4] */
	double Rii_cAMP = y_[_Rii_cAMP];                                    /* [  5] */
	double Rii_C_cAMP = y_[_Rii_C_cAMP];                                /* [  6] */
	double RiiP_CaN = y_[_RiiP_CaN];                                    /* [  7] */
	double RiiP_cAMP_CaN = y_[_RiiP_cAMP_CaN];                          /* [  8] */
	double AKAR4_C = y_[_AKAR4_C];                                      /* [  9] */
	double AKAR4p = y_[_AKAR4p];                                        /* [ 10] */
	double kf_RiiP_cAMP_CaN__CaNXRii_cAMP = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_RiiPxCaN__RiiP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiP_CaN__RiixCaN = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiPxCaN__RiiP_CaN = b_AKAP * ((kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP)/kmON ) + (1 - b_AKAP) * (kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP) / kmOFF;
	double kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP * ((kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmON) + (1 - b_AKAP) * (kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmOFF;
	double kb_RiiP_CxcAMP__RiiP_C_cAMP = kf_RiiP_CxcAMP__RiiP_C_cAMP * KD_T;
	double AKAR4 = (AKAR4_ConservedConst - (AKAR4_C+AKAR4p));
	double CaN = (CaN_ConservedConst - (RiiP_CaN+RiiP_cAMP_CaN));
	double Rii_C = (Rii_C_ConservedConst - (RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	double cAMP = (cAMP_ConservedConst - (RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	double Rii = (Rii_ConservedConst - (RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	double reaction_51 = kf_Rii_C__RiiP_C*Rii_C;
	double reaction_14 = kf_RiiPxC__RiiP_C*RiiP*C - kb_RiiPxC__RiiP_C*RiiP_C;
	double reaction_12 = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C*cAMP - kb_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_43 = kf_RiiPXcAMP__RiiP_cAMP*cAMP*RiiP - kb_RiiPXcAMP__RiiP_cAMP*RiiP_cAMP;
	double reaction_23 = kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP*C - kb_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_78 = kf_cAMPxRii__Rii_cAMP*cAMP*Rii - kb_cAMPxRii__Rii_cAMP*Rii_cAMP;
	double reaction_56 = kf_Rii_CxcAMP__Rii_C_cAMP*Rii_C*cAMP - kb_Rii_CxcAMP__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_76 = kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP*C - kb_Rii_cAMPxC__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_62 = kf_Rii_C_cAMP__RiiP_C_cAMP*Rii_C_cAMP;
	double reaction_58 = kf_RiixC__Rii_C*Rii*C - kb_RiixC__Rii_C*Rii_C;
	double reaction_44 = kf_RiiPxCaN__RiiP_CaN*RiiP*CaN - kb_RiiPxCaN__RiiP_CaN*RiiP_CaN;
	double reaction_33 = kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN*CaN*RiiP_cAMP - kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN*RiiP_cAMP_CaN;
	double reaction_48 = kf_RiiP_CaN__RiixCaN*RiiP_CaN;
	double reaction_37 = kf_RiiP_cAMP_CaN__CaNXRii_cAMP*RiiP_cAMP_CaN;
	double reaction_1 = kf_C_AKAR4*C*AKAR4 - kb_C_AKAR4*AKAR4_C;
	double reaction_2 = kcat_AKARp*AKAR4_C;
	memset(jacp_,0,sizeof(double)*363); 
	jacp_[4] = RiiP_cAMP;
	jacp_[5] = -(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))*RiiP;
	jacp_[6] = -RiiP*C;
	jacp_[7] = RiiP_C;
	jacp_[17] = -((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jacp_[18] = (1-b_AKAP)*RiiP_CaN-((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jacp_[19] = -(gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jacp_[20] = b_AKAP*RiiP_CaN-(gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jacp_[24] = ((CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP*(1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmOFF);
	jacp_[25] = ((CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP*b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmON);
	jacp_[27] = (AKAPon_3-AKAPoff_3)*RiiP_CaN-((b_AKAP*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1)/kmON+((1-b_AKAP)*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)-(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN));
	jacp_[29] = -((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP;
	jacp_[31] = -kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jacp_[35] = -RiiP_cAMP*C;
	jacp_[36] = RiiP_C_cAMP;
	jacp_[37] = -RiiP_cAMP;
	jacp_[38] = (cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))*RiiP;
	jacp_[50] = -((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP;
	jacp_[51] = (1-b_AKAP)*RiiP_cAMP_CaN-((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP;
	jacp_[52] = -(gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP;
	jacp_[53] = b_AKAP*RiiP_cAMP_CaN-(gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP;
	jacp_[57] = (RiiP_cAMP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*(1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmOFF);
	jacp_[58] = (RiiP_cAMP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmON);
	jacp_[60] = (AKAPon_3-AKAPoff_3)*RiiP_cAMP_CaN-((b_AKAP*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1)/kmON+((1-b_AKAP)*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)-(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP;
	jacp_[62] = -((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP;
	jacp_[64] = kf_RiiPXcAMP__RiiP_cAMP*RiiP;
	jacp_[66] = Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C);
	jacp_[67] = KD_T*RiiP_C_cAMP-RiiP_C*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jacp_[72] = RiiP*C;
	jacp_[73] = -RiiP_C;
	jacp_[92] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	jacp_[96] = kf_Rii_C__RiiP_C;
	jacp_[97] = -kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jacp_[100] = RiiP_C*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))-KD_T*RiiP_C_cAMP;
	jacp_[101] = RiiP_cAMP*C;
	jacp_[102] = -RiiP_C_cAMP;
	jacp_[114] = Rii_C_cAMP;
	jacp_[125] = -kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	jacp_[130] = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C;
	jacp_[134] = -RiiP_cAMP*C;
	jacp_[135] = RiiP_C_cAMP;
	jacp_[138] = -RiiP*C;
	jacp_[139] = RiiP_C;
	jacp_[144] = -(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C))*C;
	jacp_[145] = -Rii_cAMP*C;
	jacp_[146] = Rii_C_cAMP;
	jacp_[148] = Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C);
	jacp_[153] = -C*(AKAR4_ConservedConst-(AKAR4_C+AKAR4p));
	jacp_[154] = AKAR4_C;
	jacp_[155] = AKAR4_C;
	jacp_[160] = -kf_C_AKAR4*C;
	jacp_[162] = kb_RiixC__Rii_C;
	jacp_[164] = -kf_RiixC__Rii_C*C;
	jacp_[173] = (cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN))*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	jacp_[174] = -Rii_cAMP;
	jacp_[178] = -Rii_cAMP*C;
	jacp_[179] = Rii_C_cAMP;
	jacp_[182] = (1-b_AKAP)*RiiP_cAMP_CaN;
	jacp_[184] = b_AKAP*RiiP_cAMP_CaN;
	jacp_[192] = (AKAPon_1-AKAPoff_1)*RiiP_cAMP_CaN;
	jacp_[196] = kf_cAMPxRii__Rii_cAMP*(Rii_ConservedConst-(RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	jacp_[197] = kf_cAMPxRii__Rii_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jacp_[208] = (Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C))*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jacp_[209] = -Rii_C_cAMP;
	jacp_[211] = Rii_cAMP*C;
	jacp_[212] = -Rii_C_cAMP;
	jacp_[213] = -Rii_C_cAMP;
	jacp_[228] = kf_Rii_CxcAMP__Rii_C_cAMP*(cAMP_ConservedConst-(RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	jacp_[229] = kf_Rii_CxcAMP__Rii_C_cAMP*(Rii_C_ConservedConst-(RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	jacp_[248] = ((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))-(1-b_AKAP)*RiiP_CaN;
	jacp_[249] = ((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))-(1-b_AKAP)*RiiP_CaN;
	jacp_[250] = (gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))-b_AKAP*RiiP_CaN;
	jacp_[251] = (gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))-b_AKAP*RiiP_CaN;
	jacp_[255] = (-(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP*(1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmOFF);
	jacp_[256] = (-(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP*b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmON);
	jacp_[258] = ((b_AKAP*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1)/kmON+((1-b_AKAP)*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)-(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))-(AKAPon_3-AKAPoff_3)*RiiP_CaN-(AKAPon_1-AKAPoff_1)*RiiP_CaN;
	jacp_[260] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP;
	jacp_[281] = ((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP-(1-b_AKAP)*RiiP_cAMP_CaN;
	jacp_[282] = ((b_AKAP*(1-b_AKAP))/kmON+gsl_pow_2(1-b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP-(1-b_AKAP)*RiiP_cAMP_CaN;
	jacp_[283] = (gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP-b_AKAP*RiiP_cAMP_CaN;
	jacp_[284] = (gsl_pow_2(b_AKAP)/kmON+((1-b_AKAP)*b_AKAP)/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP-b_AKAP*RiiP_cAMP_CaN;
	jacp_[288] = (-RiiP_cAMP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*(1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmOFF);
	jacp_[289] = (-RiiP_cAMP*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/gsl_pow_2(kmON);
	jacp_[291] = ((b_AKAP*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)+b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1)/kmON+((1-b_AKAP)*(AKAPon_3-AKAPoff_3+AKAPon_1-AKAPoff_1)-(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*(CaN_ConservedConst-(RiiP_CaN+RiiP_cAMP_CaN))*RiiP_cAMP-(AKAPon_3-AKAPoff_3)*RiiP_cAMP_CaN-(AKAPon_1-AKAPoff_1)*RiiP_cAMP_CaN;
	jacp_[293] = ((b_AKAP*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmON+((1-b_AKAP)*(b_AKAP*AKAPon_3+(1-b_AKAP)*AKAPoff_3+b_AKAP*AKAPon_1+(1-b_AKAP)*AKAPoff_1))/kmOFF)*RiiP_cAMP;
	jacp_[318] = C*(AKAR4_ConservedConst-(AKAR4_C+AKAR4p));
	jacp_[319] = -AKAR4_C;
	jacp_[320] = -AKAR4_C;
	jacp_[325] = kf_C_AKAR4*C;
	jacp_[353] = AKAR4_C;
	return GSL_SUCCESS;
}

/* Output Function (Observables)   */
int AKAP79_func(double t, const double y_[], double *func_, void *par){
	double *p_=par;
	if (!y_ || !func__) return 1;
	double kf_Rii_C__RiiP_C = p_[_kf_Rii_C__RiiP_C];                    /* [  0] */
	double kf_RiiP_CxcAMP__RiiP_C_cAMP = p_[_kf_RiiP_CxcAMP__RiiP_C_cAMP];       /* [  1] */
	double kf_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kf_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  2] */
	double kb_RiiP_cAMPxC__RiiP_C_cAMP = p_[_kb_RiiP_cAMPxC__RiiP_C_cAMP];       /* [  3] */
	double kb_RiiPXcAMP__RiiP_cAMP = p_[_kb_RiiPXcAMP__RiiP_cAMP];       /* [  4] */
	double kf_RiiPXcAMP__RiiP_cAMP = p_[_kf_RiiPXcAMP__RiiP_cAMP];       /* [  5] */
	double kf_RiiPxC__RiiP_C = p_[_kf_RiiPxC__RiiP_C];                  /* [  6] */
	double kb_RiiPxC__RiiP_C = p_[_kb_RiiPxC__RiiP_C];                  /* [  7] */
	double kf_cAMPxRii__Rii_cAMP = p_[_kf_cAMPxRii__Rii_cAMP];          /* [  8] */
	double kb_cAMPxRii__Rii_cAMP = p_[_kb_cAMPxRii__Rii_cAMP];          /* [  9] */
	double kf_Rii_CxcAMP__Rii_C_cAMP = p_[_kf_Rii_CxcAMP__Rii_C_cAMP];       /* [ 10] */
	double kb_Rii_CxcAMP__Rii_C_cAMP = p_[_kb_Rii_CxcAMP__Rii_C_cAMP];       /* [ 11] */
	double kf_RiixC__Rii_C = p_[_kf_RiixC__Rii_C];                      /* [ 12] */
	double kf_Rii_cAMPxC__Rii_C_cAMP = p_[_kf_Rii_cAMPxC__Rii_C_cAMP];       /* [ 13] */
	double kb_Rii_cAMPxC__Rii_C_cAMP = p_[_kb_Rii_cAMPxC__Rii_C_cAMP];       /* [ 14] */
	double kf_Rii_C_cAMP__RiiP_C_cAMP = p_[_kf_Rii_C_cAMP__RiiP_C_cAMP];       /* [ 15] */
	double kb_RiixC__Rii_C = p_[_kb_RiixC__Rii_C];                      /* [ 16] */
	double AKAPoff_1 = p_[_AKAPoff_1];                                  /* [ 17] */
	double AKAPoff_3 = p_[_AKAPoff_3];                                  /* [ 18] */
	double AKAPon_1 = p_[_AKAPon_1];                                    /* [ 19] */
	double AKAPon_3 = p_[_AKAPon_3];                                    /* [ 20] */
	double kf_C_AKAR4 = p_[_kf_C_AKAR4];                                /* [ 21] */
	double kb_C_AKAR4 = p_[_kb_C_AKAR4];                                /* [ 22] */
	double kcat_AKARp = p_[_kcat_AKARp];                                /* [ 23] */
	double kmOFF = p_[_kmOFF];                                          /* [ 24] */
	double kmON = p_[_kmON];                                            /* [ 25] */
	double KD_T = p_[_KD_T];                                            /* [ 26] */
	double b_AKAP = p_[_b_AKAP];                                        /* [ 27] */
	double AKAR4_ConservedConst = p_[_AKAR4_ConservedConst];            /* [ 28] */
	double CaN_ConservedConst = p_[_CaN_ConservedConst];                /* [ 29] */
	double Rii_C_ConservedConst = p_[_Rii_C_ConservedConst];            /* [ 30] */
	double cAMP_ConservedConst = p_[_cAMP_ConservedConst];              /* [ 31] */
	double Rii_ConservedConst = p_[_Rii_ConservedConst];                /* [ 32] */
	double RiiP = y_[_RiiP];                                            /* [  0] */
	double RiiP_cAMP = y_[_RiiP_cAMP];                                  /* [  1] */
	double RiiP_C = y_[_RiiP_C];                                        /* [  2] */
	double RiiP_C_cAMP = y_[_RiiP_C_cAMP];                              /* [  3] */
	double C = y_[_C];                                                  /* [  4] */
	double Rii_cAMP = y_[_Rii_cAMP];                                    /* [  5] */
	double Rii_C_cAMP = y_[_Rii_C_cAMP];                                /* [  6] */
	double RiiP_CaN = y_[_RiiP_CaN];                                    /* [  7] */
	double RiiP_cAMP_CaN = y_[_RiiP_cAMP_CaN];                          /* [  8] */
	double AKAR4_C = y_[_AKAR4_C];                                      /* [  9] */
	double AKAR4p = y_[_AKAR4p];                                        /* [ 10] */
	double kf_RiiP_cAMP_CaN__CaNXRii_cAMP = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_RiiPxCaN__RiiP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiP_CaN__RiixCaN = b_AKAP * AKAPon_1 + (1 - b_AKAP) * AKAPoff_1;
	double kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP*AKAPon_3  +  (1 - b_AKAP)* AKAPoff_3;
	double kf_RiiPxCaN__RiiP_CaN = b_AKAP * ((kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP)/kmON ) + (1 - b_AKAP) * (kb_RiiPxCaN__RiiP_CaN + kf_RiiP_cAMP_CaN__CaNXRii_cAMP) / kmOFF;
	double kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN = b_AKAP * ((kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmON) + (1 - b_AKAP) * (kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN + kf_RiiP_CaN__RiixCaN)/kmOFF;
	double kb_RiiP_CxcAMP__RiiP_C_cAMP = kf_RiiP_CxcAMP__RiiP_C_cAMP * KD_T;
	double AKAR4 = (AKAR4_ConservedConst - (AKAR4_C+AKAR4p));
	double CaN = (CaN_ConservedConst - (RiiP_CaN+RiiP_cAMP_CaN));
	double Rii_C = (Rii_C_ConservedConst - (RiiP_C+RiiP_C_cAMP+C+Rii_C_cAMP+AKAR4_C));
	double cAMP = (cAMP_ConservedConst - (RiiP_cAMP+RiiP_C_cAMP+Rii_cAMP+Rii_C_cAMP+RiiP_cAMP_CaN));
	double Rii = (Rii_ConservedConst - (RiiP+RiiP_cAMP+Rii_cAMP+RiiP_CaN+RiiP_cAMP_CaN-C-AKAR4_C));
	double reaction_51 = kf_Rii_C__RiiP_C*Rii_C;
	double reaction_14 = kf_RiiPxC__RiiP_C*RiiP*C - kb_RiiPxC__RiiP_C*RiiP_C;
	double reaction_12 = kf_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C*cAMP - kb_RiiP_CxcAMP__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_43 = kf_RiiPXcAMP__RiiP_cAMP*cAMP*RiiP - kb_RiiPXcAMP__RiiP_cAMP*RiiP_cAMP;
	double reaction_23 = kf_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_cAMP*C - kb_RiiP_cAMPxC__RiiP_C_cAMP*RiiP_C_cAMP;
	double reaction_78 = kf_cAMPxRii__Rii_cAMP*cAMP*Rii - kb_cAMPxRii__Rii_cAMP*Rii_cAMP;
	double reaction_56 = kf_Rii_CxcAMP__Rii_C_cAMP*Rii_C*cAMP - kb_Rii_CxcAMP__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_76 = kf_Rii_cAMPxC__Rii_C_cAMP*Rii_cAMP*C - kb_Rii_cAMPxC__Rii_C_cAMP*Rii_C_cAMP;
	double reaction_62 = kf_Rii_C_cAMP__RiiP_C_cAMP*Rii_C_cAMP;
	double reaction_58 = kf_RiixC__Rii_C*Rii*C - kb_RiixC__Rii_C*Rii_C;
	double reaction_44 = kf_RiiPxCaN__RiiP_CaN*RiiP*CaN - kb_RiiPxCaN__RiiP_CaN*RiiP_CaN;
	double reaction_33 = kf_CaNxRiiP_cAMP__RiiP_cAMP_CaN*CaN*RiiP_cAMP - kb_CaNxRiiP_cAMP__RiiP_cAMP_CaN*RiiP_cAMP_CaN;
	double reaction_48 = kf_RiiP_CaN__RiixCaN*RiiP_CaN;
	double reaction_37 = kf_RiiP_cAMP_CaN__CaNXRii_cAMP*RiiP_cAMP_CaN;
	double reaction_1 = kf_C_AKAR4*C*AKAR4 - kb_C_AKAR4*AKAR4_C;
	double reaction_2 = kcat_AKARp*AKAR4_C;
	memset(func_,0,sizeof(double)*1); 
	func_[_AKAR4pOUT] = (AKAR4p*5)*71.67+100;
	return GSL_SUCCESS;
}
```
