---
title: "Generate Model Code"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

It is possible to generate C code for the model in R. The currently
available computer algebra system for this task is the
[Ryacas](https://cran.r-project.org/package=Ryacas) package that
interfaces between R and [yacas](http://www.yacas.org/) -- both need
to be installed.

The function `generateCode` accepts the return value of
`SBtabVFGEN::sbtab_to_vfgen()` (a list of ode properties) and returns
a character vector of C code:

```{r, label="load the file", eval=TRUE}
 f <- uqsa::uqsa_example("AKAP79")    # paths to the TSV files
sb <- SBtabVFGEN::sbtab_from_tsv(f)   # loads the TSV files
```

```{r, include=FALSE}
odeModel <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

Create the ODE files:
```{r, eval=FALSE}
odeModel <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

The first two steps above are unrelated to code generation. They just find, and load the biological model.
Then we convert it to an ordinary differential equation (and create intermediate files, various formats).

```{r, eval=TRUE}
C <- uqsa::generateCode(odeModel)
```

This returns the generated code as a character vector, it can be
viewed (and written to a file using the `cat` function). Here are a
few lines of the generated code:

```{r, eval=TRUE}
cat(head(C,25),sep="\n")
```

This command will print the code, use the `file=` argument of `cat` to print it into a file.
