---
title: "Global Sensitivity Analysis"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```
With UQSA you can perform global sensitivity analysis (GSA) to find out how the uncertainty of the model output depend on the uncertainty in the parameter estimates. 

UQSA computes *first order* and *total order* [Sobol sensitivity index](http://mi.mathnet.ru/mm2320) $S_i$ and $S_{iT}$ for **independent** parameter distributions using the method developed by [Saltelli, 2002](https://doi.org/10.1016/S0010-4655(02)00280-1) with help from [Halnes et al, 2009](https://doi.org/10.1007/s10827-009-0162-6) . 

We have also implemented a [method based on binning](https://doi.org/10.1093/bioinformatics/bty607) to calculate *first order* sensitivities $S_i$ also for **non-independent** distributions. This method can be applied directly on the distribution retrieved from an uncertainty quantification process.

In this [example](GSA_AKAR4.html), both methods for GSA are used and compared.

The Saltelli method requires that the parameter-matrices $M1,M2$ and $N$ is created. This can be done with the function `saltelli_prior()`. Next the corresponding model output has to be calculated (simulated). This gives corresponding matrices $FM1,FM2$ and $FN$. From $FM1,FM2$ and $FN$ the sobol indexes can next be calculated by the function `gsa_saltelli()`. For details see this [example](GSA_AKAR4.html).

The binning method can be used directly on a (in most cases non-independent) sample from an uncertainty quantification process. The function `gsa_binning()`takes the parameter and output sample as input and returns Sobols first order sensitivity indexes. For details see this [example](GSA_AKAR4.html).

 
 