---
title: "Global sensitivity analysis on AKAR4 - independent input factors"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(parallel)
```

This article provides code do perform global sensitivity analysis with the  Sobol-Saltelli method and the with the binning-approach.

Load the model files, create ODE model code and load examples similar to the example [Simulate the AKAR4 deterministic model](./simAKAR4.html).

```{r load}
f  <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m,cla=FALSE)
C <- generate_code(o)
c_path(o) <- write_c_code(C)
so_path(o) <- shlib(o)
ex <- experiments(m,o)
```

## Create simulator

We construct a simulator function and and test it on default parameter values.

```{r test}
options(mc.cores = length(ex))
s <- simulator.c(ex,o,parMap=log10ParMap,omit=3)
p <- log10(values(m$Parameter))
y <- s(p)
```

Set meta-parameters for the global sensitivity simulation

```{r}
nSamples <- 40000
```

nSamples corresponds to the number of samples in M1 (or M2) of the
Soboll-Saltelli approach. The number of simulations of the
Sobol-Saltelli approach consists of 2\*nSampels+nPars\*nSamples number
of simulations. In the binning approach below we use 2\*nSamples
number of samples (corresponding to 2\*nSamples number of simulations)
to use the same number of independent sample points as Sobol-Saltelli.


Construct parameter prior samples according to Sobol-Saltelli (M1, M2, N).

```{r}
rprior <- rNormalPrior(p, array(1, length(p)))
prior <- saltelli_prior(nSamples, rprior)
names(prior)
```

Plot parameter prior (M1)

```{r}
title <- paste("Prior distribution experiment");
boxplot(prior$M1, main = title, names=names(p), las=2, cex.main=0.9, cex.axis=0.3)
```

## Simulate from the prior

Set up simulator considering experiment 2

```{r}
E <- 2
T <- 5

fM1 <- t(s(t(prior$M1))[[E]]$state[,T,])
fM2 <- t(s(t(prior$M2))[[E]]$state[,T,])
fN <- array(
	NA,
	dim=c(nSamples, length(ex[[E]]$initialState), length(p))
)

for (i in seq_along(p)){
	fN[,,i] <- t(s(t(prior$N[,,i]))[[E]]$state[,T,])
}
```

## Calculate and plot sensitivity indexes

Calculate sensitivity indexes for sobol-saltelli
```{r}
SA <- gsa_saltelli(fM1,fM2,fN)
```

Plot first (SI) and total-order (SIT) sensitivity indexes for sobol-saltelli

```{r}
cols=rainbow(3)
par(mfrow = c(1, 2))
barplot(
	t(SA$SI[,]),
	col=cols,
	border="white",
	space=0.04,
	cex.axis=1,
	names.arg=names(ex[[E]]$initialState),
	cex.names = 0.7,
	las = 2,
	xlab="SI",
	las=2,
	cex.main=0.9
)
barplot(
	t(SA$SIT[,]),
	col=cols,
	border="white",
	space=0.04,
	cex.axis=1,
	names.arg=names(ex[[E]]$initialState),
	cex.names = 0.7,
	las = 2,
	xlab="SIT",
	legend.text=names(p),
	args.legend = list(x = "topright", inset=c(0, 0.1), cex=0.7)
)
```

Plot all states and timeponts for the experiment

```{r}
allTimesSample=s(t(prior$M2))[[E]]$state
par(mfrow = c(2, 2))
for (i in seq_along(ex[[E]]$initialState)) {
	matplot(
		ex[[E]]$outputTime,
		allTimesSample[i,,seq(500)],
		type = "l",
		lty = 1,
		col = c("red", "blue", "green"),
		ylab = "Y",
		xlab = names(ex[[E]]$initialState)[i]
	)
}
```


Calculate first order sensitivity index (SI) based on binning approach

```{r}
SIappr <- gsa_binning(rbind(prior$M1,prior$M2), rbind(fM1,fM2))
```

Plot SI for Sobol-Saltelli versus the binning approach


```{r}
par(mfrow = c(1, 3))
barplot(
	t(SA$SI[,]),
	col=cols,
	border="white",
	space=0.04,
	cex.axis=1,
	names.arg=names(ex[[E]]$initialState),
	cex.names = 0.7,
	las = 2,
	main="SI Saltelli")
barplot(
	t(SIappr),
	col=cols,
	border="white",
	space=0.04,
	cex.axis=1,
	cex.names = 0.7,
	las = 2,
	main="approximate SI")
barplot(
	c(0),
	axes=FALSE,
	col=cols,
	border="white",
	space=0.04,
	font.axis=2,
	legend.text=names(p)
)

mtext(paste("Comparision GSA methods, sample size=",as.character(2*nSamples)), side = 3, line = -1.2, outer = TRUE)
```
