---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisits

The simulators (ODE solvers) we use need the **GNU Scientific library**
[(GSL)](https://www.gnu.org/software/gsl/) to be installed in your system.

```{sh}
sudo apt install libgsl-dev # on Ubuntu
```

The model building part of UQSA (transferring the SBtab network model
to a mathematical model in C or R) also requires that you have a **C
compiler** (e.g. [gcc](https://gcc.gnu.org)).

```{sh}
sudo apt install gcc        # on Ubuntu
```

On some systems the GSL may be installed in an unexpected location (on
clusters this will almost certainly be the case). So, we rely on
[pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/) to
find this location.

```{sh}
sudo apt install pkg-config # on Ubuntu
```

You probably know the equivalent commands on your system.

The following Sections have instructions for several different
operating systems, for convenience.

## On macOS

To install libraries and science related software, people typically
opt for [homebrew](https://brew.sh/), or
[nix](https://nixos.org). Here is an example using homebrew:

```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, maybe agree to Xcode license?
brew install pkg-config gsl
```

MacOS does not have a default package manager for software of this
kind. There is probably no easier way of getting the above done on a
factory fresh device without first installing brew or nix.

It may be unnecessary to install gcc, if you already have `clang`.

## On GNU/Linux

Debian/Ubuntu derived:

```{sh ubuntu/debian, eval=FALSE}
sudo apt install pkg-config gcc libgsl-dev
```

Other distributions have some version of this:

```
sudo apk add gsl pkgconf            # Alpine
sudo guix install gsl pkg-config    # GNU Guix
sudo pacman -S gsl pkgconf          # Arch, btw.
sudo zypper install gsl pkg-config  # OpenSuSE, SLES
```

... and many [more](articles/installExplanations.html).

# Installation of the R packages

In order to use UQSA you (obviously) need the
[R language](https://www.r-project.org/)
(maybe [R-studio](https://posit.co/download/rstudio-desktop/)).
We will assume that you already have R.

To auto-generate model-code in R, you need to install the package
[Ryacas](https://cran.r-project.org/web/packages/Ryacas/index.html),
which is needed for model building:

```{r Ryacas, eval=FALSE}
install.packages("Ryacas")
```

You can now install the packages developed by the UQSA team:
```{r remotes, eval=FALSE}
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```

# MPI

MPI is intended for compute clusters (HPC) and is typically not needed
if you have one machine you want to do simulations on.

We do have several examples about MPI usage with this package (see
documentation on [Parallel chains with MPI](mpi.html) and Example
[UQ on deterministic AKAR4 with MPI](sampleAKAR4mpi.html)). However, if
you want to try them, you will need MPI.

To install MPI on Debian/Ubuntu, in the shell:

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

To install R bindings, in R:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ (R
package interfacing with MPI libraries on your system).
