---
title: "Explanation for System Prerequisites
"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Everything in this article should be already known to all users and
cover basic interactions with an operating system in the context of
modelling, scientific computing, and simulations. Read this only for
_troubleshooting_ purposes.

In this guide we explain these things in much more detail than will be
necessary for most (you probably already know how to install things).

So here is the shortest version of what you need to do, assuming that
you use `debian` or `ubuntu`:

```{sh, label="debian/ubuntu", eval=FALSE}
# tl;dr:
sudo apt install gcc pkg-config libgsl-dev r-base
```

You almost certainly already know which commands to run on your
distribution, with the package manager it uses.

The package `pkgconf` on Alpine Linux is not the original `pkg-config` but provides the `pkg-config`
command (no further action needed):

```{sh, label="alpine", eval=FALSE}
sudo apk add gcc gsl pkgconf R         # should suffice
```

The following material is the long explanation on a few example
systems.

# Motivation for C related Packages

In this package, we frequently solve an ordinary differential equation
model in the context of biological experiments. This can be done using
`deSolve`, but we prefer to use the solvers from the [GNU Scientific
Library](https://www.gnu.org/software/gsl/), GSL (the `gsl_odeiv2`
library).

**Comment** If you insist on using `deSolve` you can write a custom
`simulate(p)` function which takes as argument a parameter vector `p`
and returns a list with simulation results shaped like the results of
the simulator in this package. R functions are closures, so as long as
the necessary constant ingredients for `simulate` are in your
workspace, the function will have access to them. UQSA will not even
know the difference.

We have written an interface to the GSL solvers (`simulator.c`). The
user facing functions are designed around the idea of simulation
*experiments*: a list of instructions that result in a model
simulation, possibly with multiple calls to the ODE solver to obtain
the final time-series.

This experiment oriented interface is more convenient in our context
(and usually faster than using deSolve with R functions).

The GNU Scientific Library (GSL) is a system level dependency, which R
cannot install by itself. Usually the administrator of your machine
can do this (on the cluster). If that is you (e.g. on your laptop) and
you don't have GSL yet, then it is usually a one or two line command
to install it.

# System Prerequisites

If you are using an apple device, then you need some package manager
that macbooks don't come pre-configured with (which is a bummer). One
such package-manager is [homebrew](https://brew.sh/), this is
unofficial work by volunteers, funded by donations. Install it using
the command pasted on the homebrew homepage.

You are probably already using it or something very similar, but if
not then please do.

If you have used it in the past, a long time ago, then it would be
great to do an upgrade beforehand.

```{sh}
brew update && brew outdated && brew upgrade && brew cleanup
```

Alternatively do a fresh installation of homebrew, if all homebrew
packages are terribly out of date.

WARNING: macOS may mention xcode at some point, apple is really keen
on this. We have (of course) absolutely nothing to do with xcode,
neither does _gcc_, or _pkg-config_ (both are in no way owned by
apple). The insistence of macOS to provide _gcc_ only through xcode
with mandatory _developer ids_ is quite despicable as they have no
fundamental right to put any constraints on things that don't belong
to them.

## GNU Scientific Library and pkg-config

To compile C sources at all (with or without gsl) requires a c
compiler (any one will do, they provide the command `cc`, regardless
of vendor).

But, we also need system specific compiler options related to linking
and the location where `libgsl` is stored. These compiler options can look
like this: `-lgsl -lgslcblas -lm`, but also more complex (this depends
on your system) -- in an HPC environment they will almost certainly be
very different.

We use
[pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/) to
determine the right compiler options (or `pkgconf`, which does the
same thing (but faster)); `pkg-config` is available for macOS, GNU
Linux, Unix, and *BSD (even Windows, not that it helps).

|  Operating System | install gsl                   | install pkg-config              |
|------------------:|:------------------------------|:--------------------------------|
| Debian and Ubuntu | `sudo apt install libgsl-dev` | `sudo apt install pkg-config`   |
|      Alpine Linux | `doas apk add gsl`            | `doas apk add pkgconf`          |
|          GNU Guix | `guix install gsl`            | `guix install pkg-config`       |
|        Arch Linux | `sudo pacman -S gsl`          | `sudo pacman -S pkgconf`        |
|            Gentoo | `sudo emerge sci-libs/gsl`    | `sudo emerge virtual/pkgconfig` |
|             macOS | `brew install gsl`            | `brew install pkg-config`       |

If you are in a *normal user* shell (not a root shell, via `su -` or
similar), and not using `guix` (or `NixOS`), then the commands require
`sudo` or `doas` in front of them to have the required system
administration rights.

GNU Guix and NIXOS have package managers that don't require
administration rights.

### macOS and pkgconf

You can use `pkgconf` instead of `pkg-config` on macOS. However, it is possible that

```{sh, eval=FALSE}
brew install pkgconf
```

will not actually provide the `pkg-config` command in this spelling
(it may just be pkgconf).

This is again not something that has anything to do with us. Everyone
who wants to use `pkg-config` via pkgconf on macOS needs to consider
these things.

Make sure that the literal command `pkg-config` is actually
available after installation.  If it is not, then you can make a symbolic
link to `pkgconf` and make it available:

```{sh, eval=FALSE}
which pkg-config      # e.g.: /usr/bin/pkg-config
```

should print a path if that command is available (or nothing, then it
isn't).

## R itself

Availability of specific versions of R depend on the version of your OS,
how up-to-date your package database is, etc.. Here is a quick list of the
approximate command:

| Operating System | install R                 |
|-----------------:|:--------------------------|
|     Alpine linux | `doas apk add R`          |
|    Debian/Ubuntu | `sudo apt install r-base` |
|         GNU Guix | `guix install r`          |
|             Arch | `sudo pacman -S r`        |
|           Gentoo | `emerge --ask dev-lang/R` |

On MAC, please follow the instructions on the
[R-Studio](https://posit.co/download/rstudio-desktop/) web-site. Or
install the version from homebrew. These two will almost certainly be/have
different R versions, with different _locations_ for packages. It is
probably better to use only one of them (_R-Studio_ or _brew R_).

# R packages

Your system's package manager may have system-wide R packages, e.g.:
ubuntu has `r-cran-desolve`, such that `sudo apt install
r-cran-desolve` would install the *deSolve* package for all users of
your machine. In HPC environments this is preferable as user accounts
can have harsh quotas on *file sizes* and *file numbers*.

R packages can also be installed from within R (for individual
users), but only to a location that is writable to you, typically in
your `$HOME` diectory. Please ensure that `.libPaths()` returns at
least one location that you have write access to (this -- again -- is
true for everyone, has nothing to do with us).

```{r, .libPaths, eval=FALSE}
.libPaths()
```

If this command doesn't print a path that you can write to, then you
can create a new directory (in your home), and make it known to R:

```{r, eval=FALSE}
dir.create(
    Sys.getenv("R_LIBS_USER"),
    recursive = TRUE,
    showWarnings = FALSE
)

.libPaths(
    Sys.getenv("R_LIBS_USER")
)
```

Now, you can install R packages from *The Comprehensive R Archive Network*
[cran](https://cran.r-project.org/) and [github](https://github.com/):

```{r, label='install packages from CRAN', eval=FALSE}
install.packages("remotes")
```

UQSA is not on CRAN (yet), and can be installed like this:

```{r, remotes, eval=FALSE}
remotes::install_github("icpm-kth/uqsa")
```

# Installation of Companion packages (Optional)

We have developed two companion packages:

- [rgsl](icpm-kth/rgsl)
    - only simulates a model, no uncertainty quantification
    - solves ODEs with an interface organised around *simulation experiments*
    - uses solvers from the GNU Scientific Library
    - can be used in most of our examples, instead of `uqsa::simulator.c`
- [SBtabVFGEN](icpm-kth/SBtabVFGEN)
    - organises the storage and loading of models in SBtab form
    - loads SBtab content from tsv, ods, and excel files
    - creates vfgen, MOD, sbml, and format free ODE files
    - also loads biological data contained in the SBtab files into
      R variables (lists of *simulation experiment* setups)

Both are optional. Their capabilities resemble the ones in UQSA, but
different.

## RGSL

Like the functions in this package `rgsl` solves ODE initial value
problems, given as lists of simulation experiments, with sudden
interventions (like *activation*, or a sudden signal).

```{r, eval=FALSE}
remotes::install_github("icpm-kth/rgsl")       # requires gsl in your OS, see above
```

This package has several subtly different solver interfaces, but is
very similar to the solvers in the UQSA package. It can be used
independently of this package and has no uncertainty quantification or
sensitivity analysis functions, just the simulator.

## SBtab Model Handling

This package loads a model written in the SBtab format, which is
designed for models in systems biology, but much easier to read than
SBML:

```{r, eval=FALSE}
remotes::install_github("icpm-kth/SBtabVFGEN") # if you plan to use SBtab
```

This is highly optional, if you have a different method of model creation
or want to write the model source files by hand (which is hard for large models).

It is an alternative for the TSV model format in this package. Use
SBtabVFGEN if your model alfready is in SBtab form.

