---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites for Model Building

See the Sections below for instructions. Short summary of what they do:

* Install the [GNU Scientific Library](https://www.gnu.org/software/gsl/)
* [yacas](http://www.yacas.org/index.html) and [Ryacas](https://cran.r-project.org/package=Ryacas)
  - `install.packages("Ryacas")` after _yacas_

## Pick One

```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to xcode license
brew install pkg-config gcc gsl yacas
```

```{sh ubuntu/debian, eval=FALSE}
# other linux distributions have some version of this:
# zypper install, apk add, guix install
sudo apt install pkg-config gcc libgsl-dev yacas
```

# Installation of the R packages

```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages(remotes)
remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```

# MPI

If you want to run on several nodes of a cluster via MPI (debian/ubuntu):

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

and in R:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ and
_pbdMPI_ (R package interfacing with MPI libraries on your system).
