---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisits

The simulators (ODE solvers) we use need the **GNU Scientific library**
[(GSL)](https://www.gnu.org/software/gsl/) to be installed in your system.

The model building part of UQSA (transferring the SBtab network model
to a mathematical model in C or R) also requires that you have a **C
compiler** (e.g. [gcc](https://gcc.gnu.org)).

On some systems the GSL may be installed in an unexpected location (on
clusters this will almost certainly be the case). So, we rely on
[pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/) to
find this location.  

All prerequisites can be installed all at once with the
following command (that you have to write in a terminal window).

### On macOS

To install libraries and science related software, people typically
opt for [homebrew](https://brew.sh/), or [nix](https://nixos.org). Here is an example using homebrew:

```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to Xcode license
brew install pkg-config gcc gsl
```

MacOS does not have a default package manager for software of this
kind. There is probably no easier way of getting the above done on a
factory fresh device without first installing brew or nix.

### On GNU/Linux

Ubuntu or Debian:

```{sh ubuntu/debian, eval=FALSE}
sudo apt install pkg-config gcc libgsl-dev
```

Other distributions have some version of this:

```
sudo apk add gsl pkgconf            # Alpine
sudo guix install gsl pkg-config    # GNU Guix
sudo pacman -S gsl pkgconf          # Arch, btw.
sudo zypper install gsl pkg-config  # OpenSuSE, SLES
```

... and many [more](articles/installExplanations.html).

# Installation of the R packages

In order to use UQSA you (obviously) need the [R language](https://www.r-project.org/) (and maybe a graphical interface to it like [R-studio](https://posit.co/download/rstudio-desktop/)). We will assume that you already have R.

To auto-generate model-code in R, you need to install the package [Ryacas](https://cran.r-project.org/web/packages/Ryacas/index.html), which is needed for model building:

```{sh Ryacas, eval=FALSE}
install.packages("Ryacas")
```

You can now install the packages developed by the UQSA team:
```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages("remotes")

remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```

# MPI

MPI is intended for compute clusters (HPC) and is typically not needed if
you have one machine you want to do simulations on.

We do have several examples about MPI usage with this package, however, if you want to try them, you will need MPI.

To install MPI on Debian/Ubuntu:

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

TO install R bindings:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ and
_pbdMPI_ (R package interfacing with MPI libraries on your system).
