---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```



# Prerequisits


The model building part of UQSA (transfering the SBtab network model to a mathematical model in C or R) requires that you have a C compiler (e.g. [gcc](https://gcc.gnu.org)), the GNU Scientific Library [(gsl)](https://www.gnu.org/software/gsl/) and [pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/).
These can be installed all at once with the following command (that you have to write in the terminal window).

**In macOS:**
```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to xcode license
brew install pkg-config gcc gsl
```

**In linux (ubuntu/debian):**
```{sh ubuntu/debian, eval=FALSE}
# other linux distributions have some version of this:
#  zypper install,
#  apk add,
#  guix install,
# etc.
sudo apt install pkg-config gcc libgsl-dev
```

# Installation of the R packages
In order to use UQSA you (obviously) need the [R language](https://www.r-project.org/) (and maybe a graphical interface to it like [R-studio](https://posit.co/download/rstudio-desktop/)).

In R, you first need install the package [Ryacas](https://cran.r-project.org/web/packages/Ryacas/index.html):
```{sh Ryacas, eval=FALSE}
install.packages("Ryacas")
```
which is needed for model building.

You can now install the packages developed by the uqsa team:
```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages(remotes)
remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```



# MPI

If you want to run on several nodes of a cluster via MPI (debian/ubuntu):

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

and in R:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ and
_pbdMPI_ (R package interfacing with MPI libraries on your system).
