---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```



# Prerequisits


The model building part of UQSA (transfering the SBtab network model to a mathematical model in C or R) requires that you have a C compiler (e.g. [gcc](https://gcc.gnu.org)), the GNU Scientific Library [(gsl)](https://www.gnu.org/software/gsl/) and [pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/).
These can be installed all at once with the following command (that you have to write in the terminal window).

## On macOS

To install libraries and science related software, people typically
opt for [homebrew](https://brew.sh/), or [nix](https://nixos.org). Here is an example using brew:

```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to xcode license
brew install pkg-config gcc gsl
```

MacOS does not have a default package manager for software of this
kind. There is probably no easier way of getting the above done on a
factory fresh device without first installing brew or nix.

## On GNU/Linux

Ubuntu or debian:

```{sh ubuntu/debian, eval=FALSE}
sudo apt install pkg-config gcc libgsl-dev
```

Other distributions have some version of this:

```
zypper install  # OpenSuSE, SLES
apk add         # Alpine
guix install    # GNU Guix
```

... and many more. 

# Installation of the R packages

In order to use UQSA you (obviously) need the [R language](https://www.r-project.org/) (and maybe a graphical interface to it like [R-studio](https://posit.co/download/rstudio-desktop/)). We will assume that you already have R.

To auto-generate model-code in R, you need install the package [Ryacas](https://cran.r-project.org/web/packages/Ryacas/index.html):
```{sh Ryacas, eval=FALSE}
install.packages("Ryacas")
```
which is needed for model building.

You can now install the packages developed by the uqsa team:
```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages("remotes")
remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```

# MPI

MPI is intended for compute clusters (HPC) and is typically not needed if
you have one machine you want to do simulations on.

We do have several examples about MPI usage with this package however and if you want to try them, you will need MPI.

On debian/ubuntu:

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

and in R:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ and
_pbdMPI_ (R package interfacing with MPI libraries on your system).
