---
title: "Installation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this package, we frequently solve an ordinary differential equation
model in the context of biological experiments. This can be done using
`deSolve`, but we prefer to use `gsl_odeiv2`, for whcih we have
written a binding into R (called `rgsl`). The user facing interface is
designed around the idea of *experiments* and is more convenient (and
faster). To use this simulation backend, your system requires the gnu
scientific library, which R cannot install by itself, usually the
admin of your machine can do this. If that is you and you don't have
GSL yet, then follow the guide in the next section.

We know that it is in principle possible to use this package on
windows, but have no experience with this, and cannot provide
instructions or help with problems.

# GNU Scientific Library (optional)

For the simulation backend using the [GNU Scientific
Library](https://www.gnu.org/software/gsl/) (gsl, odeiv2), the library
needs to be installed in your OS, it is not an R package (it's a
library dependency).

To compile sources with gsl requires c compiler options related to
linking and the librarie's location. They can look like this: `-lgsl
-lgslcblas -lm`, but also more complex (this depends on your
system). We use
[pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/) to
determine the right options (or `pkgconf`, which does the same thing
(but faster)). If `pkg-config` is missing from the system,
installation will fail (it is available for MAC, GNU Linux, Unix, BSD,
and Windows).

|  Operating System | install gsl                   | install pkg-config                     |
|------------------:|:------------------------------|:---------------------------------------|
| Debian and Ubuntu | `sudo apt install libgsl-dev` | `sudo apt install pkg-config`          |
|      Alpine Linux | `sudo apk add gsl`            | `sudo apk add pkgconf`                 |
|          GNU Guix | `guix install gsl`            | `guix install pkg-config`              |
|        Arch Linux | `sudo pacman -S gsl`          | `sudo pacman -S pkgconf`               |
|            Gentoo | `sudo emerge sci-libs/gsl`    | `sudo emerge virtual/pkgconfig`        |
|             MACOS | `brew install gsl`            | `brew install pkg-config` or `pkgconf` |


Once `gsl` and `pkg-config` are available on the system, the
`icpm-kth/rgsl` package should compile nicely.

If you are in a *normal user* shell (not a root shell, `su -`), and not
using `guix` or `NixOS`, then the commands require `sudo` or `doas` in
front of them to have the required system administration rights. GNU
Guix and NIXOS have package managers that don't require administration
rights.

# R itself

Availability of specific versions depend on the version of your OS,
how up-to-date your package database. Here is a quick list of the
approximate command:

| Operating System | install R                 |
|-----------------:|:--------------------------|
|     Alpine linux | `sudo apk add R`          |
|    Debian/Ubuntu | `sudo apt install r-base` |
|         GNU Guix | `guix install r`          |
|             Arch | `sudo pacman -S r`        |
|           Gentoo | `emerge --ask dev-lang/R` |

On MAC, please follow the instructions on the
[R-studio](https://posit.co/download/rstudio-desktop/) web-site.

We do not have experience with R on windows, specifically when it
comes to GSL and C compilers to compile the C code included in an R
package, it may just work, but if not, then maybe these could help:

- [WSL](https://en.wikipedia.org/wiki/Windows_Subsystem_for_Linux)
- [Cygwin](https://en.wikipedia.org/wiki/Cygwin))



# R packages

Your system's package manager may have system-wide R packages, e.g.:
ubuntu has `r-cran-desolve`, such that `sudo apt install
r-cran-desolve` would install the *deSolve* package for all users of
your machine. In HPC environments this is preferable as user accounts
can have harsh quotas on *file sizes* and *file numbers*.

But, R packages can also be installed from within R (for individual
users), but only to a location that is writable to you, typically in your
`$HOME` diectory. Please ensure that `.libPaths()` returns at least
one location that you have write access to.

```{r, .libPaths, eval=FALSE}
.libPaths()
```

If not, you can create a new directory, and make it known to R:

```{sh, eval=FALSE}
mkdir -p ~/R/library && echo "R_LIBS=~/R/library" >> ~/.Renviron
```

Now, you can install R packages from *The Comprehensive R Archive Network* [cran](https://cran.r-project.org/) and [github](https://github.com/):

```{r, label='install packages from CRAN', eval=FALSE}
install.packages("remotes")
```

Our packages are not on CRAN, and can be installe dlike this:

```{r, remotes, eval=FALSE}
remotes::install_github("icpm-kth/uqsa")
```

# Installation of Companion packages

We have developed two companion packages:

- [rgsl](icpm-kth/rgsl)
    - solves ODEs with an interface organised around *simulation experiments*
    - uses solvers from the GNU Scientific Library
    - is used in most of our examples
- [SBtabVFGEN](icpm-kth/SBtabVFGEN)
    - organises the storage and loading of models
    - loads SBtab content from tsv, ods, and excel files
    - creates vfgen, MOD, sbml, and format free ODE files
    - also loads biological data contained in the SBtab files into
      R variables (lists of *simulation experiment* setups)

Both are optional, but recommended (our examples use them).

## RGSL

This package solves ODE initial value problems, given as lists of
simulation experiments, with sudden interventions (like *activation*,
or a sudden signal).

```{r, eval=FALSE}
remotes::install_github("icpm-kth/rgsl")       # requires gsl in your OS, see above
```

This can be replaced with the standard solvers in the deSolve
package. In our examples, the solvers in deSolve are always slower.

## SBtab Model Handling

This package loads a model written in the SBtab format, which is
designed for models in systems biology, but much easier to read than
SBML:

```{r, eval=FALSE}
remotes::install_github("icpm-kth/SBtabVFGEN") # if you plan to use SBtab
```

This can be replaced, if you have a different method of model creation
or want to write the model source files by hand (which is hard for
large models).

# Installation of UQSA

```{r, eval=FALSE}
remotes::install_github("icpm-kth/uqsa")       # this package
library(uqsa)                                  # to load it
```

Read the articles on this website, especially the *Example* articles
to learn more: AKAR4, AKAP79, and CaMKII.

## Examples in the Package

The package itself also includes example models, with a *run* script
for each of them. We advise users to copy the run scripts either from
the repository or the installed package and adapt them. The location
of a *run* file can be found like this:

```{r, eval=FALSE}
uqsa::uqsa_example("AKAP79",pat="^run.*R$")
```

or by browsing the folders on GitHub (`inst/extdata/`). Copy one of
them and adapt it to your needs.

