---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites

* Install the [GNU Scientific Library](https://www.gnu.org/software/gsl/)
* The auto-generation of code for each model is done using the shell script in
  `icpm-kth/RPN-derivative`; that package has several backends for
  analystic (symbolc) math (the thing that computer algebra systems
  do)
    - [maxima](https://maxima.sourceforge.io/)
    - [yacas](http://www.yacas.org/index.html)
    - the [RPN-derivative](icpm-kth/RPN-derivative) package
      itself has code to perform derivatives
      (needs `make && sudo make install`)

## Pick One

maxima is optional

```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to xcode license
brew install pkg-config gcc gsl maxima
```

```{sh ubuntu/debian, eval=FALSE}
# other linux distributions have some version of this:
# zypper install, apk add, guix install
sudo apt install pkg-config gcc libgsl-dev maxima
```

## Analytical derivatives and Model Code Generation

You may have to adjust this to fit your specific setup (do you use `zsh`? Does it parse `.profile`)

```{sh derivatives, eval=FALSE}
git pull https://github.com/icpm-kth/RPN-derivative.git
cd RPN-derivative
make
# these commands (or similar) make the ode script available from any location
[ -d "$HOME/.local/bin" ] && ln -s "`pwd`/sh/ode.sh" "$HOME/.local/bin/ode"
[ -L "$HOME/.local/bin/ode" ] && echo "export PATH=\"$HOME/.local/bin:$PATH\"" >> $HOME/.profile
```

```{sh optionally}
# this installs the derivative program in RPN-derivative
sudo make install
```

# Installation

```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages(remotes)
remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```

