---
title: "Explanation for System Prerequisites
"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

Everything in this article should be already known to all users and
cover basic interactions with an operating system in the context of
modelling, scientific computing, and simulations. Read this only for
_troubleshooting_ purposes.

In this guide we explain these things in much more detail than will be
necessary for most (you probably already know how to install things).

So here is the shortest version of what you need to do, assuming that
you use `debian` or `ubuntu`:

```{sh, label="debian/ubuntu", eval=FALSE}
# tl;dr:
sudo apt install gcc pkg-config libgsl-dev r-base
```

You already know which commands to run on your distribution, with the
package manager it uses.

The package `pkgconf` on Alpine Linux is not the original `pkg-config` but provides the `pkg-config`
command (no further action needed):

```{sh, label="alpine", eval=FALSE}
sudo apk add gcc gsl pkgconf R         # should suffice
```

The following material is the long explanation on a few example
systems (it is impossible to list all of them). 

# Motivation

In this package, we frequently solve an ordinary differential equation
model in the context of biological experiments. This can be done using
`deSolve`, but we prefer to use the solvers from the [GNU Scientific
Library](https://www.gnu.org/software/gsl/), GSL (the
`gsl_odeiv2` library). We have written an interface to these solvers
(called `rgsl`). The user facing interface is designed around the idea
of simulation *experiments*: a list of instructions that result in a
model simulation, possibly with multiple calls to the ODE solver to
obtain the final time-series.

This experiment oriented interface is more convenient in our context
(and usually much faster than using deSolve). The GSL is a system
level dependency, which R cannot install by itself, usually the admin
of your machine can do this (on the cluster). If that is you (e.g. on
your laptop) and you don't have GSL yet, then it is usually a one or two line
command to install it.

# System Prerequisites

If you are using an apple device, then you need a package manager that
macbooks don't come pre-configured with. One such package-manager is
[homebrew](https://brew.sh/) (this is unofficial work funded by
donations) - install it using the command pasted on the homebrew
homepage. You are probably already using it or something very similar,
but if not then please do.

WARNING: macOS will definitely mention xcode at some point, apple is
really keen on this. We have absolutely nothing to do with xcode,
neither does gcc, or pkg-config. The insistence of macOS to provide
gcc only through xcode and developer ids is despicable as they have no
right to put up any constraints on things that don't belong to them.

## GNU Scientific Library and pkg-config

To compile C sources at all (with or without gsl) requires a c
compiler. We also need system specific compiler options related to
linking and the location where gsl is stored. These compiler options
can look like this: `-lgsl -lgslcblas -lm`, but also more complex
(this depends on your system) -- in an HPC environment they will
almost certainly be very different.

We use
[pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/) to
determine the right compiler options (or `pkgconf`, which does the
same thing (but faster)); `pkg-config` is available for macOS, GNU
Linux, Unix, and *BSD (even Windows, not that it helps).

|  Operating System | install gsl                   | install pkg-config              |
|------------------:|:------------------------------|:--------------------------------|
| Debian and Ubuntu | `sudo apt install libgsl-dev` | `sudo apt install pkg-config`   |
|      Alpine Linux | `sudo apk add gsl`            | `sudo apk add pkgconf`          |
|          GNU Guix | `guix install gsl`            | `guix install pkg-config`       |
|        Arch Linux | `sudo pacman -S gsl`          | `sudo pacman -S pkgconf`        |
|            Gentoo | `sudo emerge sci-libs/gsl`    | `sudo emerge virtual/pkgconfig` |
|             macOS | `brew install gsl`            | `brew install pkg-config`       |

Once `gsl` and `pkg-config` are available on the system, the
`icpm-kth/rgsl` package should compile nicely.

If you are in a *normal user* shell (not a root shell, `su -`), and not
using `guix` (or `NixOS`), then the commands require `sudo` or `doas` in
front of them to have the required system administration rights. GNU
Guix and NIXOS have package managers that don't require administration
rights.

### macOS and pkgconf

You can use `pkgconf` instead of `pkg-config` on macOS. However, it is possible that 

```{sh, eval=FALSE}
brew install pkgconf
```

will not actually provide the `pkg-config` command in this spelling
(it may just be pkgconf).

This is again not something that has anything to do with us. Everyone
who wants to use `pkg-config` via pkgconf on macOS needs to consider
these things.

Make sure that the literal command `pkg-config` is actually
available after installation.  If it is not, then you can make a symbolic
link to `pkgconf` and make it available:

```{sh, eval=FALSE}
which pkg-config      # e.g.: /usr/bin/pkg-config
```

should print a path if that command is available (or nothing, then it
isn't).

## R itself

Availability of specific versions of R depend on the version of your OS,
how up-to-date your package database is, etc.. Here is a quick list of the
approximate command:

| Operating System | install R                 |
|-----------------:|:--------------------------|
|     Alpine linux | `sudo apk add R`          |
|    Debian/Ubuntu | `sudo apt install r-base` |
|         GNU Guix | `guix install r`          |
|             Arch | `sudo pacman -S r`        |
|           Gentoo | `emerge --ask dev-lang/R` |

On MAC, please follow the instructions on the
[R-Studio](https://posit.co/download/rstudio-desktop/) web-site. Or
install the version from homebrew. These two will almost certainly be
different versions, with different locations for packages. It is
probably better to use only one of them (_R-Studio_ or _brew R_).

# R packages

Your system's package manager may have system-wide R packages, e.g.:
ubuntu has `r-cran-desolve`, such that `sudo apt install
r-cran-desolve` would install the *deSolve* package for all users of
your machine. In HPC environments this is preferable as user accounts
can have harsh quotas on *file sizes* and *file numbers*.

R packages can also be installed from within R (for individual
users), but only to a location that is writable to you, typically in
your `$HOME` diectory. Please ensure that `.libPaths()` returns at
least one location that you have write access to (this -- again -- is
true for everyone, has nothing to do with us).

```{r, .libPaths, eval=FALSE}
.libPaths()
```

If this command doesn't print a path that you can write to, then you
can create a new directory (in your home), and make it known to R:

```{sh, eval=FALSE}
mkdir -p ~/R/library && echo "R_LIBS=~/R/library" >> ~/.Renviron
```

Now, you can install R packages from *The Comprehensive R Archive Network* [cran](https://cran.r-project.org/) and [github](https://github.com/):

```{r, label='install packages from CRAN', eval=FALSE}
install.packages("remotes")
```

Our packages are not on CRAN, and can be installed like this:

```{r, remotes, eval=FALSE}
require(remotes)
install_github("icpm-kth/uqsa")
```

# Installation of Companion packages

We have developed two companion packages:

- [rgsl](icpm-kth/rgsl)
    - solves ODEs with an interface organised around *simulation experiments*
    - uses solvers from the GNU Scientific Library
    - is used in most of our examples
- [SBtabVFGEN](icpm-kth/SBtabVFGEN)
    - organises the storage and loading of models
    - loads SBtab content from tsv, ods, and excel files
    - creates vfgen, MOD, sbml, and format free ODE files
    - also loads biological data contained in the SBtab files into
      R variables (lists of *simulation experiment* setups)

Both are optional, but recommended (our examples use them).

## RGSL

This package solves ODE initial value problems, given as lists of
simulation experiments, with sudden interventions (like *activation*,
or a sudden signal).

```{r, eval=FALSE}
remotes::install_github("icpm-kth/rgsl")       # requires gsl in your OS, see above
```

This can be replaced with the standard solvers in the deSolve
package. In our examples, the solvers in deSolve are always slower.

It is possible to use shared libraries with deSolve, to make solutions
faster, but this is not very easy to do.

## SBtab Model Handling

This package loads a model written in the SBtab format, which is
designed for models in systems biology, but much easier to read than
SBML:

```{r, eval=FALSE}
remotes::install_github("icpm-kth/SBtabVFGEN") # if you plan to use SBtab
```

This can be replaced, if you have a different method of model creation
or want to write the model source files by hand (which is hard for
large models).

# Installation of UQSA

```{r, eval=FALSE}
remotes::install_github("icpm-kth/uqsa")       # this package
library(uqsa)                                  # to load it
```

## Examples in the Package

The package itself also includes example models, with a *run* script
for each of them. We advise users to copy the run scripts either from
the repository or the installed package and adapt them. The location
of a *run* file can be found like this:

```{r, eval=FALSE}
uqsa::uqsa_example("AKAP79",pat="^run.*R$")
```

or by browsing the folders on GitHub (`inst/extdata/`). Copy one of
them and adapt it to your needs.

