---
title: "Short Installation Instructions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


# Installation of the R packages
In order to use UQSA you (obviously) need the 
   [R language](https://www.r-project.org/) (and maybe a graphical interface to it like
   [R-studio](https://posit.co/download/rstudio-desktop/)).
With this you can run most of our [examples](./examples_overview.html) by installing the following packages:
```{r remotes, eval=FALSE}
# requires the 'remotes' package
if (!require(remotes)) install.packages(remotes)
remotes::install_github("icpm-kth/SBtabVFGEN",dependencies=TRUE)
remotes::install_github("icpm-kth/rgsl",dependencies=TRUE)
remotes::install_github("icpm-kth/uqsa",dependencies=TRUE)
```
To also build your own model from SBtab files you need a few external programs and one additional R package. 

# Model building
The model building part of UQSA (transfering the SBtab network model to a mathematical model in C or R) requires that you have a C compiler (e.g. [gcc](https://gcc.gnu.org)), the GNU Scientific Library [(gsl)](https://www.gnu.org/software/gsl/) and [pkg-config](https://www.freedesktop.org/wiki/Software/pkg-config/).
These can be installed all at once with the following command (that you have to write in the terminal window of your computer).

**In macOS:**
```{sh macOS, eval=FALSE}
# macOS has a lot of issues with xcode, it loves it for some reason; maybe agree to xcode license
brew install pkg-config gcc gsl
```

**In linux (ubuntu/debian):**
```{sh ubuntu/debian, eval=FALSE}
# other linux distributions have some version of this:
#  zypper install,
#  apk add,
#  guix install,
# etc.
sudo apt install pkg-config gcc libgsl-dev
```

Model building also requires an additional R package (to be installed in R) based on [yacas](http://www.yacas.org/index.html):
```{sh Ryacas, eval=FALSE}
install.packages("Ryacas")
```

# MPI

If you want to run on several nodes of a cluster via MPI (debian/ubuntu):

```{sh MPI, eval=FALSE}
sudo apt install libopenmpi-dev
```

and in R:

```{r pbdMPI, eval=FALSE}
install.packages("pbdMPI")
```

On other systems, install the equivalent package for _OpenMPI_ and
_pbdMPI_ (R package interfacing with MPI libraries on your system).
