---
title: "Importing Models into R"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Systems Biology Models are usually described with reactions, kinetic
laws, parameters, and possibly inputs and outputs.

In this R package, we use plain TSV files to describe these
quantities. These TSV files have very few rules. One universal rule is
that the first column is always reserved for row-names. Row-names must
work as variable names in R and C. Certan names would conflict with
builtin commands (`if`, `while`, `do`, `continue`, `break`). Obviously
such names cannot be used for reacting compounds or parameters.  In
our examples, we label the first column with `id` (but it can have any
name).

They can be imported into R using the command `uqsa::model_from_tsv()`:

```{r, eval=FALSE}
library(uqsa)
library(errors)

## Read the TSV files, create an ODE model
m <- model_from_tsv()    # defaults to files from $(pwd)
o <- as_ode(m)           # the ODE

C <- generateCode(o)     # list of strings
odeModel <- write_and_compile(C)
ex <- experiments(m,o)
```

The model tables follow these guidelines:

- A list of quantities, like parameters have these columns:
  + id (first column, any heading)
  + value
  + unit
  + scale (optional, assumed to be linear)
- A list of mathematical expressions can use the term `formula`
  instead of `value`:
  + id
  + formula
  + unit
- Any quantity that is defined using a row in some table can get
  updated values in another table, or values to comapre against
- Data tables need to specifiy values using a notation that keeps
  values and their uncertainties together, e.g.:
  + `1.234(21)`
  + `1.5(1)e+4`
- The tables are found by name, the names our import scripts will be looking for are:
  + Constant.tsv
  + Parameter.tsv
  + Input.tsv
  + Expression.tsv
  + Reaction.tsv
  + Output.tsv
  + Experiment.tsv
- The Experiment table describes the experimental conditions the model
  needs to be subjected to in order to replicate measured data:
  + each experimental setting gets a row
  + each row id is used to find the data
  + the data is found in the TSV file named like the row-id
  + Data for experiment `EX1` is expected to be in the file `EX1.tsv`

This package will assume that the directory/folder where the TSV files reside in is the name of the model. This is an example folder structure:

```{sh}
~/Documents/AKAP79$ ls -1 *.tsv *.md
cAMP_increase.tsv
Compound.tsv
EX00___cAMP___CaN_AKAP79.md
EX11____0nM__TRUE___TRUE.tsv
EX12____0nM__TRUE__FALSE.tsv
EX13____0nM_FALSE__FALSE.tsv
EX21__100nM__TRUE___TRUE.tsv
EX22__100nM__TRUE__FALSE.tsv
EX23__100nM_FALSE__FALSE.tsv
EX31__200nM__TRUE___TRUE.tsv
EX32__200nM__TRUE__FALSE.tsv
EX33__200nM_FALSE__FALSE.tsv
EX41__500nM__TRUE___TRUE.tsv
EX42__500nM__TRUE__FALSE.tsv
EX43__500nM_FALSE__FALSE.tsv
EX51_1000nM__TRUE___TRUE.tsv
EX52_1000nM__TRUE__FALSE.tsv
EX53_1000nM_FALSE__FALSE.tsv
EX61_2000nM__TRUE___TRUE.tsv
EX62_2000nM__TRUE__FALSE.tsv
EX63_2000nM_FALSE__FALSE.tsv
EX71___dose__TRUE___TRUE.tsv
EX72__event__TRUE___TRUE.tsv
Experiments.tsv
Expression.tsv
Input.tsv
Output.tsv
Parameter.tsv
Reaction.tsv
README.md
Transformation.tsv
```

Here is an example command to view a TSV file nicely formatted:

```{sh}
~/Documents/AKAP79$ column -t -s $'\t' Reaction.tsv
id   kinetic law                                reactants        arw  products        is reversible
r51  k5_1*Rii_C                                 Rii_C            <=>  RiiP_C          0
r14  k4_1*RiiP*C - k1_4*RiiP_C                  RiiP + C         <=>  RiiP_C          1
r12  k1_2*RiiP_C*cAMP - k2_1*RiiP_C_cAMP        RiiP_C + cAMP    <=>  RiiP_C_cAMP     1
r43  k4_3*cAMP*RiiP - k3_4*RiiP_cAMP            cAMP + RiiP      <=>  RiiP_cAMP       1
r23  k3_2*RiiP_cAMP*C - k2_3*RiiP_C_cAMP        RiiP_cAMP + C    <=>  RiiP_C_cAMP     1
r78  k8_7*cAMP*Rii - k7_8*Rii_cAMP              cAMP + Rii       <=>  Rii_cAMP        1
r56  k5_6*Rii_C*cAMP - k6_5*Rii_C_cAMP          Rii_C + cAMP     <=>  Rii_C_cAMP      1
r76  k7_6*Rii_cAMP*C - k6_7*Rii_C_cAMP          Rii_cAMP + C     <=>  Rii_C_cAMP      1
r62  k6_2*Rii_C_cAMP                            Rii_C_cAMP       <=>  RiiP_C_cAMP     0
r58  k8_5*Rii*C - k5_8*Rii_C                    Rii + C          <=>  Rii_C           1
r44  k4_4p*RiiP*CaN - k4p_4*RiiP_CaN            RiiP + CaN       <=>  RiiP_CaN        1
r33  k3_3p*CaN*RiiP_cAMP - k3p_3*RiiP_cAMP_CaN  CaN + RiiP_cAMP  <=>  RiiP_cAMP_CaN   1
r48  k4p_8*RiiP_CaN                             RiiP_CaN         <=>  Rii + CaN       0
r37  k3p_7*RiiP_cAMP_CaN                        RiiP_cAMP_CaN    <=>  CaN + Rii_cAMP  0
r1   kf_C_AKAR4*C*AKAR4 - kb_C_AKAR4*AKAR4_C    C + AKAR4        <=>  AKAR4_C         1
r2   kcat_AKARp*AKAR4_C                         AKAR4_C          <=>  AKAR4p + C      0
```

In the following sections, we give examples for each of the tables, as
markdown tables.

# Reaction

+-------:+:---------------------------------------------:+----------------:+:-------:+:---------------+:---------------------+
| **id** | **kinetic law**                               | **reactants**   | **arw** | **products**   | **is reversible**    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r51    | k5\_1\*Rii\_C                                 | Rii\_C          | \<=\>   | RiiP\_C        | 0                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r14    | k4\_1\*RiiP\*C - k1\_4\*RiiP\_C               | RiiP + C        | \<=\>   | RiiP\_C        | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r12    | k1\_2\*RiiP\_C\*cAMP - k2\_1\*RiiP\_C\_cAMP   | RiiP\_C + cAMP  | \<=\>   | RiiP\_C\_cAMP  | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r43    | k4\_3\*cAMP\*RiiP - k3\_4\*RiiP\_cAMP         | cAMP + RiiP     | \<=\>   | RiiP\_cAMP     | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r23    | k3\_2\*RiiP\_cAMP\*C - k2\_3\*RiiP\_C\_cAMP   | RiiP\_cAMP + C  | \<=\>   | RiiP\_C\_cAMP  | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r78    | k8\_7\*cAMP\*Rii - k7\_8\*Rii\_cAMP           | cAMP + Rii      | \<=\>   | Rii\_cAMP      | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r56    | k5\_6\*Rii\_C\*cAMP - k6\_5\*Rii\_C\_cAMP     | Rii\_C + cAMP   | \<=\>   | Rii\_C\_cAMP   | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r76    | k7\_6\*Rii\_cAMP\*C - k6\_7\*Rii\_C\_cAMP     | Rii\_cAMP + C   | \<=\>   | Rii\_C\_cAMP   | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r62    | k6\_2\*Rii\_C\_cAMP                           | Rii\_C\_cAMP    | \<=\>   | RiiP\_C\_cAMP  | 0                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r58    | k8\_5\*Rii\*C - k5\_8\*Rii\_C                 | Rii + C         | \<=\>   | Rii\_C         | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r44    | k4\_4p\*RiiP\*CaN - k4p\_4\*RiiP\_CaN         | RiiP + CaN      | \<=\>   | RiiP\_CaN      | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r33    |k3\_3p\*CaN\*RiiP\_cAMP-k3p\_3\*RiiP\_cAMP\_CaN| CaN + RiiP\_cAMP| \<=\>   | RiiP\_cAMP\_CaN| 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r48    | k4p\_8\*RiiP\_CaN                             | RiiP\_CaN       | \<=\>   | Rii + CaN      | 0                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r37    | k3p\_7\*RiiP\_cAMP\_CaN                       | RiiP\_cAMP\_CaN | \<=\>   | CaN + Rii\_cAMP| 0                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r1     | kf\_C\_AKAR4\*C\*AKAR4-kb\_C\_AKAR4\*AKAR4\_C | C + AKAR4       | \<=\>   | AKAR4\_C       | 1                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+
| r2     | kcat\_AKARp\*AKAR4\_C                         | AKAR4\_C        | \<=\>   | AKAR4p + C     | 0                    |
+--------+-----------------------------------------------+-----------------+---------+----------------+----------------------+

# Compound

Reacting compounds, or molecular species. These will be turned into state variables.

  --------------- ----------- ---------- ------------- --------------------------------------------------
  **id**          **value**   **Unit**   **uniprot**   **Comment**
  Rii             6.3         µM         KAP2\_HUMAN   also KAP3\_HUMAN because Rii = RII\_alpha+RII\_beta
  cAMP            0           µM         NONE          NONE
  RiiP            0           µM         NONE          phosphorylated
  Rii\_C          0.63        µM         NONE          NONE
  RiiP\_cAMP      0           µM         NONE          NONE
  RiiP\_C         0           µM         NONE          NONE
  RiiP\_C\_cAMP   0           µM         NONE          NONE
  C               0           µM         KAPCB\_HUMAN  NONE
  Rii\_cAMP       0           µM         NONE          NONE
  Rii\_C\_cAMP    0           µM         NONE          NONE
  CaN             1.5         µM         PP2BA\_HUMAN  PP3R1\_HUMAN
  RiiP\_CaN       0           µM         NONE          NONE
  RiiP\_cAMP\_CaN 0           µM         NONE          NONE
  AKAR4           0.2         µM         NONE          RRID:addgene\_61619
  AKAR4\_C        0           µM         NONE          NONE
  AKAR4p          0           µM         NONE          NONE
  --------------- ----------- ---------- ------------- --------------------------------------------------

# Parameter

These are parameters, that will be subject to optimization or
sampling. The values are assumed to be crude estimates.

Known parameters should go into a different table.

  ------------ ----------- ------------ ---------- ------------------------ ----------- ----------
        **id**  **value**    **median** **stdv**    **prior distribution**   **scale**  **unit**
         k5\_1  1.667837          1.519 3            Normal distribution       log10    s\^-1
         k1\_2  0.0946379        −0.306 3            Normal distribution       log10    1/µM\*s
         k3\_2  −2.483545        −2.264 3            Normal distribution       log10    1/µM\*s
         k2\_3  −0.230731        −1.807 3            Normal distribution       log10    s\^-1
         k3\_4  −1.136198        −2.796 3            Normal distribution       log10    s\^-1
         k4\_3  −2.147382        −1.824 3            Normal distribution       log10    1/µM\*s
         k4\_1  −2.284185         −1.42 3            Normal distribution       log10    1/µM\*s
         k1\_4  −2.928144        −2.585 3            Normal distribution       log10    s\^-1
         k8\_7  −2.149449        −1.824 3            Normal distribution       log10    1/µM\*s
         k7\_8  −1.07931         −2.796 3            Normal distribution       log10    s\^-1
         k5\_6  −0.177019        −0.305 3            Normal distribution       log10    1/µM\*s
         k6\_5  −0.225915          0.15 3            Normal distribution       log10    s\^-1
         k8\_5  −0.998736         0.322 3            Normal distribution       log10    1/µM\*s
         k7\_6  −1.00916         −0.525 3            Normal distribution       log10    1/µM\*s
         k6\_7  −1.653377        −1.745 3            Normal distribution       log10    s\^-1
         k6\_2  1.5293947         1.519 3            Normal distribution       log10    s\^-1
         k5\_8  −3.696929        −3.523 3            Normal distribution       log10    s\^-1
    AKAPoff\_1  −0.531584         0.415 3            Normal distribution       log10    s\^-1
    AKAPoff\_3  1.1736716         1.301 3            Normal distribution       log10    s\^-1
     AKAPon\_1  −0.365851        −0.347 0.279        Normal distribution       log10    s\^-1
     AKAPon\_3  0.2361535         0.301 3            Normal distribution       log10    s\^-1
  kf\_C\_AKAR4  −1.74248         −1.745 0.0969       Normal distribution       log10    1/µM\*s
  kb\_C\_AKAR4  −0.98046         −0.975 0.0969       Normal distribution       log10    s\^-1
   kcat\_AKARp  1.0077982         1.009 0.0969       Normal distribution       log10    s\^-1
         kmOFF  2.0096025             2 0.176        Normal distribution       log10    µM
          kmON  −0.005704             0 0.176        Normal distribution       log10    µM
         KD\_T  −0.175253        −0.155 0.301        Normal distribution       log10    µM
  ------------ ----------- ------------ ---------- ------------------------ ----------- ----------

# Input

These are known parameter values. Parameters that are set by us in
exactly the same way the experimentalost controls the conditions of
the experiment.

  -------- ----------- ----------
  **id**    **value**   **unit**
  b\_AKAP        0          1
  -------- ----------- ----------

# Expression

These are mathematical expression that can be used as abbreviations,
re-used quantities that perhaps appear in several kinetic laws.

  -------- --------------------------------------------------------------------------- ----------------------
    **id**                             **formula**                                     **unit**
    k3p\_7              b\_AKAP\*AKAPon\_1 + (1 - b\_AKAP)\*AKAPoff\_1                 1/s
    k4p\_4              b\_AKAP\*AKAPon\_3 + (1 - b\_AKAP)\*AKAPoff\_3                 1/s
    k4p\_8              b\_AKAP\*AKAPon\_1 + (1 - b\_AKAP)\*AKAPoff\_1                 1/s
    k3p\_3              b\_AKAP\*AKAPon\_3 + (1 - b\_AKAP)\*AKAPoff\_3                 1/s
    k4\_4p  b\_AKAP\*(k4p\_4 + k3p\_7)/kmON + (1 - b\_AKAP)\*(k4p\_4 + k3p\_7)/kmOFF   1/(micromolarity\*s)
    k3\_3p  b\_AKAP\*(k3p\_3 + k4p\_8)/kmON + (1 - b\_AKAP)\*(k3p\_3 + k4p\_8)/kmOFF   1/(micromolarity\*s)
     k2\_1                             k1\_2 \* KD\_T                                  1/s
  -------- --------------------------------------------------------------------------- ----------------------

# Experiment

This is the table of experiments. Each column can name an input or
compound, these value smodify the initial conditions (and inputs) for
that experiment.

+----------------------------------:+:--------:+:-------:+:-----------:+-------:+:-------------:+:-------------:+:------------------------------------------------+
| **id**                            | **cAMP** | **CaN** | **b\_AKAP** | **t0** | **type**      | **event**     | **comment**                                     |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX11\_\_\_\_0nM\_\_TRUE\_\_\_TRUE | 0        | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX12\_\_\_\_0nM\_\_TRUE\_\_FALSE  | 0        | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX13\_\_\_\_0nM\_FALSE\_\_FALSE   | 0        | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX21\_\_100nM\_\_TRUE\_\_\_TRUE   | 0.1      | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX22\_\_100nM\_\_TRUE\_\_FALSE    | 0.1      | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX23\_\_100nM\_FALSE\_\_FALSE     | 0.1      | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX31\_\_200nM\_\_TRUE\_\_\_TRUE   | 0.2      | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX32\_\_200nM\_\_TRUE\_\_FALSE    | 0.2      | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX33\_\_200nM\_FALSE\_\_FALSE     | 0.2      | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX41\_\_500nM\_\_TRUE\_\_\_TRUE   | 0.5      | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX42\_\_500nM\_\_TRUE\_\_FALSE    | 0.5      | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX43\_\_500nM\_FALSE\_\_FALSE     | 0.5      | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX51\_1000nM\_\_TRUE\_\_\_TRUE    | 1        | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX52\_1000nM\_\_TRUE\_\_FALSE     | 1        | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX53\_1000nM\_FALSE\_\_FALSE      | 1        | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX61\_2000nM\_\_TRUE\_\_\_TRUE    | 2        | 1.5     | 1           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX62\_2000nM\_\_TRUE\_\_FALSE     | 2        | 1.5     | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX63\_2000nM\_FALSE\_\_FALSE      | 2        | 0       | 0           | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX71\_\_\_dose\_\_TRUE\_\_\_TRUE  | 0        | 1.5     | 1           | −30    | dose response |               | none                                            |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+
| EX72\_\_event\_\_TRUE\_\_\_TRUE   | 0        | 1.5     | 1           | −30    | time series   | cAMP\_increase| none                                            |
+-----------------------------------+----------+---------+-------------+--------+---------------+---------------+-------------------------------------------------+

# Output

Each row describes an observable, and how to measure it with this
model (as a function of state variables).

  ----------- ------------------------ ----------
  **id**            **formula**        **unit**
  AKAR4pOUT    (AKAR4p\*5)\*71.67+100  µM
  ----------- ------------------------ ----------

# Transformation

Each row describes a transformation, each column names the quantity to
be transformed, any input and compound can be named here. Each cell
contains a formula that calculates the new value. The id can be used
in event schedules to reference this transformation.

  ---------- ----------
  **id**     **cAMP**
  SET\_cAMP   dose
  ---------- ----------

# Event

An experiment can have an event schedule: a timed list of sudden
transformation, each event occurs at a given time, applies a
transformation (by name) and assigns a \_dose\_, the intensity of the
transformation.

  -------- ---------- -------------------- ----------
    **id**  **time**   **transformation**   **dose**
       tr1    610           SET\_cAMP          0.1
       tr2    810           SET\_cAMP          0.2
       tr3    1010          SET\_cAMP          0.5
       tr4    1210          SET\_cAMP          1.0
       tr5    1410          SET\_cAMP          2.0
  -------- ---------- -------------------- ----------

# Data

This is an example data-set. Each measured value is given with the
uncertainty in the same cell. The value `9.994(45)E1` means `99.94`
with standard error of `0.45`. Sometimes this is written as
`99.94±0.45` (whichthe package also accepts). But it's the inferior
syntax for this as it collides with specifying two solutions of some
sort, e.g.: `1 ± sqrt(2)`.

## A Time Series Experiment

+----------:+:--------:+:---------+:--------+
| **id**    | **time** | **AKAR4pOUT**      |
+-----------+----------+--------------------+
| E0103T001 | −15      | 9.994(45)E1        |
+-----------+----------+--------------------+
| E0103T002 | −10      | 9.997(88)E1        |
+-----------+----------+--------------------+
| E0103T003 | −5       | 9.960(31)E1        |
+-----------+----------+--------------------+
| E0103T004 | 0        | NA                 |
+-----------+----------+--------------------+
| E0103T005 | 5        | 9.80(14)E1         |
+-----------+----------+--------------------+
| E0103T006 | 10       | 9.81(19)E1         |
+-----------+----------+--------------------+
| E0103T007 | 15       | 9.701(98)E1        |
+-----------+----------+--------------------+
| E0103T008 | 20       | 9.79(15)E1         |
+-----------+----------+--------------------+
| E0103T009 | 25       | 9.86(11)E1         |
+-----------+----------+--------------------+
| E0103T010 | 30       | 9.95(14)E1         |
+-----------+----------+--------------------+
| E0103T011 | 35       | 9.87(15)E1         |
+-----------+----------+--------------------+
| E0103T012 | 40       | 9.91(14)E1         |
+-----------+----------+--------------------+
| E0103T013 | 45       | 9.93(13)E1         |
+-----------+----------+--------------------+
| E0103T014 | 50       | 9.87(13)E1         |
+-----------+----------+--------------------+
| E0103T015 | 55       | 9.815(68)E1        |
+-----------+----------+--------------------+
| E0103T016 | 60       | 9.8(1)E1           |
+-----------+----------+--------------------+
| E0103T017 | 65       | 9.94(11)E1         |
+-----------+----------+--------------------+
| E0103T018 | 70       | 9.73(11)E1         |
+-----------+----------+--------------------+
| E0103T019 | 75       | 9.769(43)E1        |
+-----------+----------+--------------------+
| E0103T020 | 80       | 9.95(9)E1          |
+-----------+----------+--------------------+
| E0103T021 | 85       | 9.90(14)E1         |
+-----------+----------+--------------------+
| E0103T022 | 90       | 9.7(1)E1           |
+-----------+----------+--------------------+
| E0103T023 | 95       | 9.833(88)E1        |
+-----------+----------+--------------------+
| E0103T024 | 100      | 9.812(29)E1        |
+-----------+----------+--------------------+
| E0103T025 | 105      | 9.883(77)E1        |
+-----------+----------+--------------------+
| E0103T026 | 110      | 1.006(11)E2        |
+-----------+----------+--------------------+
| E0103T027 | 115      | 9.960(78)E1        |
+-----------+----------+--------------------+
| E0103T028 | 120      | 9.91(12)E1         |
+-----------+----------+--------------------+
| E0103T029 | 125      | 1.006(14)E2        |
+-----------+----------+--------------------+
| E0103T030 | 130      | 9.81(15)E1         |
+-----------+----------+--------------------+
| E0103T031 | 135      | 9.99(14)E1         |
+-----------+----------+--------------------+
| E0103T032 | 140      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T033 | 145      | 9.87(15)E1         |
+-----------+----------+--------------------+
| E0103T034 | 150      | 9.80(16)E1         |
+-----------+----------+--------------------+
| E0103T035 | 155      | 9.812(85)E1        |
+-----------+----------+--------------------+
| E0103T036 | 160      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T037 | 165      | 9.948(88)E1        |
+-----------+----------+--------------------+
| E0103T038 | 170      | 9.85(12)E1         |
+-----------+----------+--------------------+
| E0103T039 | 175      | 9.88(14)E1         |
+-----------+----------+--------------------+
| E0103T040 | 180      | 9.90(13)E1         |
+-----------+----------+--------------------+
| E0103T041 | 185      | 9.85(22)E1         |
+-----------+----------+--------------------+
| E0103T042 | 190      | 9.963(71)E1        |
+-----------+----------+--------------------+
| E0103T043 | 195      | 9.948(94)E1        |
+-----------+----------+--------------------+
| E0103T044 | 200      | 9.858(65)E1        |
+-----------+----------+--------------------+
| E0103T045 | 205      | 9.966(69)E1        |
+-----------+----------+--------------------+
| E0103T046 | 210      | 9.88(8)E1          |
+-----------+----------+--------------------+
| E0103T047 | 215      | 9.969(59)E1        |
+-----------+----------+--------------------+
| E0103T048 | 220      | 9.972(14)E1        |
+-----------+----------+--------------------+
| E0103T049 | 225      | 9.809(87)E1        |
+-----------+----------+--------------------+
| E0103T050 | 230      | 9.90(13)E1         |
+-----------+----------+--------------------+
| E0103T051 | 235      | 9.89(14)E1         |
+-----------+----------+--------------------+
| E0103T052 | 240      | 9.96(12)E1         |
+-----------+----------+--------------------+
| E0103T053 | 245      | 9.79(15)E1         |
+-----------+----------+--------------------+
| E0103T054 | 250      | 9.7(2)E1           |
+-----------+----------+--------------------+
| E0103T055 | 255      | 9.849(48)E1        |
+-----------+----------+--------------------+
| E0103T056 | 260      | 9.95(24)E1         |
+-----------+----------+--------------------+
| E0103T057 | 265      | 9.935(86)E1        |
+-----------+----------+--------------------+
| E0103T058 | 270      | 9.82(16)E1         |
+-----------+----------+--------------------+
| E0103T059 | 275      | 9.94(4)E1          |
+-----------+----------+--------------------+
| E0103T060 | 280      | 9.84(13)E1         |
+-----------+----------+--------------------+
| E0103T061 | 285      | 9.97(14)E1         |
+-----------+----------+--------------------+
| E0103T062 | 290      | 9.966(45)E1        |
+-----------+----------+--------------------+
| E0103T063 | 295      | 9.79(14)E1         |
+-----------+----------+--------------------+
| E0103T064 | 300      | 9.93(12)E1         |
+-----------+----------+--------------------+
| E0103T065 | 305      | 9.843(14)E1        |
+-----------+----------+--------------------+
| E0103T066 | 310      | 9.97(17)E1         |
+-----------+----------+--------------------+
| E0103T067 | 315      | 9.84(13)E1         |
+-----------+----------+--------------------+
| E0103T068 | 320      | 9.8(1)E1           |
+-----------+----------+--------------------+
| E0103T069 | 325      | 9.88(12)E1         |
+-----------+----------+--------------------+
| E0103T070 | 330      | 9.827(47)E1        |
+-----------+----------+--------------------+
| E0103T071 | 335      | 9.83(12)E1         |
+-----------+----------+--------------------+
| E0103T072 | 340      | 9.78(16)E1         |
+-----------+----------+--------------------+
| E0103T073 | 345      | 9.988(86)E1        |
+-----------+----------+--------------------+
| E0103T074 | 350      | 10.0(1)E1          |
+-----------+----------+--------------------+
| E0103T075 | 355      | 9.972(51)E1        |
+-----------+----------+--------------------+
| E0103T076 | 360      | 9.923(57)E1        |
+-----------+----------+--------------------+
| E0103T077 | 365      | 9.96(11)E1         |
+-----------+----------+--------------------+
| E0103T078 | 370      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T079 | 375      | 9.88(15)E1         |
+-----------+----------+--------------------+
| E0103T080 | 380      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T081 | 385      | 9.806(28)E1        |
+-----------+----------+--------------------+
| E0103T082 | 390      | 1.0000(61)E2       |
+-----------+----------+--------------------+
| E0103T083 | 395      | 9.92(15)E1         |
+-----------+----------+--------------------+
| E0103T084 | 400      | 9.91(14)E1         |
+-----------+----------+--------------------+
| E0103T085 | 405      | 1.009(14)E2        |
+-----------+----------+--------------------+
| E0103T086 | 410      | 1.0019(77)E2       |
+-----------+----------+--------------------+
| E0103T087 | 415      | 9.985(99)E1        |
+-----------+----------+--------------------+
| E0103T088 | 420      | 1.0000(92)E2       |
+-----------+----------+--------------------+
| E0103T089 | 425      | 9.86(15)E1         |
+-----------+----------+--------------------+
| E0103T090 | 430      | 9.877(48)E1        |
+-----------+----------+--------------------+
| E0103T091 | 435      | 9.99(14)E1         |
+-----------+----------+--------------------+
| E0103T092 | 440      | 9.892(44)E1        |
+-----------+----------+--------------------+
| E0103T093 | 445      | 9.92(13)E1         |
+-----------+----------+--------------------+
| E0103T094 | 450      | 1.006(14)E2        |
+-----------+----------+--------------------+
| E0103T095 | 455      | 1.0086(98)E2       |
+-----------+----------+--------------------+
| E0103T096 | 460      | 9.96(12)E1         |
+-----------+----------+--------------------+
| E0103T097 | 465      | 9.95(11)E1         |
+-----------+----------+--------------------+
| E0103T098 | 470      | 9.932(99)E1        |
+-----------+----------+--------------------+
| E0103T099 | 475      | 9.99(12)E1         |
+-----------+----------+--------------------+
| E0103T100 | 480      | 1.0049(22)E2       |
+-----------+----------+--------------------+
| E0103T101 | 485      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T102 | 490      | 10.0(1)E1          |
+-----------+----------+--------------------+
| E0103T103 | 495      | 1.00(1)E2          |
+-----------+----------+--------------------+
| E0103T104 | 500      | 9.98(12)E1         |
+-----------+----------+--------------------+
| E0103T105 | 505      | 1.0114(32)E2       |
+-----------+----------+--------------------+
| E0103T106 | 510      | 1.007(12)E2        |
+-----------+----------+--------------------+
| E0103T107 | 515      | 9.972(45)E1        |
+-----------+----------+--------------------+
| E0103T108 | 520      | 1.0025(92)E2       |
+-----------+----------+--------------------+
| E0103T109 | 525      | 1.010(15)E2        |
+-----------+----------+--------------------+
| E0103T110 | 530      | 9.98(2)E1          |
+-----------+----------+--------------------+
| E0103T111 | 535      | 1.0114(79)E2       |
+-----------+----------+--------------------+
| E0103T112 | 540      | 1.004(16)E2        |
+-----------+----------+--------------------+
| E0103T113 | 545      | 9.96(4)E1          |
+-----------+----------+--------------------+
| E0103T114 | 550      | 1.0034(82)E2       |
+-----------+----------+--------------------+
| E0103T115 | 555      | 9.95(18)E1         |
+-----------+----------+--------------------+
| E0103T116 | 560      | 1.0022(68)E2       |
+-----------+----------+--------------------+
| E0103T117 | 565      | 9.89(24)E1         |
+-----------+----------+--------------------+
| E0103T118 | 570      | 9.963(69)E1        |
+-----------+----------+--------------------+
| E0103T119 | 575      | 1.010(11)E2        |
+-----------+----------+--------------------+
| E0103T120 | 580      | 9.94(12)E1         |
+-----------+----------+--------------------+
| E0103T121 | 585      | 9.938(92)E1        |
+-----------+----------+--------------------+
| E0103T122 | 590      | 1.0090(94)E2       |
+-----------+----------+--------------------+
| E0103T123 | 595      | 9.94(14)E1         |
+-----------+----------+--------------------+
| E0103T124 | 600      | 1.008(14)E2        |
+-----------+----------+--------------------+
| E0103T125 | 605      | 1.0028(73)E2       |
+-----------+----------+--------------------+

There are many tables like this, one per experiment.

## A Dose Response Experiment

A Dose Response table is a more convenient way of describing a series
of experiments, each with its own, very \_similar\_ conditions, and
exactly one measurement per time-course (a series of micro
experiments). Such a table will be split up by our import script, and
one additional simulation experiment per line will be created. The
newly created simulation-experiments will have very similar names and
vcan be found this way.

  ----------- ---------- ---------- ---------------
  **id**       **cAMP**   **time**    **AKAR4pOUT**
  E0103T125       0         605        1.0028(73)E2
  E0203T125      0.1        605        1.0003(91)E2
  E0403T125      0.2        605           1.10(6)E2
  E0503T125      0.5        605          1.33(12)E2
  E0603T125       1         605        1.5336(97)E2
  E0303T125       2         605         1.675(37)E2
  ----------- ---------- ---------- ---------------

If you don't want these values to be split up into several
micro-experiments, then a very similar effect can be achieved using
events and transformations. See below.

## An Event Schedule Experiment

This is one time series experiment that replicates the same data-set
as the dose reponse curve in the previous section. But, in this
sequence, we also have sudden events that change the `cAMP` value in
the system, and then we allow the model to find a new equilibrium (in
200 seconds).

  ------------ ---------- ---------------
  **id**        **time**    **AKAR4pOUT**
  EVENTEXT01      605        1.0028(73)E2
  EVENTEXT02      805        1.0003(91)E2
  EVENTEXT03      1005          1.10(6)E2
  EVENTEXT04      1205         1.33(12)E2
  EVENTEXT05      1405       1.5336(97)E2
  EVENTEXT06      1605        1.675(37)E2
  ------------ ---------- ---------------

The event schedule we used here is the same as in the section about events:

  -------- ---------- -------------------- ----------
    **id**  **time**   **transformation**   **dose**
       tr1    610           SET\_cAMP          0.1
       tr2    810           SET\_cAMP          0.2
       tr3    1010          SET\_cAMP          0.5
       tr4    1210          SET\_cAMP          1.0
       tr5    1410          SET\_cAMP          2.0
  -------- ---------- -------------------- ----------

These two work together to create a sequence of letting the moment
converge to an equilibrium, recording the value of `AKAR4pOUT`, then
setting a new `cAMP` level and repeat.

