---
title: "Importing Models into R"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Systems Biology Models are usually described with reactions, kinetic
laws, parameters, and possibly inputs and outputs.

In this R package, we use plain TSV files to describe these
quantities. These TSV files have very few rules. One universal rule is
that the first column is always reserved for row-names. Row-names must
work as variable names in R and C. Certan names would conflict with
builtin commands (`if`, `while`, `do`, `continue`, `break`). Obviously
such names cannot be used for reacting compounds or parameters.  In
our examples, we label the first column with `id` (but it can have any
name).

- A list of quantities, like parameters have these columns:
  + id (first column, any heading)
  + value
  + unit
  + scale (optional, assumed to be linear)
- A list of mathematical expressions can use the term `formula`
  instead of `value`:
  + id
  + formula
  + unit
- Any quantity that is defined using a row in some table can get
  updated values in another table, or values to comapre against
- Data tables need to specifiy values using a notation that keeps
  values and their uncertainties together, e.g.:
  + `1.234(21)`
  + `1.5(1)e+4`
- The tables are found by name, the names our import scripts will be looking for are:
  + Constant.tsv
  + Parameter.tsv
  + Input.tsv
  + Expression.tsv
  + Reaction.tsv
  + Output.tsv
  + Experiment.tsv
- The Experiment table describes the experimental conditions the model
  needs to be subjected to in order to replicate measured data:
  + each experimental setting gets a row
  + each row id is used to find the data
  + the data is found in the TSV file named like the row-id
  + Data for experiment `EX1` is expected to be in the file `EX1.tsv`

In the following sections, we give examples for each of the tables (as
markdown tables):

# Reaction

+-------:+:--------------------------------------------:+----------------:+:-------:+:---------------+:---------------------+
| **id** | **kinetic law**                              | **reactants**   | **arw** | **products**   | **is reversible**    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r51    | k5_1\*Rii_C                                  | Rii_C           | \<=\>   | RiiP_C         | 0                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r14    | k4_1\*RiiP\*C - k1_4\*RiiP_C                 | RiiP + C        | \<=\>   | RiiP_C         | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r12    | k1_2\*RiiP_C\*cAMP - k2_1\*RiiP_C_cAMP       | RiiP_C + cAMP   | \<=\>   | RiiP_C_cAMP    | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r43    | k4_3\*cAMP\*RiiP - k3_4\*RiiP_cAMP           | cAMP + RiiP     | \<=\>   | RiiP_cAMP      | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r23    | k3_2\*RiiP_cAMP\*C - k2_3\*RiiP_C_cAMP       | RiiP_cAMP + C   | \<=\>   | RiiP_C_cAMP    | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r78    | k8_7\*cAMP\*Rii - k7_8\*Rii_cAMP             | cAMP + Rii      | \<=\>   | Rii_cAMP       | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r56    | k5_6\*Rii_C\*cAMP - k6_5\*Rii_C_cAMP         | Rii_C + cAMP    | \<=\>   | Rii_C_cAMP     | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r76    | k7_6\*Rii_cAMP\*C - k6_7\*Rii_C_cAMP         | Rii_cAMP + C    | \<=\>   | Rii_C_cAMP     | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r62    | k6_2\*Rii_C_cAMP                             | Rii_C_cAMP      | \<=\>   | RiiP_C_cAMP    | 0                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r58    | k8_5\*Rii\*C - k5_8\*Rii_C                   | Rii + C         | \<=\>   | Rii_C          | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r44    | k4_4p\*RiiP\*CaN - k4p_4\*RiiP_CaN           | RiiP + CaN      | \<=\>   | RiiP_CaN       | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r33    | k3_3p\*CaN\*RiiP_cAMP - k3p_3\*RiiP_cAMP_CaN | CaN + RiiP_cAMP | \<=\>   | RiiP_cAMP_CaN  | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r48    | k4p_8\*RiiP_CaN                              | RiiP_CaN        | \<=\>   | Rii + CaN      | 0                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r37    | k3p_7\*RiiP_cAMP_CaN                         | RiiP_cAMP_CaN   | \<=\>   | CaN + Rii_cAMP | 0                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r1     | kf_C_AKAR4\*C\*AKAR4 - kb_C_AKAR4\*AKAR4_C   | C + AKAR4       | \<=\>   | AKAR4_C        | 1                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+
| r2     | kcat_AKARp\*AKAR4_C                          | AKAR4_C         | \<=\>   | AKAR4p + C     | 0                    |
+--------+----------------------------------------------+-----------------+---------+----------------+----------------------+

: Reaction {cellspacing="0" cellpadding="3"}

# Compound

Reacting compounds, or molecular species. These will be turned into state variables.

  --------------- ----------- ---------- ------------- --------------------------------------------------
  **id**          **value**   **Unit**   **uniprot**   **Comment**
  Rii             6.3         µM         KAP2_HUMAN    also KAP3_HUMAN because Rii = RII_alpha+RII_beta
  cAMP            0           µM         NONE          NONE
  RiiP            0           µM         NONE          phosphorylated
  Rii_C           0.63        µM         NONE          NONE
  RiiP_cAMP       0           µM         NONE          NONE
  RiiP_C          0           µM         NONE          NONE
  RiiP_C_cAMP     0           µM         NONE          NONE
  C               0           µM         KAPCB_HUMAN   NONE
  Rii_cAMP        0           µM         NONE          NONE
  Rii_C_cAMP      0           µM         NONE          NONE
  CaN             1.5         µM         PP2BA_HUMAN   PP3R1_HUMAN
  RiiP_CaN        0           µM         NONE          NONE
  RiiP_cAMP_CaN   0           µM         NONE          NONE
  AKAR4           0.2         µM         NONE          RRID:addgene_61619
  AKAR4_C         0           µM         NONE          NONE
  AKAR4p          0           µM         NONE          NONE
  --------------- ----------- ---------- ------------- --------------------------------------------------

  : Compound {cellspacing="0" cellpadding="3"}

# Parameter

These are parameters, that will be subject to optimization or
sampling. The values are assumed to be crude estimates.

Known parameters should go into a different table.

  ------------ ----------- ------------ ---------- ------------------------ ----------- ----------
        **id**  **value**    **median** **stdv**    **prior distribution**   **scale**  **unit**
          k5_1  1.667837          1.519 3            Normal distribution       log10    s\^-1
          k1_2  0.0946379        −0.306 3            Normal distribution       log10    1/µM\*s
          k3_2  −2.483545        −2.264 3            Normal distribution       log10    1/µM\*s
          k2_3  −0.230731        −1.807 3            Normal distribution       log10    s\^-1
          k3_4  −1.136198        −2.796 3            Normal distribution       log10    s\^-1
          k4_3  −2.147382        −1.824 3            Normal distribution       log10    1/µM\*s
          k4_1  −2.284185         −1.42 3            Normal distribution       log10    1/µM\*s
          k1_4  −2.928144        −2.585 3            Normal distribution       log10    s\^-1
          k8_7  −2.149449        −1.824 3            Normal distribution       log10    1/µM\*s
          k7_8  −1.07931         −2.796 3            Normal distribution       log10    s\^-1
          k5_6  −0.177019        −0.305 3            Normal distribution       log10    1/µM\*s
          k6_5  −0.225915          0.15 3            Normal distribution       log10    s\^-1
          k8_5  −0.998736         0.322 3            Normal distribution       log10    1/µM\*s
          k7_6  −1.00916         −0.525 3            Normal distribution       log10    1/µM\*s
          k6_7  −1.653377        −1.745 3            Normal distribution       log10    s\^-1
          k6_2  1.5293947         1.519 3            Normal distribution       log10    s\^-1
          k5_8  −3.696929        −3.523 3            Normal distribution       log10    s\^-1
     AKAPoff_1  −0.531584         0.415 3            Normal distribution       log10    s\^-1
     AKAPoff_3  1.1736716         1.301 3            Normal distribution       log10    s\^-1
      AKAPon_1  −0.365851        −0.347 0.279        Normal distribution       log10    s\^-1
      AKAPon_3  0.2361535         0.301 3            Normal distribution       log10    s\^-1
    kf_C_AKAR4  −1.74248         −1.745 0.0969       Normal distribution       log10    1/µM\*s
    kb_C_AKAR4  −0.98046         −0.975 0.0969       Normal distribution       log10    s\^-1
    kcat_AKARp  1.0077982         1.009 0.0969       Normal distribution       log10    s\^-1
         kmOFF  2.0096025             2 0.176        Normal distribution       log10    µM
          kmON  −0.005704             0 0.176        Normal distribution       log10    µM
          KD_T  −0.175253        −0.155 0.301        Normal distribution       log10    µM
  ------------ ----------- ------------ ---------- ------------------------ ----------- ----------

  : Parameter {cellspacing="0" cellpadding="3"}

# Input

These are known parameter values. Parameters that are set by us in
exactly the same way the experimentalost controls the conditions of
the experiment.

  -------- ----------- ----------
  **id**    **value**   **unit**
  b_AKAP        0          1
  -------- ----------- ----------

  : Input {cellspacing="0" cellpadding="3"}

# Expression

These are mathematical expression that can be used as abbreviations,
re-used quantities that perhaps appear in several kinetic laws.

  -------- -------------------------------------------------------------------- ----------------------
    **id**                             **formula**                                     **unit**
     k3p_7              b_AKAP\*AKAPon_1 + (1 - b_AKAP)\*AKAPoff_1                       1/s
     k4p_4              b_AKAP\*AKAPon_3 + (1 - b_AKAP)\*AKAPoff_3                       1/s
     k4p_8              b_AKAP\*AKAPon_1 + (1 - b_AKAP)\*AKAPoff_1                       1/s
     k3p_3              b_AKAP\*AKAPon_3 + (1 - b_AKAP)\*AKAPoff_3                       1/s
     k4_4p  b_AKAP\*(k4p_4 + k3p_7)/kmON + (1 - b_AKAP)\*(k4p_4 + k3p_7)/kmOFF   1/(micromolarity\*s)
     k3_3p  b_AKAP\*(k3p_3 + k4p_8)/kmON + (1 - b_AKAP)\*(k3p_3 + k4p_8)/kmOFF   1/(micromolarity\*s)
      k2_1                             k1_2 \* KD_T                                      1/s
  -------- -------------------------------------------------------------------- ----------------------

  : Expression {cellspacing="0" cellpadding="3"}

# Experiment

This is the table of experiments. Each column can name an input or
compound, these value smodify the initial conditions (and inputs) for
that experiment.

+----------------------------------:+:--------:+:-------:+:----------:+-------:+:-------------:+:-------------:+:------------------------------------------------+
| **id**                            | **cAMP** | **CaN** | **b_AKAP** | **t0** | **type**      | **event**     | **comment**                                     |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX11\_\_\_\_0nM\_\_TRUE\_\_\_TRUE | 0        | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX12\_\_\_\_0nM\_\_TRUE\_\_FALSE  | 0        | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX13\_\_\_\_0nM_FALSE\_\_FALSE    | 0        | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX21\_\_100nM\_\_TRUE\_\_\_TRUE   | 0.1      | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX22\_\_100nM\_\_TRUE\_\_FALSE    | 0.1      | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX23\_\_100nM_FALSE\_\_FALSE      | 0.1      | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX31\_\_200nM\_\_TRUE\_\_\_TRUE   | 0.2      | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX32\_\_200nM\_\_TRUE\_\_FALSE    | 0.2      | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX33\_\_200nM_FALSE\_\_FALSE      | 0.2      | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX41\_\_500nM\_\_TRUE\_\_\_TRUE   | 0.5      | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX42\_\_500nM\_\_TRUE\_\_FALSE    | 0.5      | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX43\_\_500nM_FALSE\_\_FALSE      | 0.5      | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX51_1000nM\_\_TRUE\_\_\_TRUE     | 1        | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX52_1000nM\_\_TRUE\_\_FALSE      | 1        | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX53_1000nM_FALSE\_\_FALSE        | 1        | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX61_2000nM\_\_TRUE\_\_\_TRUE     | 2        | 1.5     | 1          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX62_2000nM\_\_TRUE\_\_FALSE      | 2        | 1.5     | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX63_2000nM_FALSE\_\_FALSE        | 2        | 0       | 0          | −30    | time series   |               | Figure 3 in https://doi.org/10.7554/eLife.68164 |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX71\_\_\_dose\_\_TRUE\_\_\_TRUE  | 0        | 1.5     | 1          | −30    | dose response |               | none                                            |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+
| EX72\_\_event\_\_TRUE\_\_\_TRUE   | 0        | 1.5     | 1          | −30    | time series   | cAMP_increase | none                                            |
+-----------------------------------+----------+---------+------------+--------+---------------+---------------+-------------------------------------------------+

: Experiments {cellspacing="0" cellpadding="3"}

# Output

Each row describes an observable, and how to measure it with this
model (as a function of state variables).

  ----------- ------------------------ ----------
  **id**            **formula**        **unit**
  AKAR4pOUT    (AKAR4p\*5)\*71.67+100  µM
  ----------- ------------------------ ----------

  : Output {cellspacing="0" cellpadding="3"}


# Transformation

Each row describes a transformation, each column names the quantity to
be transformed, any input and compound can be named here. Each cell
contains a formula that calculates the new value. The id can be used
in event schedules to reference this transformation.

  ---------- ----------
  **id**     **cAMP**
  SET_cAMP   dose
  ---------- ----------

  : Transformation {cellspacing="0" cellpadding="3"}

# Event

An experiment can have an event schedule: a timed list of sudden
transformation, each event occurs at a given time, applies a
transformation (by name) and assigns a _dose_, the intensity of the
transformation.

  -------- ---------- -------------------- ----------
    **id**  **time**   **transformation**   **dose**
       tr1    610           SET_cAMP          0.1
       tr2    810           SET_cAMP          0.2
       tr3    1010          SET_cAMP          0.5
       tr4    1210          SET_cAMP          1.0
       tr5    1410          SET_cAMP          2.0
  -------- ---------- -------------------- ----------

  : cAMP_increase {cellspacing="0" cellpadding="3"}

# Data

This is an example data-set. Each measured value is given with the
uncertainty in the same cell. The value `9.994(45)E1` means `99.94`
with standard error of `0.45`. Sometimes this is written as
`99.94±0.45` (whichthe package also accepts). But it's the inferior
syntax for this as it collides with specifying two solutions of some
sort, e.g.: `1 ± sqrt(2)`.

## A Time Series Experiment

+----------:+:--------:+:---------+:--------+
| **id**    | **time** | **AKAR4pOUT**      |
+-----------+----------+--------------------+
| E0103T001 | −15      | 9.994(45)E1        |
+-----------+----------+--------------------+
| E0103T002 | −10      | 9.997(88)E1        |
+-----------+----------+--------------------+
| E0103T003 | −5       | 9.960(31)E1        |
+-----------+----------+--------------------+
| E0103T004 | 0        | NA                 |
+-----------+----------+--------------------+
| E0103T005 | 5        | 9.80(14)E1         |
+-----------+----------+--------------------+
| E0103T006 | 10       | 9.81(19)E1         |
+-----------+----------+--------------------+
| E0103T007 | 15       | 9.701(98)E1        |
+-----------+----------+--------------------+
| E0103T008 | 20       | 9.79(15)E1         |
+-----------+----------+--------------------+
| E0103T009 | 25       | 9.86(11)E1         |
+-----------+----------+--------------------+
| E0103T010 | 30       | 9.95(14)E1         |
+-----------+----------+--------------------+
| E0103T011 | 35       | 9.87(15)E1         |
+-----------+----------+--------------------+
| E0103T012 | 40       | 9.91(14)E1         |
+-----------+----------+--------------------+
| E0103T013 | 45       | 9.93(13)E1         |
+-----------+----------+--------------------+
| E0103T014 | 50       | 9.87(13)E1         |
+-----------+----------+--------------------+
| E0103T015 | 55       | 9.815(68)E1        |
+-----------+----------+--------------------+
| E0103T016 | 60       | 9.8(1)E1           |
+-----------+----------+--------------------+
| E0103T017 | 65       | 9.94(11)E1         |
+-----------+----------+--------------------+
| E0103T018 | 70       | 9.73(11)E1         |
+-----------+----------+--------------------+
| E0103T019 | 75       | 9.769(43)E1        |
+-----------+----------+--------------------+
| E0103T020 | 80       | 9.95(9)E1          |
+-----------+----------+--------------------+
| E0103T021 | 85       | 9.90(14)E1         |
+-----------+----------+--------------------+
| E0103T022 | 90       | 9.7(1)E1           |
+-----------+----------+--------------------+
| E0103T023 | 95       | 9.833(88)E1        |
+-----------+----------+--------------------+
| E0103T024 | 100      | 9.812(29)E1        |
+-----------+----------+--------------------+
| E0103T025 | 105      | 9.883(77)E1        |
+-----------+----------+--------------------+
| E0103T026 | 110      | 1.006(11)E2        |
+-----------+----------+--------------------+
| E0103T027 | 115      | 9.960(78)E1        |
+-----------+----------+--------------------+
| E0103T028 | 120      | 9.91(12)E1         |
+-----------+----------+--------------------+
| E0103T029 | 125      | 1.006(14)E2        |
+-----------+----------+--------------------+
| E0103T030 | 130      | 9.81(15)E1         |
+-----------+----------+--------------------+
| E0103T031 | 135      | 9.99(14)E1         |
+-----------+----------+--------------------+
| E0103T032 | 140      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T033 | 145      | 9.87(15)E1         |
+-----------+----------+--------------------+
| E0103T034 | 150      | 9.80(16)E1         |
+-----------+----------+--------------------+
| E0103T035 | 155      | 9.812(85)E1        |
+-----------+----------+--------------------+
| E0103T036 | 160      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T037 | 165      | 9.948(88)E1        |
+-----------+----------+--------------------+
| E0103T038 | 170      | 9.85(12)E1         |
+-----------+----------+--------------------+
| E0103T039 | 175      | 9.88(14)E1         |
+-----------+----------+--------------------+
| E0103T040 | 180      | 9.90(13)E1         |
+-----------+----------+--------------------+
| E0103T041 | 185      | 9.85(22)E1         |
+-----------+----------+--------------------+
| E0103T042 | 190      | 9.963(71)E1        |
+-----------+----------+--------------------+
| E0103T043 | 195      | 9.948(94)E1        |
+-----------+----------+--------------------+
| E0103T044 | 200      | 9.858(65)E1        |
+-----------+----------+--------------------+
| E0103T045 | 205      | 9.966(69)E1        |
+-----------+----------+--------------------+
| E0103T046 | 210      | 9.88(8)E1          |
+-----------+----------+--------------------+
| E0103T047 | 215      | 9.969(59)E1        |
+-----------+----------+--------------------+
| E0103T048 | 220      | 9.972(14)E1        |
+-----------+----------+--------------------+
| E0103T049 | 225      | 9.809(87)E1        |
+-----------+----------+--------------------+
| E0103T050 | 230      | 9.90(13)E1         |
+-----------+----------+--------------------+
| E0103T051 | 235      | 9.89(14)E1         |
+-----------+----------+--------------------+
| E0103T052 | 240      | 9.96(12)E1         |
+-----------+----------+--------------------+
| E0103T053 | 245      | 9.79(15)E1         |
+-----------+----------+--------------------+
| E0103T054 | 250      | 9.7(2)E1           |
+-----------+----------+--------------------+
| E0103T055 | 255      | 9.849(48)E1        |
+-----------+----------+--------------------+
| E0103T056 | 260      | 9.95(24)E1         |
+-----------+----------+--------------------+
| E0103T057 | 265      | 9.935(86)E1        |
+-----------+----------+--------------------+
| E0103T058 | 270      | 9.82(16)E1         |
+-----------+----------+--------------------+
| E0103T059 | 275      | 9.94(4)E1          |
+-----------+----------+--------------------+
| E0103T060 | 280      | 9.84(13)E1         |
+-----------+----------+--------------------+
| E0103T061 | 285      | 9.97(14)E1         |
+-----------+----------+--------------------+
| E0103T062 | 290      | 9.966(45)E1        |
+-----------+----------+--------------------+
| E0103T063 | 295      | 9.79(14)E1         |
+-----------+----------+--------------------+
| E0103T064 | 300      | 9.93(12)E1         |
+-----------+----------+--------------------+
| E0103T065 | 305      | 9.843(14)E1        |
+-----------+----------+--------------------+
| E0103T066 | 310      | 9.97(17)E1         |
+-----------+----------+--------------------+
| E0103T067 | 315      | 9.84(13)E1         |
+-----------+----------+--------------------+
| E0103T068 | 320      | 9.8(1)E1           |
+-----------+----------+--------------------+
| E0103T069 | 325      | 9.88(12)E1         |
+-----------+----------+--------------------+
| E0103T070 | 330      | 9.827(47)E1        |
+-----------+----------+--------------------+
| E0103T071 | 335      | 9.83(12)E1         |
+-----------+----------+--------------------+
| E0103T072 | 340      | 9.78(16)E1         |
+-----------+----------+--------------------+
| E0103T073 | 345      | 9.988(86)E1        |
+-----------+----------+--------------------+
| E0103T074 | 350      | 10.0(1)E1          |
+-----------+----------+--------------------+
| E0103T075 | 355      | 9.972(51)E1        |
+-----------+----------+--------------------+
| E0103T076 | 360      | 9.923(57)E1        |
+-----------+----------+--------------------+
| E0103T077 | 365      | 9.96(11)E1         |
+-----------+----------+--------------------+
| E0103T078 | 370      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T079 | 375      | 9.88(15)E1         |
+-----------+----------+--------------------+
| E0103T080 | 380      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T081 | 385      | 9.806(28)E1        |
+-----------+----------+--------------------+
| E0103T082 | 390      | 1.0000(61)E2       |
+-----------+----------+--------------------+
| E0103T083 | 395      | 9.92(15)E1         |
+-----------+----------+--------------------+
| E0103T084 | 400      | 9.91(14)E1         |
+-----------+----------+--------------------+
| E0103T085 | 405      | 1.009(14)E2        |
+-----------+----------+--------------------+
| E0103T086 | 410      | 1.0019(77)E2       |
+-----------+----------+--------------------+
| E0103T087 | 415      | 9.985(99)E1        |
+-----------+----------+--------------------+
| E0103T088 | 420      | 1.0000(92)E2       |
+-----------+----------+--------------------+
| E0103T089 | 425      | 9.86(15)E1         |
+-----------+----------+--------------------+
| E0103T090 | 430      | 9.877(48)E1        |
+-----------+----------+--------------------+
| E0103T091 | 435      | 9.99(14)E1         |
+-----------+----------+--------------------+
| E0103T092 | 440      | 9.892(44)E1        |
+-----------+----------+--------------------+
| E0103T093 | 445      | 9.92(13)E1         |
+-----------+----------+--------------------+
| E0103T094 | 450      | 1.006(14)E2        |
+-----------+----------+--------------------+
| E0103T095 | 455      | 1.0086(98)E2       |
+-----------+----------+--------------------+
| E0103T096 | 460      | 9.96(12)E1         |
+-----------+----------+--------------------+
| E0103T097 | 465      | 9.95(11)E1         |
+-----------+----------+--------------------+
| E0103T098 | 470      | 9.932(99)E1        |
+-----------+----------+--------------------+
| E0103T099 | 475      | 9.99(12)E1         |
+-----------+----------+--------------------+
| E0103T100 | 480      | 1.0049(22)E2       |
+-----------+----------+--------------------+
| E0103T101 | 485      | 9.9(1)E1           |
+-----------+----------+--------------------+
| E0103T102 | 490      | 10.0(1)E1          |
+-----------+----------+--------------------+
| E0103T103 | 495      | 1.00(1)E2          |
+-----------+----------+--------------------+
| E0103T104 | 500      | 9.98(12)E1         |
+-----------+----------+--------------------+
| E0103T105 | 505      | 1.0114(32)E2       |
+-----------+----------+--------------------+
| E0103T106 | 510      | 1.007(12)E2        |
+-----------+----------+--------------------+
| E0103T107 | 515      | 9.972(45)E1        |
+-----------+----------+--------------------+
| E0103T108 | 520      | 1.0025(92)E2       |
+-----------+----------+--------------------+
| E0103T109 | 525      | 1.010(15)E2        |
+-----------+----------+--------------------+
| E0103T110 | 530      | 9.98(2)E1          |
+-----------+----------+--------------------+
| E0103T111 | 535      | 1.0114(79)E2       |
+-----------+----------+--------------------+
| E0103T112 | 540      | 1.004(16)E2        |
+-----------+----------+--------------------+
| E0103T113 | 545      | 9.96(4)E1          |
+-----------+----------+--------------------+
| E0103T114 | 550      | 1.0034(82)E2       |
+-----------+----------+--------------------+
| E0103T115 | 555      | 9.95(18)E1         |
+-----------+----------+--------------------+
| E0103T116 | 560      | 1.0022(68)E2       |
+-----------+----------+--------------------+
| E0103T117 | 565      | 9.89(24)E1         |
+-----------+----------+--------------------+
| E0103T118 | 570      | 9.963(69)E1        |
+-----------+----------+--------------------+
| E0103T119 | 575      | 1.010(11)E2        |
+-----------+----------+--------------------+
| E0103T120 | 580      | 9.94(12)E1         |
+-----------+----------+--------------------+
| E0103T121 | 585      | 9.938(92)E1        |
+-----------+----------+--------------------+
| E0103T122 | 590      | 1.0090(94)E2       |
+-----------+----------+--------------------+
| E0103T123 | 595      | 9.94(14)E1         |
+-----------+----------+--------------------+
| E0103T124 | 600      | 1.008(14)E2        |
+-----------+----------+--------------------+
| E0103T125 | 605      | 1.0028(73)E2       |
+-----------+----------+--------------------+

There are many tables like this, one per experiment.

## A Dose Response Experiment

A Dose Response table is a more convenient way of describing a series
of experiments, each with its own, very _similar_ conditions, and
exactly one measurement per time-course (a series of micro
experiments). Such a table will be split up by our import script, and
one additional simulation experiment per line will be created. The
newly created simulation-experiments will have very similar names and
vcan be found this way.

  ----------- ---------- ---------- ---------------
  **id**       **cAMP**   **time**    **AKAR4pOUT**
  E0103T125       0         605        1.0028(73)E2
  E0203T125      0.1        605        1.0003(91)E2
  E0403T125      0.2        605           1.10(6)E2
  E0503T125      0.5        605          1.33(12)E2
  E0603T125       1         605        1.5336(97)E2
  E0303T125       2         605         1.675(37)E2
  ----------- ---------- ---------- ---------------

  : EX71\_\_\_dose\_\_TRUE\_\_\_TRUE {cellspacing="0" cellpadding="3"}

If you don't want these values to be split up into several
micro-experiments, then a very similar effect can be achieved using
events and transformations. See below.

## An Event Schedule Experiment

This is one time series experiment that replicates the same data-set
as the dose reponse curve in the previous section. But, in this
sequence, we also have sudden events that change the `cAMP` value in
the system, and then we allow the model to find a new equilibrium (in
200 seconds).

  ------------ ---------- ---------------
  **id**        **time**    **AKAR4pOUT**
  EVENTEXT01      605        1.0028(73)E2
  EVENTEXT02      805        1.0003(91)E2
  EVENTEXT03      1005          1.10(6)E2
  EVENTEXT04      1205         1.33(12)E2
  EVENTEXT05      1405       1.5336(97)E2
  EVENTEXT06      1605        1.675(37)E2
  ------------ ---------- ---------------

  : EX72\_\_event\_\_TRUE\_\_\_TRUE {cellspacing="0" cellpadding="3"}


The event schedule we used here is the same as in the section about events:

  -------- ---------- -------------------- ----------
    **id**  **time**   **transformation**   **dose**
       tr1    610           SET_cAMP          0.1
       tr2    810           SET_cAMP          0.2
       tr3    1010          SET_cAMP          0.5
       tr4    1210          SET_cAMP          1.0
       tr5    1410          SET_cAMP          2.0
  -------- ---------- -------------------- ----------

  : cAMP_increase {cellspacing="0" cellpadding="3"}

These two work together to create a sequence of letting the moment
converge to an equilibrium, recording the value of `AKAR4pOUT`, then
setting a new `cAMP` level and repeat.

# OLD SBtab content


If the model can be described in terms of reactions, inputs, and
outputs, then the most straightforward way to define it is through
[SBtab tables](./SBtab.html).^[For completely different models (or solvers), the user must either write their own likelihood function (cost-function/objective-function) or write a .vf file manually, to generate C code from.] ^[It is also possible to directly write the model functions in C (for `gsl_odeiv2`).]

The following R commands load the SBtab model and data into R as `data.frames`, and convert the model into an ODE. First, you will need to get to the directory with the SBtab files by the command `setwd(...)`.

```{r init, eval=FALSE}
f <- dir(pattern="[.]tsv$",full.names=TRUE)          # file names
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
ml <- SBtabVFGEN::sbtab_to_vfgen(sb)                 # ODE model (as a list)
ex <- SBtabVFGEN::sbtab.data(sb,ml$conservationLaws) # includes the data
C <- uqsa::generateCode(ml)                          # C Code
cat(C, sep = "\n", file = "AKAR4cl_gvf.c")

```

Some of the R commands above belong to our external package [SBtabVFGEN](https://github.com/icpm-kth/SBtabVFGEN). 

More information on the R commands used in the previous code block, and further instructions on building and simulating your own model can be found in the pages ["Build and simulate your own model"](user_model.html) and ["Importing Data"](data.html).

The rest of this page describes the variables and the files that we just created through the previous R commands.

---

# Real Example AKAR4

```{r init-evaluation, eval=TRUE, include=FALSE}
library(uqsa)
library(SBtabVFGEN)

f <- uqsa_example('AKAR4')                           # file names
print(f[1])
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
ml <- SBtabVFGEN::sbtab_to_vfgen(sb)                 # model, as character vector
cl <- ml$conservationLaws                            # with conservation laws (if any)
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # list of simulation experiments
```

```{r init-display, eval=FALSE, include=TRUE}
library(uqsa)
library(SBtabVFGEN)

f <- uqsa_example('AKAR4')                           # file names
print(f[1])
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
ml <- SBtabVFGEN::sbtab_to_vfgen(sb)                 # model, as character vector
cl <- ml$conservationLaws                            # with conservation laws (if any)
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # list of simulation experiments
```

The `ex` variable holds simulation instructions for the different
experiments as an R list that includes information on:

- inputs,
- initial values,
- measurement times,
- scheduled events,
- measured data.

SBtab contents include:

Reactions and conservation laws
```{r sb-reactions, eval=TRUE}
print(sb$Reaction[,c(2,3)])
cat(cl$Text,sep="\n")
```

The measured data, e.g. for experiment `1`:
```{r sb-data, eval=TRUE}
head(ex[[1]]$outputValues); cat("[...]\n"); tail(ex[[1]]$outputValues)
```

The standard error of the data:
```{r sb-err, eval=TRUE}
head(ex[[1]]$errorValues); cat("[...]\n"); tail(ex[[1]]$errorValues)
```

Data as a plot:
```{r error-bars, eval=TRUE}
i <- seq(70)
t <- ex[[1]]$outputTimes[i]
y <- ex[[1]]$outputValues[[1]][i]
dy <- ex[[1]]$errorValues[[1]][i]

if (require("errors")){
    t <- set_errors(t,0)
    y <- set_errors(y,dy)
    plot(t,y)
} else {
    plot(t,y,bty='n',ylim=c(100,200))
    arrows(t,y,t,y+dy,angle=90, length=0.01)
    arrows(t,y,t,y-dy,angle=90, length=0.01)
}
```

# Files

The function `sbtab_to_vfgen` will produce several files, with the
model's name (`Document='model_name'` attribute of SBtab) and different extensions:

- `model_name.vf`, a vfgen compatible file
- `model_name.mod`, a NEURON mod file
- (optional) `model_name.xml`, an SBML file, if you have the R bindings to libsbml installed in R and libsbml in your system
- `model_name.tar.gz`, compressed archive of a text representation of all ODE elements, no specific format
- `model_name.zip`, the same as `tar.gz`, but as a zip archive

Let's look at the state variables of the ode:

```{sh, label='show the ODE right hand side',eval=TRUE}
grep '<StateVariable' AKAR4.vf
```

Two State variables have been commented out (through conservation law analysis). These two have been replaced by vfgen expressions:

```{sh, label='show Expressions',eval=TRUE}
grep '<Expression' AKAR4.vf
```

Finally, we clean up the generated files:

```{sh, eval=TRUE, include=FALSE}
for t in vf tar.gz zip ; do rm ./AKAR4.$t ; done
[ -f AKAR4.xml ] && rm AKAR4.xml # since that one is optional
echo $?
```

```{sh, eval=FALSE}
for t in vf tar.gz zip ; do rm ./AKAR4.$t ; done
[ -f AKAR4.xml ] && rm AKAR4.xml # since that one is optional
echo $?
```

## The Vector Field file

The vf file is compatible with the
[vfgen](https://warrenweckesser.github.io/vfgen/) tool, which creates
code in many languages from this `vf` file (MATLAB, CVODE, Octave,
Python, XPPAUT). Use this tool if you want to use the created source
code for other programming languages. Some SBtab content will be
missing here (e.g., scheduled events). The vfgen tool uses the
[GiNaC](https://www.ginac.de/) library to calculate Jacobians. GiNaC
cannot parse vf files with inequality expressions in them (e.g. `a < b`).

### Converting the ODE to code

We use our tool, [RPN-derivative](github.com/icpm-kth/RPN-derivative),
which reads the same file (or one of the archive files:
`zip`/`tar.gz`) to create slightly different C code. The main
advantages are:

- a vector valued output function with a predictable name: `int MODEL_func()`
- the error code of each function is the length of the output buffers that need to be allocated
- more functions are created:
    + initial conditions: `int MODEL_init()`
	+ default parameter values: `int MODEL_default()`

The repository of this tool, [icpm-kth/RPN-derivative](icpm-kth/RPN-derivative),
contains a shell script [ode.sh](icpm-kth/RPN/derivative/sh/ode.sh);
this script uses one of three methods to calculate analytical
derivatives for the Jacobians:
1. [maxima](https://maxima.sourceforge.io/),
2. [yacas](https://yacas.readthedocs.io/en/latest/),
3. the `bin/derivative` binary that RPN-derivative contains (after `make && make install`)

Assuming that you followed all [installation instructions](articles/installation.html):

```{sh, eval=FALSE}
ode -R --maxima ./myModel.vf > ./myModel.R 
##     both     ./myModel.{zip,tar.gz}     also work here
```

```{sh, eval=FALSE}
ode -C --maxima ./myModel.tar.gz > ./myModel_gvf.c
```

Note: `ode.sh` is compatible with the same vf files from SBtabVFGEN,
but doesn't support all of the output formats that VFGEN can do, and
does nothing related to delays.

# Background on Parameters vs Inputs

Our goal is to perform *parameter estimation*. A *systems biology*
model typically has parameters that are either reaction rate
coefficients $k_f, k_b$, equilibrium constants $K_d$, Hill coefficients,
and many other quantities related to *kinetc laws*. Parameters are often
unknown or not uniquely determined yet. However, not all parameters of a
model are unknown, and not all parameters are intrinsic to the
system we study. Some of them describe the interventions we do to the
system when performing an experiment. This could be the frequency of a
driving force for a mechanical model, or the amount of a treatment
dose (added to the system) that affects the time course of reactions:

- a substrate, or buffer
- an enzyme,
- a silencing agent,
- an inhibitor

These parameters are known to us, because they are written down in the
protocol. If the input is time dependent, then its dynamics (or
explicit algebraic functions) have to become part of the model. These
input parameters can be different between experiments, but the
intrinsic parameters (e.g., reaction rate coefficients) are always the same.

The ordinary differential equation, on the other hand, does not need to
know the distinction between known or unknown parameters. For these
reasons, we may write:

```R
parMCMC <- [...]             # some sampling variable
k <- parMap(parMCMC)         # a model compatible parameter vector
u <- experiments[[i]]$input  # a valid model input vector
p <- c(k,u)
# solve ODE using p
```

where `k` are intrinsic, unknown parameters (of which we quantify the uncertainty via sampling algorithms in UQSA), and
`u` are the known input parameters that are encoded in the `input`
field of an experiment.

If input parameters `u` exist, then they are always concatenated (`p <-c(k,u)`) 
in that order (after intrinsic, unknown parameters) and passed to the model. The model's C code only sees `p`. The wrapper functions in the `rgsl` package do this
concatenation when the solver is called. We sample in
logarithmic space, but the solver gets `k` and appends
the right `u` for each experiment, both in linear space.

So, if `parMCMC` is in log10-space, then `parMap` is the
transformation to linear-space for the solver to work in:

```
parMap <- function(par) {
    return(10^(par))
}
```
