---
title: "Parallel chains with MPI"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE,
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# Motivation

Markov chain Monte Carlo scales exponentially with the problem
size. For models with ~50 parameters it already becomes hard to
intialize a Markov chain with good values:
- starting parameters
- step size

The effect is a slow moving chain, and a sample with high auto-correlation.

To reduce auto-correlation, we suggest the parallel tempering
technique, where several Markov chains run in parallel and each
experiences a different likelihood: $p(D|\theta)^\beta$ where $\theta$
is the Markov chain variable (maps to the model's kinetic parameters)
and $D$ is the available data. In this context $0 \le \beta \le 1$
plays the role of an inverse temperature. A value of $\beta = 1.0$ is
equivalent to the original problem; the lower $\beta$ gets, the more
does the target distribution resemble the prior distribution.

The different chains are then allowed to exchange their positions if
it is benefitial (according to the specific rules of parallel
tempering). This exchange of information makes the chain with $\beta =
1$ explore the sampling space much faster than the Markov chain
algorithm by itself would allow.

Even though the implementation of Parallel Tempering is possible
without MPI, we now have several experiments to simulate on many
parallel chains, for AKAP79 this is 20 experiments and e.g. 32
chains. At this point, it becomes benefitial to do the simulations on
more than one machine (mine has 16 cores). The _high performance
computing_ cluster at KTH (Parallel Dator Center, Dardel cluster) has
128 cores per node. This means that we can simulate 6 chains per node
and still process all experiments at the same time (using
`parLapply` or `mclapply`), but also start 24 chains distributed accross 3 nodes.

The distribution requires message passing between nodes, for this
prupose we use [pbdMPI](https://cran.r-project.org/package=pbdMPI)

MPI code is difficult to understand, write, test, and debug. So, the
below instructions may not work for everyone.

# AKAP79

MPI has several mechanisms to launch worker processes, one ow which is
the `spawn` method (`MPI_Comm_spawn`). We don't use this here.

Instead we launch an Rscript via `mpirun` (and the `send` and
`recv` functions on `MPI_COMM_WORLD`, which corresponds to `comm=0` in
pbdMPI).

The only difference between the workers is the `rank` and possibly the
random number seed. The rank will determine the value of $\beta$ and
thus the kind of sample the worker obtains.


Locally (on your machine) you can test the MPI script using the
`mpirun` command, if you have some mpi library installed:

```{sh, eval=FALSE, label="mpirun"}
# find number of cores on GNU/linux
N = $(( `grep -c processor /proc/cpuinfo` )) # or just type the right number
mpirun -H localhost:$N -N $N ./AKAP79.R 5000
```

For this to work, `AKAP79.R` starts with the line:

```{sh, eval=FALSE}
#!/usr/bin/env Rscript
## ^^^^^^^^^^^^^^^^^^^
## The above has to be the first line.
## Nothing may be written before that.
[...]

R code follows below
```

So, Rscript will be found and used to process the rest.

If you want to call `Rscript` explicitly, then:

```{sh, eval=FALSE, label="Rscript"}
mpirun -H localhost:$N -N $N Rscript ./AKAP79.R 5000
```

On the cluster the `-H` option is not necessary, because a queue
system like slurm will determine which machines to use.

The argument `5000` (the sample size) is passed to
the R program `AKAP79.R`, it's not an `mpirun` option.

`-N ...` means that we launch this many MPI workers per node (node
means the same as machine or host). We have one node (localhost), the
`-H localhost:$SLOTS` option determines the maximum number of jobs per
node.

If you have 16 cores:

```
mpirun -H localhost:$16 -N 16 Rscript ./AKAP79.R 5000
```

In this example, we use the MPI chains for parallel chains on
different temperatures. But, this is not mandatory. It is possible to
run `$N` chains on the same temperature, or split the work in some other
way.

# MPI Script Template

This is an example that will sample a model using the Metropolis
algorithm, using [`pbdMPI`](https://cran.r-project.org/package=pbdMPI)
and the parallel tempering technique.

```{r, eval=FALSE, label="mpi-pt-sampling.R"}
#!/usr/bin/env Rscript

library(rgsl)
library(SBtabVFGEN)
library(uqsa)
library(parallel)
library(pbdMPI)

start_time <- Sys.time()
comm  <- 0
pbdMPI::init()
r <- pbdMPI::comm.rank(comm=comm)
cs <- pbdMPI::comm.size(comm=comm)

attr(comm,"rank") <- r
attr(comm,"size") <- cs
N <- 10000                 # default sample size
h <- 1e-2                  # step size

beta <- (1.0 - (r/cs))^2   # PT: inverse temperature

## ----label="SBtab content"----------------------------------------------------

modelFiles <- dir("..",pattern="[.]tsv$",full.names=TRUE)
sb <- SBtabVFGEN::sbtab_from_tsv(modelFiles)

modelName <- checkModel(comment(sb),sprintf("./%s_gvc.c",comment(sb)))

## ----ConservationLaws----------------------------------------------------------
if (file.exists("../ConservationLaws.RDS")){
	ConLaw <- readRDS("../ConservationLaws.RData")
} else {
	ConLaw <- NULL
}
experiments <- sbtab.data(sb,ConLaw)

## ----default------------------------------------------------------------------
n <- length(experiments[[1]]$input)
if (n>0){
	parMCMC <- log10(head(model$par(),-n))
} else {
	parMCMC <- log10(model$par())
}

## ----range--------------------------------------------------------------------
stdv <- head(sb$Parameter[["!Std"]],length(parMCMC))
if (is.null(stdv) || any(is.na(stdv))){ # some alternative default value
	stdv <- parMCMC*0.05 + max(parMCMC)*0.05 + 1e-2
}
dprior <- dNormalPrior(mean=parMCMC,sd=stdv)
rprior <- rNormalPrior(mean=parMCMC,sd=stdv)

## ----simulate-----------------------------------------------------------------
sim <- simulator.c(experiments,modelName,log10ParMap)

y <- sim(parMCMC) ## little test
stopifnot(is.list(y) && length(y)==length(experiments))
stopifnot(all(c("state","func") %in% names(y[[1]])))

## ----likelihood---------------------------------------------------------------
logLH <- function(y,h,stdv,name){
	n <- sum(!is.na(stdv))
	llf_const <- sum(log(stdv),na.rm=TRUE) + 0.5*log(2*pi)*n
	llf_sq <- 0.5*sum(((y - h)/stdv)^2,na.rm=TRUE)
	return(-llf_const-llf_sq)
}

llf <- logLikelihoodFunc(experiments,simpleUserLLF=logLH)

## ----update-------------------------------------------------------------------
metropolis <- mcmcUpdate(simulate=sim,
	experiments=experiments,
	model=model,
	logLikelihood=llf,
	dprior=dprior)

## ----init---------------------------------------------------------------------
ptMetropolis <- mcmc_mpi(metropolis,comm=comm,swapDelay=0,swapFunc=pbdMPI_bcast_reduce_temperatures)
## ----sample-------------------------------------------------------------------
x <- mcmcInit(
	beta,
	parMCMC,
	simulate=sim,
	logLikelihood=llf,
	dprior)
s <- ptMetropolis(x,N,h) # the main amount of work is done here
saveRDS(s,file=sprintf("sample-rank-%i.RDS",r))

# ---- when all are done, we load the sampled points from the files but only for the right temperature:
pbdMPI::barrier()
f <- dir(pattern=sprintf('sample-rank-.*RDS$'))
stopifnot(length(f) == cs)                                          # MPI comm size
X <- uqsa::gatherSample(f,beta,size=N/2)

Copula <- fitCopula(X)
dprior <- dCopulaPrior(Copula)
rprior <- rCopulaPrior(Copula)
parMCMC <- as.numeric(tail(X,1))
attr(Copula,"beta") <- beta
attr(X,"beta") <- beta

saveRDS(Copula,file=sprintf("Copula-for-beta-%i-l10b-%i.RDS",r,round(100*log10(beta))))
saveRDS(X,file="parameter-sample-for-beta-%i-l10b-%i.RDS",r,round(100*log10(beta)))

time_ <- difftime(Sys.time(),start_time,units="min")
print(time_)
finalize()
```
