---
title: "ODE Model"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

In some cases, the model already is an Ordinary Differential Equation,
and we don't want to first think of it as a reaction network.

In such cases, these approaches are equally good/plausible:

- write the model in vfgen's `vf` format
    + avoid line-breaks in self-closed tags: `<SomeElement attr="value"/>` (always one line)
    + [`ode.sh`](icpm-kth/RPN-erivative) does not use an xml parser since there is no mathml
	+ The attribute values are extracted using regular expressions and POSIX tools
	+ standard _regular expressions_ usually don't deal well with
      content that is spread across lines (Rob Pike
      [wrote](https://doc.cat-v.org/bell_labs/structural_regexps/se.pdf)
      about this a long time ago)
- use the `ode` format of `xpp` and `xppaut`
    + there is limited support for this format as `xpp` has a huge list of capabilities
	+ differentiation has to be `var' = formula`, with a `'` (prime)
	+ no delay equations
	+ no DAE
	+ no `bdry`, `global`, `wiener`, `table`, and other special functions
- convert the model into the text form used by `RPN-derivative/sh/ode.sh`, by imitating the examples
    + `ode.sh` has an `--inspect` option that leaves the created temporary files intact for inspection
	+ this can be used for debugging as well
	+ whichever format `ode.sh` receives, it will convert everything into `txt` files that end up in the _temp_ folder
	+ on GNU/Linux the temp folder is `/dev/shm/ode_gen/` (it doesn't survive reboots)
- write the model's C source code directly using a computer algebra system
    + maxima and yacas are good
	+ this is not automated at all
    + feasible if you have experience with these tools

On this page, we will use the `.ode` method.

# Simple Example

Let's take this [model of a
pendulum](https://warrenweckesser.github.io/vfgen/menu_xpp.html) as an
example, in xpp's `.ode` format because it is very compact and
beautiful:

```{c, eval=FALSE}
# pendulum.ode
par g=9.81
par b=0.0
par L=1.0
par m=1.0

theta'=v
init theta=3.13159

v'=-v*m^(-1)*b*L^(-2)-g*sin(theta)*L^(-1)
init v=0.0

# Output
aux energy=-m*g*cos(theta)*L+0.5*v^2*m*L^2
#
done
```

```{sh, include=FALSE}
(
cat<<EOF
# The Model starts here
par g=9.81
par b=0.0
par L=1.0
par m=1.0

theta'=v
init theta=3.13159

v'=-v*m^(-1)*b*L^(-2)-g*sin(theta)*L^(-1)
init v=0.0

#
aux energy=-m*g*cos(theta)*L+0.5*v^2*m*L^2
#
done
EOF
) > pendulum.ode
```

We will use this model, and convert it to the text format that `ode.sh` can read using standard command line tools:

```{sh, eval=FALSE}
f="pendulum.ode"
echo "$f"
egrep '^par' "$f" | tr '=' '\t' | sed 's/^par //' > Parameters.txt
egrep '^init' "$f" | tr '=' '\t' | sed 's/^init //' > StateVariables.txt
egrep "'" "$f" | sed -E "s/([^=]*)'=/\1\t/;" > ODE.txt
egrep '^aux' "$f" | tr '=' '\t' | sed 's/^aux[ ]*//' > OutputFunctions.txt
tar czf pendulum.tar.gz Parameters.txt StateVariables.txt ODE.txt OutputFunctions.txt
[ -f pendulum.tar.gz ] && ode -C --maxima pendulum.tar.gz > pendulum_gvf.c
```

At this stage, we could add transformations (manually), if we wanted.

---

But, actually, there is a limited parser for `ode` files that does something like the above.
It runs if the model file ends in ode:

```{sh}
ode -C --maxima pendulum.ode > pendulum_gvf.c
```

The resulting code can be found at the bottom of this page.

In R, we can use this code directly:

```{r, eval=TRUE}
library(rgsl)
library(uqsa)

f <- "pendulum_gvf.c"
modelName <- checkModel("pendulum",f)

ex <- list(
	test=list(
		initialTime=-1.0,
		outputTimes=seq(0,10,0.1))
	)

SIM <- simcf(ex,modelName)
par <- c(g=9.81, b=0.0, L=1.0, m=1.0)
print(par)

y <- SIM(par)
plot(ex[[1]]$outputTimes,y[[1]]$state[1,,1],xlab='t',ylab='theta',bty='n',type='l')
```

## Generated C Code

```{c, eval=FALSE}
#include <stdlib.h>
#include <math.h>
#include <gsl/gsl_errno.h>
#include <gsl/gsl_odeiv2.h>
#include <gsl/gsl_math.h>

enum stateVariable { _theta,_v, numStateVar }; /* state variable indexes  */
enum param { _g,_b,_L,_m, numParam }; /* parameter indexes  */
enum func { _energy, numFunc }; /* parameter indexes  */

/* The error code indicates how to pre-allocate memory
 * for output values such as `f_`. The _vf function returns
 * the number of state variables, if any of the args are `NULL`.
 * evaluation errors can be indicated by negative return values.
 * GSL_SUCCESS (0) is returned when no error occurred.
 */

/* ode vector field: y'=f(t,y;p) */
int pendulum_vf(double t, const double y_[], double f_[], void *par)
{
	double *p_=par;
	if (!y_ || !f_) return 2;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
	f_[_theta] = v; /* theta */
	f_[_v] = -v*pow(m,-1)*b*pow(L,-2)-g*sin(theta)*pow(L,-1); /* v */
	return GSL_SUCCESS;
}
/* ode Jacobian df(t,y;p)/dy */
int pendulum_jac(double t, const double y_[], double *jac_, double *dfdt_, void *par)
{
	double *p_=par;
	if (!y_ || !jac_) return 2*2;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
/* column 1 (df/dy_0) */
	jac_[0] = 0; /* [0, 0] */
	jac_[2] = -(g*cos(theta))/L; /* [1, 0] */
/* column 2 (df/dy_1) */
	jac_[1] = 1; /* [0, 1] */
	jac_[3] = -b/(gsl_pow_2(L)*m); /* [1, 1] */
	return GSL_SUCCESS;
}
/* ode parameter Jacobian df(t,y;p)/dp */
int pendulum_jacp(double t, const double y_[], double *jacp_, double *dfdt_, void *par)
{
	double *p_=par;
	if (!y_ || !jacp_) return 2*4;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
/* column 1 (df/dp_0) */
	jacp_[0] = 0; /* [0, 0] */
	jacp_[4] = -sin(theta)/L; /* [1, 0] */
/* column 2 (df/dp_1) */
	jacp_[1] = 0; /* [0, 1] */
	jacp_[5] = -v/(gsl_pow_2(L)*m); /* [1, 1] */
/* column 3 (df/dp_2) */
	jacp_[2] = 0; /* [0, 2] */
	jacp_[6] = (2*b*v)/(gsl_pow_3(L)*m)+(g*sin(theta))/gsl_pow_2(L); /* [1, 2] */
/* column 4 (df/dp_3) */
	jacp_[3] = 0; /* [0, 3] */
	jacp_[7] = (b*v)/(gsl_pow_2(L)*gsl_pow_2(m)); /* [1, 3] */
	return GSL_SUCCESS;
}
/* ode Functions F(t,y;p) */
int pendulum_func(double t, const double y_[], double *func_, void *par)
{
	double *p_=par;
	if (!y_ || !func_) return 1;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
	func_[_energy] = -m*g*cos(theta)*L+0.5*gsl_pow_2(v)*m*gsl_pow_2(L); /* energy */
	return GSL_SUCCESS;
}
/* Function Jacobian dF(t,y;p)/dy */
int pendulum_funcJac(double t, const double y_[], double *funcJac_, void *par)
{
	double *p_=par;
	if (!y_ || !funcJac_) return 2;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
/* column 1 (dF/dy_0) */
	funcJac_[0] = L*g*m*sin(theta); /* [0, 0] */
/* column 2 (dF/dy_1) */
	funcJac_[1] = 1.0*gsl_pow_2(L)*m*v; /* [0, 1] */
	return GSL_SUCCESS;
}
/* Function parameter Jacobian dF(t,y;p)/dp */
int pendulum_funcJacp(double t, const double y_[], double *funcJacp_, void *par)
{
	double *p_=par;
	if (!y_ || !funcJacp_) return 4;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	double theta=y_[0];
	double v=y_[1];
/* column 1 (dF/dp_0) */
	funcJacp_[0] = -L*m*cos(theta); /* [0, 0] */
/* column 2 (dF/dp_1) */
	funcJacp_[1] = 0; /* [0, 1] */
/* column 3 (dF/dp_2) */
	funcJacp_[2] = 1.0*L*m*gsl_pow_2(v)-g*m*cos(theta); /* [0, 2] */
/* column 4 (dF/dp_3) */
	funcJacp_[3] = 0.5*gsl_pow_2(L)*gsl_pow_2(v)-L*g*cos(theta); /* [0, 3] */
	return GSL_SUCCESS;
}
/* ode default parameters */
int pendulum_default(double t, void *par)
{
	double *p_=par;
	if (!p_) return 4;
	p_[_g] = 9.81;
	p_[_b] = 0.0;
	p_[_L] = 1.0;
	p_[_m] = 1.0;
	return GSL_SUCCESS;
}
/* ode initial values */
int pendulum_init(double t, double *y_, void *par)
{
	double *p_=par;
	if (!y_) return 2;
	double g=p_[0];
	double b=p_[1];
	double L=p_[2];
	double m=p_[3];
	/* the initial value of y may depend on the parameters. */
	y_[_theta] = 3.13159;
	y_[_v] = 0.0;
	return GSL_SUCCESS;
}
```
