---
title: "Uncertainty Quantification on AKAR4 (deterministic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=TRUE,
  cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(errors)
```

# Load Model and Data

The first step is to locate and load the SBtab model files. In this
example we use the SBtab files from an already existing example
("AKAR4").

In the article ["Build and simulate your own Model](user_model.html)
you can read about how the set up your own SBtab model. You can list
the files in a directory with the `dir()` command.


```{r load}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m)
C <- generateCode(o)
modelName <- write_and_compile(C)
```

The chain of functions that we use to build the model tracks the name
of the model via comments:

```{r comments}
print(comment(m))
print(comment(o))
print(comment(C))
print(comment(modelName))

```

This comment is ultimately used to determine the automatic file-name
of the shared library.

```{r data}
ex <- experiments(m,o)
```

## Take a parameter Sample for this Model

We use just one core as this is a tiny model. No MPI, no parallel tempering. First we construct the prior:

```{r prior}
mu <- values(m$Parameter)
sigma <- log(m$Parameter$stdv)

dprior <- dNormalPrior(
	mean=log(mu^2/sqrt(mu^2+sigma^2)),
	sd=log(1+sigma^2/mu^2)
)
rprior <- rNormalPrior(
	mean=log(mu)/sqrt(mu^2+sigma^2),
	sd=log(1+sigma^2/mu^2)
)
```

Next, the MCMC method via `metropolis_update`, which determines the
sampling algorithm we are going to use. Finally, we construct the
sampler (`MH` below), which actually performs the MCMC run, this
function takes the inital values, number of steps and the stepsize as
arguments.

```{r Metropolis-Hastings}
options(mc.cores=length(ex))
s <- simulator.c(ex,modelName,parMap=logParMap, omit=2)

MH <- mcmc(
    metropolis_update(
        s,
        dprior=dprior
    )
)
```

Finally, to obtain a sample of parameters:

```{r sample, fig.width=12, fig.height=12, dpi=100}
x <- mcmc_init(
    beta=1.0,
    parMCMC=log(values(m$Parameter)),
    simulate=s,
    dprior=dprior)

h <- tune_step_size(MH,x)
N <- 5e4

# sampling:
X <- MH(x,N,eps=h)

print(attr(X,"acceptanceRate"))

if (require(hexbin)){
    hexplom(X)
} else {
    pairs(X)
}
```

## Better Sample

We can try to obtain a higher quality sample, through simple means:

1. Combine several samples (replicas) into one big sample
2. Use several random starting locations
3. Remove burn-in phase from each chain
4. Adjust the step size at least once to get close to 25% acceptance rate



