---
title: "Uncertainty Quantification of the AKAR4 (stochastic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = FALSE,
  eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(uqsa)
library(errors)
library(parallel)
```

This article provides code to perform uncertainty quantification (UQ)
of a stochastic reaction network model. As an example we use the AKAR4
stochastic model. The UQ method that we use is ABCMCMC.

For information about the stochastic model, please read the article
[Simulate AKAR4 stochastically](simAKAR4stochastic.html).


# Load the Model

This model is included with the package, as one of the example models.

```{r load-files}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m,cla=FALSE)
ex <- experiments(m,o)
```
## Make a Stochastic Version of AKAR4

```{r write}
sm <- makeGillespieModel(m)
C <- generateGillespieCode(sm)
cat(C,sep="\n",file="AKAR4_st.c")
modelName <- checkModel("AKAR4","AKAR4_st.c")
```

To simulate the stochastic reaction network model we use the
*Gillespie algorithm*.

# Define MCMC settings

```{r configure}
# scale to determine prior values
defRange <- 1000
par_val <- values(m$Parameter)

# Define Lower and Upper Limits for logUniform prior distribution for the parameters
ll <- c(par_val/defRange)
ul <- c(par_val*defRange)
ll = log10(ll) # log10-scale
ul = log10(ul) # log10-scale

# Define the prior distribution. In this case Uniform(ll,ul)
dprior <- dUniformPrior(ll, ul) # dprior evaluates the prior density function
rprior <- rUniformPrior(ll, ul) # rprior generates random samples from the prior
```

# Generate an objective function

An *objective function* in this case is a function that can read the
experimental conditions (initial state, inputs etc.) and the
experimental data in a list of experiments (variable `experiments` in
the code). Give a parameter in input, for each experimental condition
specified in the list of experiments, the *objective function*:

1. simulates the stochastic model given such parameter, and
2. computes the distance between the simulated trajectories and the
   observed trajectories (experimental data).

To generate the objective function for stochastic models you can use
the `uqsa` function `makeObjectiveSSA`.

```{r makeObjective}
s <- simstoch(
	ex,
	model.so=comment(modelName),
	parMap=logParMap
)
objFunc <- makeObjective(ex,s)
d <- objFunc(log(values(m$Parameter)))
print(d)
```

# Run PreCalibration sampling

Before running the ABCMCMC code, we run a precalibration to determine
good initial values and algorithmic hyperparameters for the ABCMCMC
chains.

```{r precalibration}
nChains <- detectCores()
pC <- preCalibration(
	objFunc,
	1000,
	rprior,
	rep=1,
	p=0.01,
	sfactor=0.05,
	delta=0.5,
	num=nChains
)
```

Next we check the results of pre-calibration:

```{r use-pre-calibration}
Sigma <- pC$Sigma
startPar <- pC$startPar
for(j in seq(nChains)){
	cat("Chain", j, "\n")
	cat("\tMin distance of starting parameter for chain",j," = ", min(objFunc(startPar[,j])),"\n")
	cat("\tMean distance of starting parameter for chain",j," = ", mean(objFunc(startPar[,j])),"\n")
	cat("\tMax distance of starting parameter for chain",j," = ", max(objFunc(startPar[,j])),"\n")
}
```

# Run ABC-MCMC Sampling

The sampling is parallelized on several cores using the `parLapply`
function. Running this code will take approximately 10 minutes on an
8-core laptop.

```{r abc}
t_ <- Sys.time()
options(mc.cores = nChains)
res <- mclapply(
	seq(nChains),
	function(j) {
		tryCatch(
			ABCMCMC(
				objFunc,
				startPar[,j],
				100,
				Sigma,
				delta=0.5,
				dprior=dprior
			),
			error=function(cond) {
				message("ABCMCMC crashed");
				return(NULL)
			}
		)
	}
)

print(difftime(Sys.time(),t_))
```

Collect the fork-cluster results:

```{r collect}
ABCMCMCoutput <- do.call(Map, c(rbind,res))
colnames(ABCMCMCoutput$draws) <- names(par_val)
```

# Visualize the sampled parameters

The parameter-values in `ABCMCMCoutput$draws` are approximate samples
from the posterior distribution, given the experiments.

The sampled three dimensional parameters for the AKAR4 model can be
visualized via pairs of scatter plots, or two dimensional binning:

```{r pairs-2, out.width="100%", fig.width=12, fig.height=12, dpi=100}
if (requireNamespace(hexbin)){
	hexbin::hexplom(ABCMCMCoutput$draws, xbins = 16)
} else {
	pairs(ABCMCMCoutput$draws, pch=20)
}
```

The marginal distributions of each parameter in the parameter vector
can be visualized via histograms as follows.

```{r}
for(i in seq_along(par_val)){
	hist(
		ABCMCMCoutput$draws[,i],
		breaks = 20,
		main=names(par_val)[i],
		xlab = "Value in log scale"
	)
}
```
