---
title: "Sample AKAR4 with Parallel Tempering"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  eval = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

To take a sufficiently large sample of AKAP79, we shall use MPI and
parallel tempering. There are several ways to launch MPI processes (e.g. spawn processes).

We will use the simplest form: launch all MPI workers right from the
start, using `mpirun -N 8` (or similar) and have all processes
communicate on `MPI_COMM_WORLD`, this requires very minimal MPI code in R.

# Create C Source Code

First we load the model in its SBtab form and create all derived
files. We do this step only once and have the MPI part of the process
load the coimpleted results of this:

```{r model}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m)
C <- generate_code(o)
c_file <- "/dev/shm/AKAR4.c"
cat(C,sep="\n",file=c_file)
c_path(o) <- c_file
so_path(o) <- shlib(o)
ex <- experiments(m,o)

saveRDS( m,file="/dev/shm/AKAR4-m.RDS")
saveRDS( o,file="/dev/shm/AKAR4-o.RDS")
saveRDS(ex,file="/dev/shm/AKAR4-e.RDS")
```

# Sample via MPI

Here we run a prepared R script from the command line (any POSIX shell
will do: bash/zsh/fish/dash). This way, the entire R program is
wrapped in an mpirun call.

```{sh MPI, eval=TRUE}
N=4 # by default
[ -e '/proc/cpuinfo' ] && N=$((`grep -c processor /proc/cpuinfo`)) && nm="`grep -m1 'model name' /proc/cpuinfo`"
echo "We will use $N cores with $nm"
sample_size=30000

start_time=$((`date +%s`))
date
mpirun -H localhost:$N -N $N ./pt-mh-akar4.R $sample_size > ./log.txt 2>&1
date
end_time=$((`date +%s`))
echo "Time spent sampling: $((end_time - start_time)) seconds (for $sample_size points)."
```

On a high performance computing (hpc) cluster, the above would be in a
slurm script or similar workload manager.

# Inspect the Results

Here we check the integrated auto-correlation length (Markov chain
time)

```{r uwerr, out.width="100%", fig.width=5, fig.height=7}
Sample <- readRDS(file="/dev/shm/AKAR4-parameter-sample-for-rank-0.RDS")
n <- NROW(Sample)
l <- attr(Sample,"logLikelihood")

if (requireNamespace("hadron")){
    par(mfrow=c(3,1))
    res <- hadron::uwerr(data=l,pl=TRUE)
    tau <- ceiling(res$tauint+res$dtauint)
} else {
    A <- acf(l)
    tau <- sum(A$acf[A$acf>0.2])
    plot(l,type="l",xlab="mcmc index",ylab="log-likelihood")
}
cat(sprintf("Effective sampple size: %i\n",round(n/(2*tau))))
```

We reduce the sample for plotting purposes, by showing only every \(\tau_{\text{int.}}\)-th point and create a pairs plot for the first 6 parameters:

```{r pairs, fig.width=12, fig.height=12, fig.fullwidth=TRUE, dpi=100}
X <- Sample[seq(1,n,by=tau),]
colnames(X) <- rownames(m$Parameter)
if (require(hexbin)){
    hexbin::hexplom(X,xbins=16)
} else {
    pairs(X)
}
```

# The R script for Parallel Tempering

The R-Script executed via `mpirun` (MPI context) in the section above
is this:

```{r pt-mh-akar4, eval=FALSE}
#!/usr/bin/env Rscript
library(parallel)
library(uqsa)
library(pbdMPI)

comm  <- 0
pbdMPI::init()
r <- pbdMPI::comm.rank(comm=comm)
cs <- pbdMPI::comm.size(comm=comm)
attr(comm,"rank") <- r
attr(comm,"size") <- cs

ca <- commandArgs(trailingOnly=TRUE)
if (length(ca)>0){
	N <- as.integer(ca[1])
} else {
	N <- 1000              # default sample size
}

beta <- (1.0 - (r/cs))^2   # PT: inverse temperature

m <- readRDS(file="/dev/shm/AKAR4-m.RDS")
o <- readRDS(file="/dev/shm/AKAR4-o.RDS")
ex <- readRDS(file="/dev/shm/AKAR4-e.RDS")

parMCMC <- log10(values(m$Parameter))
stdv <- rep(2,length(parMCMC))
dprior <- dNormalPrior(mean=parMCMC,sd=stdv)
rprior <- rNormalPrior(mean=parMCMC,sd=stdv)

sim <- simfi(ex,o,log10ParMap,omit=2) # or simulator.c

metropolis <- metropolis_update(
	simulate=sim,
	dprior=dprior
)

rwm <- mcmc(metropolis) # non-parallel

ptMetropolis <- mcmc_mpi( # parallel
	metropolis,
	comm=comm
)

x <- mcmc_init(
	beta,
	parMCMC,
	simulate=sim,
	dprior=dprior
)

h <- tune_step_size(rwm,x,1e-3)
gc()
Z <- ptMetropolis(x,N,h) # the main amount of work is done here
saveRDS(Z,file=sprintf("/dev/shm/sample-rank-%i.RDS",r))
pbdMPI::barrier()
rm(Z)

f <- dir("/dev/shm",pattern=sprintf('sample-rank-.*RDS$'),full.names=TRUE)
X <- gatherSample(f,beta)
saveRDS(X,file=sprintf("/dev/shm/AKAR4-parameter-sample-for-rank-%i.RDS",r))
gc()

time_ <- difftime(Sys.time(),start_time,units="min")
print(time_)

cat(
	sprintf(
		"rank %02i of %02i finished with an acceptance rate of %f %% and swap rate of %f.\n",
		round(r),
		round(cs),
		attr(Z,"acceptanceRate"),
		attr(Z,"swapRate")
	)
)

finalize()
```
