---
title: "sampleAKAP79"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache=FALSE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

To take a sufficiently large sample of AKAP79, we shall use MPI and
parallel tempering. There are several ways to launch MPI processes (e.g. spawn processes).

We will use the simplest form: launch all MPI workers right from the
start, using `mpirun -N 8` (or similar) and have all processes
communicate on `MPI_COMM_WORLD`, this requires very minimal MPI code in R.

# Create C Source Code

First we load the model in its SBtab form and create all derived
files. We do this step only once and have the MPI part of the process
load the coimpleted results of this:

```{r SBtab}
f <- uqsa::uqsa_example("AKAP79")
sb <- SBtabVFGEN::sbtab_from_tsv(f)
```

Create the ODE interpretation (output omitted)
```{r sbtab-to-vfgen, eval=FALSE}
m <- SBtabVFGEN::sbtab_to_vfgen(sb)
```
```{r sbtab-to-vfgen-invisible, include=FALSE}
m <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

```{r experiments}
ex <- SBtabVFGEN::sbtab.data(sb,m$conservationLaws)
saveRDS(sb,file="AKAP79-sb.RDS")
saveRDS(ex,file="AKAP79-ex.RDS")
```

The next step is to generate the code, write it into a file ending in
`.c` and compiling it into a shared library:

```{r GSL}
C <- uqsa::generateCode(m)
cFile <- sprintf("./%s_gvf.c",comment(sb)) # GSL Vector Field file name
cat(C,sep="\n",file=cFile)                 # write sources to that file
modelName <- uqsa::checkModel(comment(sb),cFile) # this compiles the sources
```

# Sample via MPI

Here we run a prepared R script from the command line (any POSIX shell will do: bash/zsh/fish/dash)

```{sh MPI, eval=TRUE}
N=4 # by default
[ -e '/proc/cpuinfo' ] && N=$((`grep -c processor /proc/cpuinfo`)) && nm="`grep -m1 'model name' /proc/cpuinfo`"
echo "We will use $N cores with $nm"

start_time=$((`date +%s`))
date
mpirun -H localhost:$N ./pt-smmala-akap79.R 100000
date
end_time=$((`date +%s`))
echo "Time spent sampling: $((end_time - start_time)) seconds ($(( (end_time - start_time)/60 )) minutes)."
```

On a high performance computing (hpc) cluster, the above would be in a
slurm script or similar workload manager.

# Inspect the Results

Here we check the integrated auto-correlation length (Markov chain
time)

```{r uwerr}
Sample <- readRDS(file="./AKAP79-temperature-ordered-smmala-sample-for-rank-0.RDS")
n <- NROW(Sample)
l <- attr(Sample,"logLikelihood")

plot(l,type="l",xlab="mcmc index",ylab="log-likelihood")

if (require(hadron)){
    res <- hadron::uwerr(data=l,pl=TRUE)
}
```

We reduce the sample for plotting purposes, by showing only every \(\tau_{\text{int.}}\)-th point and create a pairs plot for the first 6 parameters:

```{r pairs, fig.width=12, fig.height=12, fig.fullwidth=TRUE, dpi=100}
X <- Sample[seq(1,n,by=ceiling(res$tauint)),]
colnames(X) <- rownames(sb$Parameter)
if (require(hexbin)){
    hexbin::hexplom(X[,seq(6)],xbins=12)
} else {
    pairs(X[,seq(6)])
}
```
