---
title: "Uncertainty Quantification of AKAP79 (deterministic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache=TRUE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

To take a sufficiently large sample of AKAP79, we shall use MPI and
parallel tempering. There are several ways to launch MPI processes
(e.g. spawn processes).  We will use the simplest form: launch all MPI
workers right from the start, using `mpirun -N 8` (or similar) and
have all processes communicate on `MPI_COMM_WORLD`, this requires very
minimal MPI code in R.  This approach doesn't allow us to adapt the
number of temperatures, which is probably not optimal but makes it
easier to write and understand the code and increases
performance significantly compared to a sequential Markov chain.

The R code we are going to execute is not a code
block on this page that we run in an interactive R session. Instead,
it has to be a file that we pass on to `mpirun`.

In this article we show the results of a parallel tempering SMMALA
(simplified manifold Metropolis adjusted Langevin algorithm) run. The
article titled [Parallel chains with MPI](articles/mpi.html) explains
how to prepare the R-script for mpirun execution -- it includes the
listing of the script we used in this article.

# Create C Source Code

First we load the model in its SBtab form and create all derived model
files. We do this step only once and have the MPI part of the process
load the coimpleted results of this:

```{r tsv}
f <- uqsa::uqsa_example("AKAP79")
m <- model_from_tsv(f)
```

Create the ODE interpretation of `m`:

```{r sbtab-to-vfgen, eval=FALSE}
o <- as_ode(m)
```

Load the simulation instructions for `o`:

```{r experiments}
ex <- experiments(m,o)
saveRDS(ex,file="AKAP79-ex.RDS")
```

There is a high-level function that does most of what we need to do automatically:

```{r}
p <- log10(values(m$Parameter))
smmala <- high_level_smmala(m,o,ex,p)
```

# Plain Sample

```{r, quick-sample}
h <- tune_step_size(smmala)
X <- smmala(smmala %@% "init",1000,h)
```

We shall only plot the pairs of the first 6 dimensions (this is a big model):

```{r, hexbin-plot, out.width="100%"}
if (require(hexbin)){
    hexplom(X[,seq(6)])
} else {
    pairs(X[,seq(6)])
}
```

# Sample via MPI

Here we run a prepared R script from the command line (any POSIX shell will do: bash/zsh/fish/dash)

```{sh MPI, eval=TRUE}
N=4 # by default
[ -e '/proc/cpuinfo' ] && N=$((`grep -c processor /proc/cpuinfo`)) && nm="`grep -m1 'model name' /proc/cpuinfo`"
echo "We will use $N cores with $nm"

start_time=$((`date +%s`))
date
mpirun -H localhost:$N ./pt-smmala-akap79.R 16000 > akap79.log
date
end_time=$((`date +%s`))
echo "Time spent sampling: $((end_time - start_time)) seconds ($(( (end_time - start_time)/60 )) minutes)."
```

On a high performance computing (hpc) cluster, the above would be in a
slurm script or similar workload manager.

# Inspect the Results

Here we check the integrated auto-correlation length (Markov chain
time)

```{r uwerr}
Sample <- readRDS(file="AKAP79-temperature-ordered-pt-smmala-sample-2-for-rank-0.RDS")
n <- NROW(Sample)
l <- attr(Sample,"logLikelihood")

plot(l,type="l",xlab="mcmc index",ylab="log-likelihood")

if (require(hadron)){
    res <- hadron::uwerr(data=l,pl=TRUE)
    tau <- ceiling(res$tauint + res$dtauint) # upper bound
} else {
    A <- acf(l,lag.max=10000)
    tau <- ceiling(sum(A$acf))
}
```

We reduce the sample for plotting purposes, by showing only every
\(\tau_{\text{int.}}\)-th point and create a
_pairs_ plot for the first
6 parameters:

```{r pairs, fig.width=12, fig.height=12, fig.fullwidth=TRUE, dpi=100}
X <- Sample[seq(1,n,by=tau),]
colnames(X) <- rownames(sb$Parameter)
if (require(hexbin)){
    hexplom(X[,seq(6)],xbins=12)
} else {
    pairs(X[,seq(6)])
}
```
