---
title: "Uncertainty Quantification of AKAP79 (deterministic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval=TRUE,
  cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(parallel)
options(mc.cores=detectCores())
```

AKAP79 is a comparativelyh big model (for a demonstartive example). To
even attempt to take a somewhat sufficiently large sample of AKAP79,
we shall use MPI and parallel tempering. There are several ways to
launch MPI processes (e.g. spawn processes).  We will use the simplest
form: launch all MPI workers right from the start, using `mpirun -N 8`
(or similar) and have all processes communicate on `MPI_COMM_WORLD`,
this requires very minimal MPI code in R.

This approach doesn't allow us to adapt the number of temperatures,
which is probably not optimal but makes it easier to write and
understand the code and increases performance significantly compared
to a sequential Markov chain.

The R code we are going to execute is not a code block on this page
that we run in an interactive R session. Instead, it has to be a file
that we pass on to `mpirun` (a system command, in the shell).

In this article we show the results of a parallel tempering SMMALA
(simplified manifold Metropolis adjusted Langevin algorithm) run. The
article titled [Parallel chains with MPI](articles/mpi.html) explains
how to prepare the R-script for mpirun execution -- it includes the
listing of the script we used in this article.

We will store all files related to this in a very predictable
location: `/dev/shm`; not R's `tempdir()`. This is because the R
blocks in this article are not executed in the same R session as the
MPI workers will run in. Each MPI worker is its own process; thus they
all have different temporary directories. The directory `/dev/shm` is
a virtual directory that lives in the machine's RAM. It is standard on
GNU/Linux systems, but not on MACOS, so this example will not work
verbatim there (it can be adjusted to work). The sampling will take a
while, even for a very small sample, so it is not recommended to try
out this article for no reason.

# Create C Source Code

First we load the model in its tabular form and create all derived model
files:

```{r tsv}
f <- uqsa::uqsa_example("AKAP79")
m <- model_from_tsv(f)
```

Create the ODE interpretation of `m`:

```{r ode-model}
o <- as_ode(m)
cat(generate_code(o),sep="\n",file="/dev/shm/AKAP79.c")
c_path(o) <- "/dev/shm/AKAP79.c"
so_path(o) <- shlib(o)
saveRDS(o,file="/dev/shm/AKAP79-ode.RDS")
```

Load the simulation instructions for `o`:

```{r experiments}
ex <- experiments(m,o)
saveRDS(ex,file="/dev/shm/AKAP79-ex.RDS")
```

There is a high-level function that does most of what we need to do automatically:

```{r, label="default-parameter-values"}
p <- values(m$Parameter)
smmala <- high_level_smmala(m,o,ex,p)
```

# Plain Sample

```{r quick-sample, out.width="100%"}
h <- tune_step_size(smmala)
X <- smmala(smmala %@% "init",1000,h)
plot(
  X %@% "logLikelihood",
  type="s",
  xlab="iteration",
  ylab="logLikelihood"
)
```

We shall only plot the pairs of the first 6 dimensions (this is a big model):

```{r, label="hexbin-plot", out.width="100%", fig.width=12, fig.height=12, dpi=200}
if (require(hexbin)){
    hexplom(X[,seq(6)])
} else {
    pairs(X[,seq(6)])
}
```

# Sample via MPI

Here we run a prepared R script from the command line (any POSIX shell will do: bash/zsh/fish/dash)

```{bash MPI, eval=TRUE}
N=4 #  default number of MPI workers
[ -e '/proc/cpuinfo' ] && N=$((`grep -c processor /proc/cpuinfo`)) && nm="`grep -m1 'model name' /proc/cpuinfo`"
echo "We will use $N cores with $nm"

start_time=$((`date +%s`))
date
mpirun -H localhost:$N ./pt-smmala-akap79.R 10000 > /dev/null 2>&1
date
end_time=$((`date +%s`))
echo "Time spent sampling: $((end_time - start_time)) seconds ($(( (end_time - start_time)/60 )) minutes)."
```

On a high performance computing (hpc) cluster, the above would be in a
slurm script or similar workload manager.

# Inspect the Results

Here we check the integrated auto-correlation length (Markov chain
time)

```{r uwerr, out.width="100%"}
Sample <- readRDS(file="/dev/shm/AKAP79-temperature-ordered-pt-smmala-sample-2-for-rank-0.RDS")
n <- NROW(Sample)
l <- attr(Sample,"logLikelihood")

plot(l,type="l",xlab="mcmc index",ylab="log-likelihood")

if (require(hadron)){
    res <- hadron::uwerr(data=l,pl=TRUE)
    tau <- ceiling(res$tauint + res$dtauint) # upper bound
} else {
    A <- acf(l,lag.max=10000)
    tau <- ceiling(sum(A$acf))
}
```

We reduce the sample for plotting purposes, by showing only every
\(\tau_{\text{int.}}\)-th point and create a
_pairs_ plot for the first
6 parameters:

```{r pairs, out.width="100%", fig.width=12, fig.height=12, dpi=200}
X <- Sample[seq(1,n,by=tau),]
colnames(X) <- rownames(m$Parameter)
if (require(hexbin)){
    hexplom(X[,seq(6)],xbins=12)
} else {
    pairs(X[,seq(6)])
}
```


