---
title: "Calibrate the AKAR4 stochastic model (UQ)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(uqsa)
library(parallel)
library(SBtabVFGEN)
library(GillespieSSA2)
```

This article provides code to perform uncertainty quantification (UQ) of a stochastic reaction network model. As an example we use the AKAR4 stochastic model. The UQ method that we use is ABCMCMC.

For information about the stochastic model, please read the article [Simulate AKAR4 stochastically](simAKAR4stochastic.html).


# Load the Model

This model is included with the package. To load your own model, see
the [user model](user_model.html) article.

```{r}
modelFiles <- uqsa_example("AKAR4",full.names=TRUE)
SBtab <- SBtabVFGEN::sbtab_from_tsv(modelFiles)

# model related functions, in R, e.g. AKAP79_default() parameters
source(uqsa_example("AKAR4",pat='^AKAR4[.]R$'))
print(AKAR4_default())
```

To simulate the stochastic reaction network model we use the *Gillespie algorithm*. Specifically, we use the functions implemented in the package GillespieSSA2. To use this package we need to create a variable `reactions` using the uqsa function `importReactionsSSA` with the SBtab model as input.

```{r}
# Create reactions to simulate the stochastic model with the GillespieSSA2 package
reactions <- importReactionsSSA(SBtab)
```

Moreover, to simulate the stochastic reactions with the Gillespie algorithm we need to specify the volume of space where the reactions take place and obtain the parameter `Phi` which will be needed for simulations.

```{r}
AvoNum <- 6.022e23 #Avogadro constant
unit <- 1e-6    # unit of measure with respect to m^3 (1e-6 corresponds to micrometers)
vol <- 4e-15    # volume where the reactions take place
Phi <- AvoNum * vol * unit  # parameter used in the simulations
```

# Load Experiments (data)

```{r, experiments}
experiments <- sbtab.data(SBtab)

# for example, these is the initial state of experiment 1:
print(experiments[[1]]$initialState)

# save parameters' names
par_val <- SBtab$Parameter[["!DefaultValue"]]
par_names <- SBtab$Parameter[["!Name"]]
names(par_val) <- par_names
print(par_val)
```

# Define MCMC settings

```{r configure}
# a function that tansforms the ABC variables to acceptable model
# parameters, re-indexing could also happen here
parMap <- function (parABC=0) {
  return(10^parABC)
}


# scale to determine prior values
defRange <- 1000


# Define Lower and Upper Limits for logUniform prior distribution for the parameters
ll <- c(par_val/defRange)
ul <- c(par_val*defRange)
ll = log10(ll) # log10-scale
ul = log10(ul) # log10-scale

# Define the prior distribution. In this case Uniform(ll,ul)
dprior <- dUniformPrior(ll, ul) # dprior evaluates the prior density function 
rprior <- rUniformPrior(ll, ul) # rprior generates random samples from the prior


# Define Number of Samples for the Precalibration (npc) and each ABC-MCMC chain (ns)
nChains <- 4  # number of ABCMCMC chains to run
ns <- 100   # no of samples required from each ABC-MCMC chain
npc <- 100 # pre-calibration


# Function to compute the score (distance) between experimental and simulated data
distance <- function(funcSim, dataExpr, dataErr = 1.0){
  return(mean(((funcSim-dataExpr)/max(dataExpr))^2,na.rm=TRUE))
}

# Threshold in the ABCMCMC algorithm
delta <- 0.0005
```

# Generate an objective function

An *objective function* in this case is a function that can read the experimental conditions (initial state, inputs etc.) and the experimental data in a list of experiments (variable `experiments` in the code). Give a parameter in input, for each experimental condition specified in the list of experiments, the *objective function*:
1. simulates the stochastic model given such parameter, and
2. computes the distance between the simulated trajectories and the observed trajectories (experimental data).

To generate the objective function for stochastic models you can use the `uqsa` function `makeObjectiveSSA`.

```{r makeObjective}
objectiveFunction <- makeObjectiveSSA(experiments = experiments, model.tab = SBtab, parNames = par_names, outputFunction = model$func, distance = distance, parMap = parMap, reactions = reactions, nStochSim = 3)
```

# Run PreCalibration sampling

Before running the ABCMCMC code, we run a precalibration to determine good initial values and algorithmic hyperparameters for the ABCMCMC chains.

```{r precalibration}
p <- 0.01 # Choose Top 1% samples from the precalibration samples with shortest distance to the experimental values
sfactor <- 0.1 # Scaling factor used to determine a good covariance matrix for the ABCMCMC proposal moves

pC <- preCalibration(objectiveFunction, npc, rprior, rep = 1, p=p, sfactor=sfactor, delta=delta, num=nChains)

Sigma <- pC$Sigma
startPar <- pC$startPar
for(j in 1 : nChains){
    cat("Chain", j, "\n")
    cat("\tMin distance of starting parameter for chain",j," = ", min(objectiveFunction(startPar[,j])),"\n")
    cat("\tMean distance of starting parameter for chain",j," = ", mean(objectiveFunction(startPar[,j])),"\n")
    cat("\tMax distance of starting parameter for chain",j," = ", max(objectiveFunction(startPar[,j])),"\n")
}
```


# Run ABC-MCMC Sampling
The sampling is parallelized on several cores using the `parLapply` function.

```{r}
cl <- makeForkCluster(nChains, outfile="outputMessagesABCMCMC.txt")
out_ABCMCMC <- parLapply(
  cl,
  1:nChains,
  function(j) {
    tryCatch(
      ABCMCMC(
       objectiveFunction, startPar[,j], ns, Sigma, delta, dprior, batchSize = 1),
       error=function(cond) {message("ABCMCMC crashed"); return(NULL)}
    )
  }
)
stopCluster(cl)

ABCMCMCoutput <- do.call(Map, c(rbind,out_ABCMCMC))
colnames(ABCMCMCoutput$draws) <- par_names
```

# Visualize the sampled parameters

The parameters in the variable `ABCMCMCoutput` are approximate samples from the posterior distribution of the parameters given the considered experiments.

The sampled three dimensional parameters for the AKAR4 model can be visualized via scatter plots.

```{r}
pairs(ABCMCMCoutput$draws)
```

The marginal distributions of each parameter in the parameter vector can be visualized via histograms as follows.

```{r}
for(i in 1:3){
 hist(ABCMCMCoutput$draws[,i], main=par_names[i], xlab = "Value in log scale")
}
```

# Simulate the stochastic model given posterior samples

To check the fit of the sampled posterior parameters with the experimental data, we can simulate trajectories given sampled parameters and compare them with the experimental data.

```{r, sim}

expInd <- 1 # Index of the experiment that we want to plot, with corresponding fitted trajectories
e <- experiments[[1]]

# Plot the experimental data
par(bty='n',xaxp=c(80,120,4))
plot(e$outputTimes, e$outputValues$AKAR4pOUT,ylim=c(90,200), ylab="AKAP79",
     xlab="t",
     main=sprintf("Experiment %i",exp_ind),
     lwd=1.5)

# generate a function (s) that simulates a trajectory given a parameter in input
s <- simulator.stoch(experiments = experiments, model.tab = SBtab, parMap = parMap, outputFunction = model$func, vol = vol, unit = unit, nStochSim = 3)

n_draws <- dim(ABCMCMCoutput$draws)[1] # number of sampled posterior parameters
subsample_indices <- sample(n_draws,10) # indices of 10 random parameters out of the paramneters in
for(i in subsample_indices){
  p <- ABCMCMCoutput$draws[i,] # random parameter from the posterior

  y <- s(p) # simulate a trajectory (y) given parameter p
  # "experiments" is a list of experiments
  # "y" is a list where element i of the list is a simulation performed with the experimental conditions as in the i-th element of list "experiments"
  
  lines(e$outputTimes, y[[expInd]]$output, col="blue") # plot the simulated trajectory
}

```
