---
title: "Uncertainty Quantification on AKAR4 (deterministic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=TRUE, cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(SBtabVFGEN)
library(rgsl)
```

# Load Model and Data

The first step is to locate and load the SBtab model files. In this example we use the SBtab files from an already existing example ("AKAR4"). (In the article ["Build and simulate your own Model](user_model.html) you can read about how the set up your own SBtab model. You can list the files in a directory with the `dir()` command.)  


```{r}
f  <- uqsa::uqsa_example("AKAR4") 
sb <- SBtabVFGEN::sbtab_from_tsv(f)
```

From the `sb` variable (list of data frames with the file contents) we can create an ordinary
differential equation, this will print lots of text about how it
interprets the contents (printouts are omitted here):

```{r include=FALSE}
m  <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

```{r eval=FALSE}
m  <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

We also load the experimental data:

```{r data}
ex <- SBtabVFGEN::sbtab.data(sb,m$conservationLaws)
```

We have loaded these constructs:

```{r summary}
print(names(sb))
print(names(ex))
print(names(m))
```

## Generate C ODE-code for model simulation

We next take `m` and generate c code from it, then write this code into a file (`AKAR4_gvf.c`):

```{r code}
C <- uqsa::generateCode(m)
cat(C,sep="\n",file="./AKAR4_gvf.c")
```

The `ex` variable holds the experimental data and instructions about
how they correspond to model simulations (`ex` is a list).

Next, we check that the c sources exist in the specified location,
that the model compiles on the current system, with the default c
compiler (called by `checkModel`) and a shared library can be built:

```{r check}
modelName <- checkModel("AKAR4",modelFile="./AKAR4_gvf.c")
```

The function `checkModel` also appends the location of the shared library as a comment to its return value.

For a test simulation we create the function `s` which will remember the experiments to simulate and the model to simulate them with. The function/closure `s` maps parameter vectors to model solutions (trajectories):

```{r test}
s <- simcf(ex,modelName,parMap=log10ParMap) # or simulator.c
p <- log10(sb$Parameter[["!DefaultValue"]])
names(p) <- rownames(sb$Parameter)
y <- s(p) # here the simulation happens
```

We can plot the simulated output together with the experimental data and experimental error:

```{r plot}
ylab <- colnames(ex[[1]]$outputValues)
d_ <- ex[[1]]$outputValues[[1]]
e_ <- ex[[1]]$errorValues[[1]]
t_ <- ex[[1]]$outputTimes
f_ <- y[[1]]$func[1,,1]
par(bty="n")
plot(t_,d_,type="p",main=names(ex)[1],xlab="time",ylab=ylab,ylim=c(100,200))
arrows(t_,d_,t_,d_+e_,angle=90,length=0.01)
arrows(t_,d_,t_,d_-e_,angle=90,length=0.01)
lines(t_,f_,lw=3)
```

## Take a parameter Sample for this Model

We use just one core as this is a tiny model. No MPI, no parallel tempering. First we construct the prior:

```{r prior}
sd <- rep(2,length(p))
dprior <- dNormalPrior(p,sd)
rprior <- rNormalPrior(p,sd)
```

Next, we set up the loglikelihood function and the function `mcmcUpdate` which will automatically determine that the
Metropolis Hastings Algorithm should be used with these inputs. Finally we construct the function 'MH_MCMC' (below), which actually performs the MCMC run, this takes the inital values, number of steps and the stepsize as arguments.

```{r likelihood}
llf <- logLikelihoodFunc(ex)
metropolis_hastings <- mcmcUpdate(s,ex,logLikelihood=llf,dprior=dprior)
MH_MCMC <- mcmc(metropolis_hastings)
```

Finally, to obtain a sample of parameters:

```{r sample, fig.width=12, fig.height=12, dpi=100}
x <- mcmcInit(
    1.0,
    parMCMC=p,
    simulate=s,
    logLikelihood=llf,
    dprior=dprior)

h <- 5e-2
N <- 3e4
S <- MH_MCMC(x,N,eps=h)
colnames(S) <- names(p)
print(attr(S,"acceptanceRate"))

if (require(hexbin)){
    hexplom(S)
} else {
    pairs(S)
}
```

## Better Sample

We can try to obtain a higher quality sample, through simple means:

1. Combine several samples (replicas) into one big sample
2. Use several random starting locations
3. Remove burn-in phase from each chain
4. Adjust the step size at least once to get close to 25% acceptance rate

```{r betterSample}
A <- function(a) {
    return(0.5 + a^4/(0.25^4 + a^4))
}

set.seed(137)


nCores <- parallel::detectCores()
bigSample <- parallel::mclapply(
    seq(nCores),
    \(i) {
        x <- mcmcInit(
            1.0,
            parMCMC=t(rprior(1)),
            simulate=s,
            logLikelihood=llf,
            dprior=dprior)
        ## adjust acceptance rate to 25% via step-size h
        for (i in seq(30)){
            z <- MH_MCMC(x,200,h)
            x <- attr(z,"lastPoint")
            h <- h*A(attr(z,"acceptanceRate"))
        }
        return(MH_MCMC(x,N,h))
    },
    mc.cores=nCores
)

S_ <- Reduce(\(a,b) {rbind(a,b)},bigSample,init=NULL)
L_ <- Reduce(\(a,b) {c(a,attr(b,"logLikelihood"))},bigSample,init=NULL)
colnames(S_) <- names(p)

if (require(hadron)){
    ac <- hadron::uwerr(data=L_,nrep=rep(N,nCores),pl=TRUE)
    tau <- ceiling(ac$tauint+ac$dtauint)
    i <- seq(1,NROW(S_),by=tau)
} else {
    i <- seq(1,NROW(S_),by=100)
    plot(L_,type="l",xlab="mcmc index",ylab="log-likelihood")
}
```
```{r pairs-2, fig.width=12, fig.height=12, dpi=100}
if (require(hexbin)){
    hexbin::hexplom(S_[i,])
} else {
    pairs(S_[i,])
}
```

The code above can of course also be done in a loop (sequentially), and `rbind` within the loop.

Let us also plot the log-likelihood

```{r}
plot(L_,main="Big Sample for AKAR4",xlab="Markov chain index",ylab="log-likelihood", type="l")
```

This should be enough for this model size to be used as a basis for uncertainty quantification. For bigger models we turn
to either more advanced MCMC algorithms, such as [SMMALA](smmala.html)or [parallel tempering](mpi.html) combined with the Metropolis-Hastings algorithm.

## Alternative to `checkModel()`

The steps described above to compile the model with the `checkModel`
function can also be done outside of R, by running an appropriate
command, e.g.:

```{sh, eval=FALSE}
# alternative, directly in the shell:
cc -shared -fPIC -o AKAR4.so ./AKAR4_gvf.c -lgsl -lgslcblas -lm
```


