---
title: "SBtab"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

The R package `uqsa`, typically imports systems biology models and
corresponding data stored in [SBtab](https://www.sbtab.net)
files. SBtab is a table format for system biology. The use of SBtab is
optional, but very practical. SBML models currently cannot be imported
by this package, but can be converted to SBtab using the online tools
that [sbtab.net](https://www.sbtab.net) provides. The content of SBML
files is more strict than with SBtab files, so the files created by
conversion are not guaranteed to work here. They are user-friendly,
i.e. a human can open SBtab files with a text editor and edit them
manually.

# Advantages of SBtab

Data and models are tightly linked and we also include two tables on
the relationships between the data and model simulations in all our
examples:

1. *Experiments*, a table with a list of experiments together with appropriate model settings
2. *Output*, a list of output functions that correspond to measurable
   values for this model, named the same as the corresponding data
   columns

An SBtab Document is a set of tables that represent reactions,
compounds, parameters, and measured data that correspond to
simulations of the model under certain input conditions and initial
values. Input conditions are often related to initial values and
values of the input parameters (parameters that are used within the
model of the input signal). The values of the input parameters are
always known, while the model' s internal parameters (e.g. reaction
rate coefficients) can be subject to optimization or sampling.

SBML level 2 lacks the ability to carry data with the model and also
lacks the vocabulary to define which parameters are *inputs* and which
are *model internal*. There is also no way to define an *observable*
in SBML (a function that models the measurement device, an
observation). For this reason, an SBML document cannot be converted
into an SBtab document that has these qualities (which we need).

# Issues to be aware of with any File format

According to the specifications of the SBtab authors, MS Excel
spreadsheets are an acceptable storage format - but we don't use any
of their original code and therefore do not carry over any guarantees.

Our [SBtabVFGEN](icpm-kth/SBtabVFGEN) package uses either multiple TSV
files or one [ODS](https://en.wikipedia.org/wiki/OpenDocument)
file. The TSV reading is done using core R functions, while ODS is
read through
[readODS](https://cran.r-project.org/package=readODS). Both formats
can have issues that are general and not specific to this package:

- TSV
    - line endings can be `\n` or `\n\r`
    - fields could be automatically and unnecessarily quoted by a spreadsheet software
    - encoding (UTF-8, or something else)
    - some spreadsheets may want to write a unicode minus sign `−` rather than `-` into a tsv file: `U+2212`, be careful
    - lines can be blank, but not quite, if they contain just tabs or spaces
- ODS
    - comments inside fields could be imported as field content and confuse the model parser
    - same issue with unicode minus signs
    - `readODS` could theoretically be discontinued (deprecated), despite the format continuing to exist

Models can be automatically converted between the SBtab format and
other modelling formats (such as SBML). For more information about
SBtab please refer to [the official git
repository](https://github.com/tlubitz/SBtab). Be aware that the
conversion from SBML to SBtab has to be done with the official SBtab
tools, not ours (SBML is hard).

For the most part, in our case, an SBtab document is a collection of
[tsv](https://en.wikipedia.org/wiki/Tab-separated_values) files (has
to be one file per table). This can be written by hand, in a text
editor (one that doesn't auto convert tabs to *N* spaces).  format.
Each table contains information about the model, data, and their
relationship to one another:

- Reactions
- Parameters
- Output Functions
- Compounds/Molecular Species

Each type of items (e.g. parameters) gets a TSV file
(e.g. `Parameter.tsv`), see our examples.

From these files you can automatically generate ODE code for R and C
solvers, and load the data from them.

# SBtab File Structure

Each file has a header, e.g.:

```txt
!!SBtab SBtabVersion='1.0' TableName='Reaction' TableType='Reaction' TableTitle='A list of Reactions' Document='myModel'
```

We don't parse the version attribute or any other attribute other than
`TableName` and `Document`. We use the `Document` attribute to assign
a name to each model we read from the files; it is stored as a comment
attached to the list of data.frames `SBtabVFGEN::sbtab_from_tsv()`
generates, `comment(model.tab)` will print it.

The second line contains the table column headers. The first column
can be called either `!ID` or `!Name` (it is harmless to have both,
but only the first one counts). If the first column is not called
that, the scripts will carry on regardless, assuming that it is some
kind of name (so `!Id`, `!id`, or `ID?` may also work). All other
columns are found by name, exactly. All other columns can appear in
any order. Any amount of columns can be added with names not starting
with exclamation points, they are ignored.

In the following sub-sections we will list the minimum set of columns
for each table, the sub-section heading refers to the required
`TableName` attribute for this kind of table. The set of columns are
given as example tables. Each time we omit the `!!SBtab` heading from the table to not mess up formatting.

There are two generic table-types that can be used if no specific one
is applicable: `TableType='Quantity'` and
`TableType='QuantityMatrix'`. We usually don't use the `TableType`
property at all, but it is needed for the official
([sbtab.net](www.sbtab.net)) tool-set to work.

If there is a specific table-type setting for a table, we list that
setting in the following sections.

## Constant

A list of constants, values that never change for this model.

`SBtab TableName='Constant' Document='myModel'`
|!ID|!Unit|!Value|
|--:|:---:|:-----|
|a|µM|3.9|
|...|...|...|

## Input

Known paramters of the model, can vary between experiments.

`SBtab TableName='Input' Document='myModel'`
|!ID|!DefaultValue|!Unit|
|--:|:-----------:|:---:|
|CalciumBaseLevel|2000.0|nM|
|...|...|...|

## Parameter

Possibly unknown parameters of the model, these parameters typically
refer to internal properties of the model itself rather than something
we did to the model during an experiment.

`SBtab TableName='Parameter' Document='myModel'`
|!ID|!DefaultValue|!Std|!Unit|!Scale|
|--:|------------:|:---|:----|:----:|
|kf1|-1.8|0.1|s^(-1)|natural logarithm|

Here the `!Scale` column is optional, but often useful; several values
are possible: `log10`,`log`,`linear` (in various spellings). If
missing `linear` is assumed.

## Expression

mathematical sub-expressions that can be used in reaction fluxes. Can
be used to encode thermodynamic relationships between parameters,
model an input signal. This is used to assign a name to an algebraic
expression.

`SBtab TableName='Expression' Document='myModel'`
|!ID|!Unit|!Formula|
|--:|:---:|:-------|
|KD_1|nM|(KD_3*KD_4)/KD_2|
|...|...|...|

## Compound

In SBML language, these are *species*.

`!!SBtab TableName='Compound' TableType='Compound' Document='myModel'`
|!ID|!Unit|!InitialValue|
|--:|:---:|:------------|
|PP2B_CaM|uM|0.0|
|...|...|...|

There are several optional columns, e.g. `!IsConstant` and
`!Type`. These variables are usually used as state variables in an ODE
framework.

## Reaction

`!!SBtab TableName='Reaction' TableType='Reaction' Document='myModel'`
|!ID|!KineticLaw|!IsReversible|!ReactionFormula|
|--:|:---------:|:-----------:|:--------------:|
|Reaction1|kf*A*B-kb*C|TRUE|A + B <=> C|
|...|...|...|

## Output

`!!SBtab TableName='Output' Document='myModel'`
|!ID|!Unit|!Formula|
|--:|:---:|:-------|
|CaPerCaM|1|totalCa/totalCaM|
|...|...|...|

A unit of `1` means the same as `'dimensionless'` in SBML.

## Event

Events will interrupt the solver in the rgsl package, change the valuse of input parameters, or state variables, and then re-initialize the ODE solver to continue from there.
`!!SBtab TableName='ActivationEvent' Document='myModel'`
|!TimePoint|!Time|>SET:Calcium|
|--:|:---:|:-------|
|Event0Time0|0.0|200|
|...|...|...|

Possible operations in >*OP*:*ID*: `SET`,`ADD`,`SUB`,`MUL`,`DIV`,
while ID can refer to any parameter or state variable (compound).

# Alternatives

A user can circumvent this entire format by just writing the C code or
R code for the model by hand or an entirely different tool, such as
[VFGEN](https://warrenweckesser.github.io/vfgen/) itself, or an SBML
related project like
[SBFC](https://www.ebi.ac.uk/biomodels/tools/converters/) that
generates code. Or, if the model is small enough, you can also write
these files using a text editor without any other tool.

The data we load from SBtab files is stored as a list in R, each item
is itself a list of data.frames or vectors (it's not an opaque
object). A user can just create such a list (e.g. in an R script), if
they want to avoid SBtab. If the data is stored in an entirely
different format, such as an hdf5 file, it could be read using e.g. hdf5r
and re-roganized into a list like this.
