---
title: "Simulation of a Stochastic Model for the AKAR4 reaction network"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = FALSE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(errors)
```

This article provides code to simulate the AKAR4 stochastic model; one
time, no sampling.

# The Stochastic Model

When the copy number of molecular species in a reaction network system
(AKAR4 in our case) is low, we cannot model the amount of molecules
deterministically (e.g., with an ODE model), because the stochasticity
in the reactions that take place cannot be ignored. In particular, the
time at which reactions take place is random, as well as the specific
reactions that take place (i.e., what pair of molecules react). To
model such system we can use the *chemical master equation* (CME): we
model the (integer) number of each molecule species in the system
(e.g., proteins) and how the number of each molecule type (randomly)
evolves in time, as a Markov jump process. We can easily sample
trajectories from this stochastic model using *Gillespie's
Stochastic Simulation algorithm*. Given the current amount of each
molecular species in the system at a given time point, we can sample
the time at which the next reaction takes place and we can sample the
type of reaction (i.e., what pair of molecules react).

However, we don't expect the model files the systems biology content
is stored in to change. Instead, we convert all information given in
the TSV files into a CME model, with particle counts and propensities
(automatically). Not all features of an ODE simulations are easy to
transfer into the CME formalism: conservation law analysis, and event
based transformations are currently not available.

To obtain the stochastic model for the reaction network, we need to
determine the reaction _propensities_. These can be derived from the
reaction rate coefficients: the parameters that we usually
use in our models, and on which we perform uncertainty
quantification. To derive the reaction propensities we referred to
[this article](https://doi.org/10.1186/1752-0509-5-187).


# Load the Model

The AKAR4 model is included with the package.

```{r load-model}
f <- uqsa_example("AKAR4") # or some other files
m <- model_from_tsv(f)
cmeModel <- as_cme(m)
C <- generate_code(cmeModel,LV=1e8)
c_path(cmeModel) <- write_c_code(C)
so_path(cmeModel) <- shlib(cmeModel)
```

The `shlib` function creates a shared library, and stores the path to
it in the `cmeModel` object.

To generate the code, we decide on a value for `LV`: Avogadro's number
multiplied by the system's volume. Higher numbers mean more particles
(more reactions per time-span). This mumber can be changed directly in
the source code of the model (the C file), afterwards, because it is
used as a global constant in the C file.

# Load Experiments (data)

```{r experiments}
ex <- experiments(m) # same as with ODE models
```

# Simulate

Function `simstoch` will output a function `s`, which will always
simulate the scenarios described in the experiments `ex` (i.e., same
initial conditions, same inputs), but for user supplied
parameters. This is the same structure as with `simulator.c` for ODEs

We auto-generate C code, which uses the kinetic laws, but rescales all
parameters internally to have the unit `1/s` and all compound species
are rescaled to be in particle counts. So, the internal simulator can
be used with the list of experiments that we use for ODE models, with
no changes. This rescaling is performed as described in [this
article](https://doi.org/10.1186/1752-0509-5-187).

## Simulation

We simulate this compiled model, using the shared library:

```{r simulate}
p <- values(m$Parameter)
st <- simstoch(ex,cmeModel)
y <- st(p)
```

The second argument can also be the path to the shared library. If it is `cmeModel`, the path is extracted from it using:

```{r path-print}
print(so_path(cmeModel))
```

## Plot the results

```{r plot-simstoch, fig.height=12, fig.width=7, out.width="100%"}
tm <- ex[[1]]$outputTimes

par(mfrow=c(3,1),bty="n")
for (i in seq_along(ex)){
plot(
	as.errors(tm),
	ex[[i]]$data,
	xlab="time",
	ylab="AKAR4p",
	main=names(ex)[i],
	ylim=c(90,210)
)
lines(
	tm,
	y[[i]]$func,
	lwd=2,
	col="red",
	type='s'
)
}
```

## Benchmark

Stochastic simulations are slower than ODE simulations, this is a
benchmark of the builtin stochastic solver:

```{r benchmark}
if (require(rbenchmark)){
	BM <- benchmark(
		uqsa = {y <- st(p)},
		replications = 5000
	)
	print(BM)
}
```
