---
title: "Simulation of a Stochastic Model for the AKAP79 reaction network"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(SBtabVFGEN)
library(uqsa)
library(GillespieSSA2)
```

This article provides code to simulate the AKAP79 stochastic model (one time, no
sampling). We are plotting the model with default parameters which
are not expected to fit the data (this is the starting point).

# The Stochastic Model

When the copy number of molecular species in a reaction network system
(AKAP79 in our case) is low, we cannot model the amount of molecules
deterministically (e.g., with an ODE model), because the stochasticity
in the reactions that take place cannot be ignored. In particular, the
time at which reactions take place is random, as well as the specific
reactions that take place (i.e., what pair of molecules react). To
model such system we can use the *chemical master equation* (CME): we
model the (integer) number of each molecule species in the system
(e.g., proteins) and how the number of each molecule type (randomly)
evolves in time. The likelihood of this model is hard to compute;
however, we can easily sample trajectories from this stochastic model
using the *Gillespie algorithm*. Given the current amount of each
molecular species in the system at a given time point, we can sample
the time at which the next reaction takes place and we can sample the
type of reaction (i.e., what pair of molecules react).

To obtain the stochastic model for the reaction network, we need to
compute the reaction _propensities_. These can be related to the
reaction rate coefficients, that are the parameters that we usually
use in our models, and on which we perform uncertainty
quantification. To derive the reaction propensities we referred to
[this article](https://doi.org/10.1186/1752-0509-5-187).


# Load the Model

This model is included with the package. To load your own model, see
the article ["Build and simulate your own Model"](user_model.html).

```{r}
m <- model_from_tsv(uqsa_example("AKAP79"))
cme <- as_cme(m)    # chemical master equation, stochastic model
c_path(cme) <- write_c_code(generate_code(cme))
so_path(cme) <- shlib(cme)
```

# Load Experiments (data)

```{r, experiments}
ex <- experiments(m)
print(ex)
```

Here, we set parameters that we obtained as a result of uncertainty quantification as the default parameters don't produce a good fit, they are still not optimal:
```{r parameters}
p <- c(3.38,-0.22,-0.39,0.0013,7.89e-2,-1.02,-1.08,-2.86,-0.53,-0.34,-0.51,-2.42,-1.05,-1.37,-1.29,2.08,-2.79,-0.87,0.26,-0.168,-0.331,-1.77,-0.938,1.065,2.08,0.0147,-0.09893)
names(p) <- rownames(m$Parameter)
```

If instead we wanted to use the default parameters, then we could use this code:
```{r defaults, eval=FALSE}
p <- values(m$Parameter)
```

# Simulate

Function `simstoch` will return a function `s`, which will always
simulate the scenarios described in the list of experiments (i.e.,
same initial conditions, same inputs), but for user supplied
parameters.

```{r, sim}
E <- 9 #index of the experiment to consider

# generate a function (s) that simulates a trajectory given a parameter in input
s <- simstoch(ex, cme, parMap=log10ParMap)

# simulate a trajectory (y) given parameter p
ystc <- s(p)
```

# Simulation Results via `ggplot2`

```{r, ggplot}
require(ggplot2)
require(errors)

D<-data.frame(
    time=ex[[E]]$outputTime,
    AKAP79=as.numeric(ex[[E]]$data),
    AKAP79ERR=as.numeric(errors(ex[[E]]$data)),
    sim=as.numeric(ystc[[E]]$func[1,,1]))
ggplot(D) +
  geom_linerange(mapping=aes(x=time,y=AKAP79,ymin=AKAP79-AKAP79ERR,ymax=AKAP79+AKAP79ERR),na.rm=TRUE) +
  geom_point(mapping=aes(x=time,y=AKAP79),na.rm=TRUE) +
  geom_line(mapping=aes(x=time,y=sim),color="purple",lwd=1.2)
```

```{r referenceODEsimulation}
o <- as_ode(m)
c_path(o) <- write_c_code(generate_code(o))
so_path(o) <- shlib(o)
sode <- simfi(ex,o,parMap=log10ParMap)
yode <- sode(p)
```

And also plot the results:

```{r plot_simstoch}
tm <- ex[[E]]$outputTimes
par(bty="n")
plot(
    as.errors(tm),
    ex[[E]]$data,
    xlab="time",
    ylab="AKAR4p",
    main=names(ex)[E],
    ylim=c(90,210)
)

lines(tm,as.numeric(ystc[[E]]$func["AKAR4pOUT",,1]),lwd=1,col="blue")
lines(tm,as.numeric(yode[[E]]$func["AKAR4pOUT",,1]),lwd=3)
```
