---
title: "Simulation of a Stochastic Model for the AKAR4 reaction network"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = FALSE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(errors)
```

This article provides code to simulate the AKAR4 stochastic model (one
time, no sampling).

# The Stochastic Model

When the copy number of molecular species in a reaction network system
(AKAR4 in our case) is low, we cannot model the amount of molecules
deterministically (e.g., with an ODE model), because the stochasticity
in the reactions that take place cannot be ignored. In particular, the
time at which reactions take place is random, as well as the specific
reactions that take place (i.e., what pair of molecules react). To
model such system we can use the *master equation*: we model the
(integer) number of each molecule species in the system (e.g.,
proteins) and how the number of each molecule type (randomly) evolves
in time. We can easily sample trajectories from this stochastic model
using the *Gillespie's Stochastic Simulation algorithm*. Given the
current amount of each molecular species in the system at a given time
point, we can sample the time at which the next reaction takes place
and we can sample the type of reaction (i.e., what pair of molecules
react).

To obtain the stochastic model for the reaction network, we need to
determine the reaction _propensities_. These can be derived from the
reaction rate coefficients (which are the parameters that we usually
use in our models and on which we perform uncertainty
quantification). To derive the reaction propensities we referred to
[this article](https://doi.org/10.1186/1752-0509-5-187).


# Load the Model

The AKAR4 model is included with the package. To load your own model, see
the article ["Build and simulate your own Model"](user_model.html) article.

```{r load-model}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m,cla=FALSE)
sm <- makeGillespieModel(m)
C <- generateGillespieCode(sm,LV=1e8)
```

where `LV` is Avogadro's number multiplied by the system's
volume. Higher numbers mean more particles (more reactions per
time-span).

# Load Experiments (data)

```{r experiments}
ex <- experiments(m,o) # same as with ODE models
```

# Simulate

Function `simstoch` will output a function `s`, which will always
simulate the scenarios described in the experiments `ex` (i.e., same
initial conditions, same inputs), but for user supplied parameters, as
with `simulator.c` for ODEs

We auto-generate C code, which uses the kinetic laws, but rescales all
parameters internally to have the unit `1/s` and all compound species
are rescaled to be in particle counts. So, the internal simulator can
be used with the list of experiments that we use for ODE models, with
no changes. This rescaling is performed as described in [this
article](https://doi.org/10.1186/1752-0509-5-187).

We can also write code and compile that model to a shared library
manually:

```{r write-code, eval=TRUE}
print(length(C))
cat(C,sep="\n",file="./AKAR4_st.c")
```

In the shell:
```{sh shell-compile-code, eval=TRUE}
echo $0
cc -shared -fPIC -O2 -o akar4-st.so ./AKAR4_st.c -lm
```

```{r compile-code, eval=FALSE, echo=FALSE}
system2("cc",args=c("-shared", "-fPIC", "-march=native", "-O2", "-o", "akar4-st.so", "./AKAR4_st.c", "-lm"))
```

And next we simulate this compiled model:

```{r executeCode, eval=TRUE}
p <- values(m$Parameter)
st <- simstoch(ex,model.so="./akar4-st.so")
y <- st(p)
```

## Plot the results

```{r plot_simstoch}
tm <- ex[[1]]$outputTimes

par(bty="n")
plot(
	as.errors(tm),
	ex[[1]]$data,
	xlab="time",
	ylab="AKAR4p",
	main=names(ex)[1],
	ylim=c(90,210)
)
lines(
	tm,
	y[[1]]$func,
	lwd=2,
	col="red",
	type='s'
)
```

## Benchmark

Stochastic simulations are slower than ODE simulations, this is a
benchmark of the builtin stochastic solver:

```{r benchmark}
if (require(rbenchmark)){
	BM <- benchmark(
		uqsa = {y <- st(p)},
		replications = 5000
	)
	print(BM)
}

```
