---
title: "Example Simulation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(SBtabVFGEN)
library(uqsa)
```

This article provides code to simulate the AKAP79 model (one time, no
sampling).  We are plotting the model with default parameters which
are not expected to fit the data (this is the starting point).

# Load the Model

This model is included with the package. To load your own model, see
the [user model](user_model.html) article.

```{r}
modelFiles <- uqsa_example("AKAP79",full.names=TRUE)
SBtab <- SBtabVFGEN::sbtab_from_tsv(modelFiles)
modelName <- checkModel("AKAP79",uqsa_example("AKAP79",pat="_gvf[.]c$"))
comment(modelName)

# model related functions, in R, e.g. AKAP79_default() parameters
source(uqsa_example("AKAP79",pat='^AKAP79[.]R$'))
print(AKAP79_default())
```

Conservation laws (determined by `sbtab_to_vfgen` and saved as RDS file):

```{r}
clf <- uqsa_example("AKAP79",pat='Laws[.]RDS$')
cl <- readRDS(clf)
if (any(is.na(cl))) cl <- NULL
```

With conservation laws, some species are calculated
algebraically. Their initial values are turned into input parameters
(using the found law):

With this hypothetical relationship ($c$ is a constant):

$$A+B = c\,,$$

we can determine that

$$c = A_0 + B_0\,.$$

And thus we can replac either of the two species:

$$A(t) = A_0 + B_0 - B(t)$$

And $A_0+B_0$ are turned into an input called `A_ConservedConst` (the
$c$ above) with the value determined from the stated initial
condition.


# Load Experiments (data)

This also includes instructions for the simulator.

```{r, experiments}
experiments <- sbtab.data(SBtab) # with conservation laws
# for example, these are the input
# and initial state of experiment 1:
print(experiments[[1]]$input)
print(experiments[[1]]$initialState)
# pick prameters for simulation
nInput <- length(experiments[[1]]$input)
if (nInput>0) p <- head(AKAP79_default(),-nInput)
print(p)
```

# Simulate

This will make a function `s`, which will always simulate the
scenarios described in the `experiments` list, but for user supplied
parameters.

```{r, sim}
s <- simulator.c(experiments,modelName,noise=TRUE)
y <- s(p)
```

# Plot

```{r, plotting}
E <- 10 # which experiment to plot
out <- experiments[[E]]$outputValues$AKAR4pOUT
err <- experiments[[E]]$errorValues$AKAR4pOUT
tm <- experiments[[E]]$outputTime

par(bty='n',xaxp=c(80,120,4))
plot(tm,
     y[[E]]$func[1,,1],
     type='l',
     ylim=c(90,130), ylab="AKAR4p",
     xlab="t",
     main=sprintf("Experiment %i",E),
     lwd=1.5,
     col="purple"
)

points(tm,out)
arrows(x0=tm,x1=tm,y0=out,y1=out+err,angle=90,length=0.025)
arrows(x0=tm,x1=tm,y0=out,y1=out-err,angle=90,length=0.025)
```

# gg-Plot

```{r, ggplot}
require(ggplot2)

D<-data.frame(time=experiments[[E]]$outputTime,
              AKAR4p=experiments[[E]]$outputValues$AKAR4pOUT,
              AKAR4pERR=experiments[[E]]$errorValues$AKAR4pOUT,
              sim=y[[E]]$func[1,,1])
ggplot(D) +
  geom_linerange(mapping=aes(x=time,y=AKAR4p,ymin=AKAR4p-AKAR4pERR,ymax=AKAR4p+AKAR4pERR),na.rm=TRUE) +
  geom_point(mapping=aes(x=time,y=AKAR4p),na.rm=TRUE) +
  geom_line(mapping=aes(x=time,y=sim),color="purple",lwd=1.2)
```
