---
title: "Simulate AKAP79 deterministically and add noise"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(errors)
library(uqsa)
```

This article provides code to simulate the AKAP79 model (one time, no
sampling) also considering measurement noise.  We are plotting the model with default parameters which
are not expected to fit the data (this is the starting point).

# Load the Model

This model is included with the package. To load your own model, see
the article ["Build and simulate your own Model"](user_model.html).

```{r model-loading}
f <- uqsa_example("AKAP79",full.names=TRUE)
m <- model_from_tsv(f)
o <- as_ode(m)
c_path(o) <- write_c_code(generate_code(o))
so_path(o) <- shlib(o)
print(o)
```

Conservation laws (determined by `as_ode`):

```{r conservation-laws}
print(o$conservationLaws)
```

With conservation laws, some species are calculated
algebraically. Their initial values are turned into input parameters
(using the found law):

With this hypothetical relationship ($c$ is a constant):

$$A+B = c\,,$$

we can determine that

$$c = A_0 + B_0\,.$$

And thus we can replac either of the two species:

$$A(t) = A_0 + B_0 - B(t)$$

And $A_0+B_0$ are turned into an input called `A_ConservedConst` (the
$c$ above) with the value determined from the stated initial
condition.


# Load Experiments (data)

This also includes instructions for the simulator.

```{r experiments}
ex <- experiments(m)
print(m)
```

# Simulate

This will make a function `s`, which will always simulate the
scenarios described in the `experiments` list, but for user supplied
parameters.

```{r sim}
s <- simulator.c(ex,o,parMap=log10ParMap,omit=2)
p <- values(m$Parameter)
y <- s(p)
```

# Plot

```{r plotting}

e.g. <- 10 # which experiment to plot
tm <- ex[[e.g.]]$outputTime

par(bty='n',xaxp=c(80,120,4))
plot(
	as.errors(tm),
	ex[[e.g.]]$data[1,],
	type='l',
	ylim=c(90,130), ylab="AKAR4p",
	xlab="t",
	main=sprintf("Experiment %s",names(ex)[e.g.])
)
lines(tm,y[[e.g.]]$func[1,,1],lwd=1.5,col="purple")
```
