---
title: "Simulation of a Stochastic Model for the AKAR4 reaction network"
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(SBtabVFGEN)
library(uqsa)
library(GillespieSSA2)
```

This article provides code to simulate the AKAR4 stochastic model (one time, no
sampling). We are plotting the model with default parameters which
are not expected to fit the data (this is the starting point).

# The Stochastic Model

When the copy number of molecular species in a reaction network system (AKAP79 in our case) is low, we cannot model the amount of molecules deterministically (e.g., with an ODE model), because the stochasticity in the reactions that take place cannot be ignored. In particular, the time at which reactions take place is random, as well as the specific reactions that take place (i.e., what pair of molecules react). To model such system we can use the *master equation*: we model the (integer) number of each molecule species in the system (e.g., proteins) and how the number of each molecule type (randomly) evolves in time. The likelihood of this model is hard to compute; however, we can easily sample trajectories from this stochastic model using the *Gillespie algorithm*. Given the current amount of each molecular species in the system at a given time point, we can sample the time at which the next reaction takes place and we can sample the type of reaction (i.e., what pair of molecules react).

To obtain the stochastic model for the reaction network, we need to compute the reaction _propensities_. These can be related to the reaction rate coefficients, that are the parameters that we usually use in our models, and on which we perform uncertainty quantification. To derive the reaction propensities we referred to [this article](https://doi.org/10.1186/1752-0509-5-187).


# Load the Model

This model is included with the package. To load your own model, see
the article ["Build and simulate your own Model"](user_model.html) article.

```{r}
modelFiles <- uqsa_example("AKAR4",full.names=TRUE)
SBtab <- SBtabVFGEN::sbtab_from_tsv(modelFiles)

# model related functions, in R, e.g. AKAP79_default() parameters
source(uqsa_example("AKAR4",pat='^AKAR4[.]R$'))
print(AKAR4_default())
```

# Load Experiments (data)

```{r, experiments}
experiments <- sbtab.data(SBtab)

# for example, these is the initial state of experiment 1:
print(experiments[[1]]$initialState)

# pick parameters for simulation
p <- SBtabVFGEN::sbtab_quantity(SBtab$Parameter)
print(p)
```
# Simulate

Function `simulator.stoch` will output a function `s`, which will always simulate the
scenarios described in experiment `e` (i.e., same initial conditions, same inputs), but for user supplied parameters.

```{r, sim}
# generate a function (s) that simulates a trajectory given a parameter in input
ssa2 <- simulator.stoch(experiments[1], model.tab = SBtab, vol = 4e-16, nStochSim = 3, outputFunction=AKAR4_func)
# vol indicates the volume in m^3 where the reactions take place
# nStochSim indicates how many times we simulate the stochastic model; the output trajectory is the average of all (in this case nStochSim = 3) the simulated trajectories

# simulate a trajectory (y) given parameter p
y_ssa2 <- ssa2(p)
```


# Plot

```{r, plotting}
par(bty='n',xaxp=c(80,120,4))
plot(
    experiments[[1]]$outputTimes,
    experiments[[1]]$outputValues$AKAR4pOUT,
    ylim=c(90,200), ylab="AKAR4p", xlab="time",
    main=names(experiments)[1],
    lwd=1.5)
lines(experiments[[1]]$outputTimes, as.numeric(y_ssa2[[1]]$output), col="blue")
```

# Show Results via `ggplot2`

```{r, ggplot2}
if (require(ggplot2)){
D<-data.frame(time=experiments[[1]]$outputTime,
              AKAR4p=experiments[[1]]$outputValues$AKAR4pOUT,
              AKAR4pERR=experiments[[1]]$errorValues$AKAR4pOUT,
              sim=y_ssa2[[1]]$output)
ggplot(D) +
  geom_linerange(mapping=aes(x=time,y=AKAR4p,ymin=AKAR4p-AKAR4pERR,ymax=AKAR4p+AKAR4pERR),na.rm=TRUE) +
  geom_point(mapping=aes(x=time,y=AKAR4p),na.rm=TRUE) +
  geom_line(mapping=aes(x=time,y=sim),color="purple",lwd=1.2)
}
```

# Built-in simulator

This package (`uqsa`) has a built-in stochastic simulator which
assumes that the model in question comes from an SBtab file and is
formulated with concentrations and kinetic laws.

We auto-generate C code, which uses the kinetic laws, but rescales all
parameters internally to have the unit `1/s` and all compound species
are rescaled to be in particle counts. So, the internal simulator can
be used with the list of experiments that we use for ODE models, with
no changes. This rescaling is performed as described in [this article](https://doi.org/10.1186/1752-0509-5-187).

```{r generateCode}
C <- generateGillespieCode(SBtab)
cat(C,sep="\n",file="AKAR4_st.c")
```

We compile that model to a shared library:

```{sh compileCode}
cc -shared -fPIC -O2 -o akar4-st.so AKAR4_st.c -lm
```

And next we simulate this compiled model:

```{r executeCode}
st <- uqsa::simstoch(experiments,model.so="./akar4-st.so")
y <- st(p)
```

And also plot the results:

```{r plot_simstoch}
tm <- experiments[[1]]$outputTimes
ov <- experiments[[1]]$outputValues$AKAR4pOUT
ev <- experiments[[1]]$errorValues$AKAR4pOUT
par(bty="n")
plot(tm,ov,xlab="time",ylab="AKAR4p",main=names(experiments)[1],ylim=c(90,210))
arrows(tm,ov,tm,ov+ev,angle=90,length=0.02)
arrows(tm,ov,tm,ov-ev,angle=90,length=0.02)
lines(tm,y[[1]]$func["AKAR4pOUT",],lwd=2)
```

## Benchmark

Here we test which of the two simulates faster:

```{r benchmark}
if (require(rbenchmark)){
	BM <- benchmark(
		GillespieSSA2 = {y1<-ssa2(p)},
		uqsa = {y2 <- st(p)},
	replications = 500
	)
}
print(BM)
```
