---
title: "Simulating a model"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=FALSE,
  cache = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Reaction network models can be simulated as deterministic models or
stochastic models. In this article we show the deterministic
approach. An example of how to simualte from the stochastic model is
available in [this article](simAKAR4stochastic.html).

Given a reaction network model, we can use the law of mass action to
derive an ODE system that describes how the concentrations of the
compounds in the system change in time.

To simulate the reaction network model deterministically with UQSA you
can use the `simulator.c` function:

```{r simulate, eval=FALSE}
simulate <- simulator.c(ex, modelName, parMap = parMap)
```

This has created a [closure](https://en.wikipedia.org/wiki/Closure_(computer_programming)) (`simulate`), with a single argument `p`:

```{r run-simulations, eval=FALSE}
y <- simulate(p)
```

The `simulate` function remembers the experiments that it was created
with and produces results of the same length as `ex` (the list of
simulation experiments), with the same experimental conditions.

It is often convenient to modify the parameters before passing them to
the model. Here are possible reasons:

1. the uncertainty is log normal
   - you want to pass `exp(log(p) + rnorm(...))` to the model rather than `p` itself
2. the Markov chain is in log-space
   - the sampler uses `p`, but the model needs `10^p`
3. the model parameters are linearly dependent
   - we have to reliably pass `c(p[1]+p[2], p[2]+p[3], p[3]-p[1])` to the model, every time

In such cases, you can write a mapping function, and use the `parMap` argument-slot of `simulator.c`:

```{r, eval=TRUE}
library(uqsa)
library(errors)
library(parallel)

f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m)
C <- generateCode(o)
modelName <- write_and_compile(C)

ex <- experiments(m,o)


```

For example, here parameters are in log-space.

```{r, eval=TRUE}
simulate <- simulator.c(ex,modelName,parMap=logParMap)
p <- log(values(m$Parameter))
print(p)

rprior <- rNormalPrior(mean=p,sd=0.2)

P <- rprior(150)
print(dim(P))
```

All simulations happen here (timed):

```{r timed-simulation, eval=TRUE}
options(mc.cores=parallel::detectCores())

ti <- Sys.time()
y <- simulate(t(P))
tf <- Sys.time()

print(difftime(tf,ti))
```

The matrix `P` has columns of model parameters that are log-normally
distributed around OK-ish values (taken from the TSV file
`AKAR4/Parameter.tsv`):

```{r, label="log-normal-sample", eval=TRUE, out.width="100%", fig.height=16}
par(mfrow=c(3,1))
for (i in seq(ex)){
	tm <- ex[[i]]$outputTimes
	plot(               # data
		as.errors(tm),
		ex[[i]]$data,
		main=names(ex)[i],
		xlab="time",
		ylab='AKAR4pOUT'
	)
	matplot(            # simulation
		tm,
		y[[i]]$func['AKAR4pOUT',,],
		add=TRUE, type="l",
		lwd=3, lty=1, col=rgb(0,0,1,0.1)
	)
}
```
