---
title: "Simulating a model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=FALSE,
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

The main method for model simulation is:

```{r simulate}
simulate <- simulator.c(experiments, modelName, parMap = identity)`
```

This has created a closure (`simulate`), with a single argument `par`:

```{r}
sr <- simulate(par)  # simulation results
```

The `simulate` function remembers the experiments that it was created
with and produces results of the same length as `experiments`, in the same experimental conditions (e.g., same initial conditions).

It is often convenient to modify the parameters before passing them to
the model. Here are possible reasons:

1. the uncertainty is log normal
   - you want to pass `exp(log(p) + rnorm(...))` to the model rather than `p` itself
2. the Markov chain is in log-space
   - the sampler uses `p`, but the model needs `10^p`
3. the model parameters are linearly dependent
   - we have to reliably pass `c(p[1]+p[2], p[2]+p[3], p[3]-p[1])` to the model, every time
   
In such cases, you can write a mapping function, and use the `parMap` argument-slot of `simulator.c`:

```{r, eval=TRUE, include=FALSE}
library(uqsa)
library(SBtabVFGEN)

f <- uqsa_example("AKAR4cl")
cl <- readRDS(uqsa_example("AKAR4cl",f="RDS"))
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # includes the data

parMap <- function(p){
	return(exp(p))
}

modelName <- checkModel(comment(sb),uqsa_example("AKAR4cl",pat="_gvf.c$"))
sim <- simulator.c(ex,modelName,parMap)
```

```{r, eval=FALSE}
library(uqsa)
library(SBtabVFGEN)

f <- uqsa_example("AKAR4cl")
cl <- readRDS(uqsa_example("AKAR4cl",f="RDS"))
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # includes the data

parMap <- function(p){
	return(exp(p))
}

modelName <- checkModel(comment(sb),uqsa_example("AKAR4cl",pat="_gvf.c$"))
sim <- simulator.c(ex,modelName,parMap)
```

For example, here parameters are in log-space.

```{r, eval=TRUE}
t <- as.numeric(ex[[1]]$outputTimes)
par <- log(sb$Parameter[["!DefaultValue"]])
np <- length(par)

stdv <- 0.2
REPS <- 50
P <- matrix(rnorm(np*REPS,mean=par,sd=stdv),np,REPS)
dim(P)

stm <- Sys.time()
sr <- sim(P)
etm <- Sys.time()
difftime(etm,stm)
```

Now we sample trajectories associated with parameters in log-space:

```{r, label="log-normal-sample", eval=TRUE}
T <- rep(c(t,NA),REPS)                                  # the NA value will break the line
Z <- as.numeric(sr[[1]]$func[1,c(seq_along(t),NA),])    # at the end, so it doesn't loop
plot(T,Z,type='l',bty='n')
points(t,ex[[1]]$outputValues[[1]])
file.remove("AKAR4cl.so")
```
