---
title: "UQ for AKAR4 (stochastic model)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  eval = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(uqsa)
library(errors)
library(parallel)
```

We use a particle filter (aka SMC) on the stochastic AKAR4 model,
combined with ABC (due to the stochasticity of the model)

For information about the stochastic model, please read the article
[Simulate AKAR4 stochastically](simAKAR4stochastic.html).

# Load the Model

This model is included with the package, as one of the example models.

```{r load-files}
f <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
ex <- experiments(m)
```

## Make a Stochastic Version of AKAR4

```{r write}
sm <- as_cme(m)
C <- generate_code(sm)
c_path(sm) <- write_c_code(C)
so_path(sm) <- shlib(sm)
```

To simulate the stochastic reaction network model we use the
*Gillespie algorithm*.

# Define MCMC settings

```{r configure}
# scale to determine prior values
defRange <- 1000
par_val <- values(m$Parameter)

# Define Lower and Upper Limits for logUniform prior distribution for the parameters
ll <- c(par_val/defRange)
ul <- c(par_val*defRange)
ll = log10(ll) # log10-scale
ul = log10(ul) # log10-scale

# Define the prior distribution. In this case Uniform(ll,ul)
dprior <- dUniformPrior(ll, ul) # dprior evaluates the prior density function
rprior <- rUniformPrior(ll, ul) # rprior generates random samples from the prior
```

# Generate an objective function

The objective function simulates the model, to do this it needs a
simulator, it also needs direct access to the data, so `ex` is also an
explicit arument for `makeObjective`:

```{r makeObjective}
s <- simstoch(
	ex, # list of experiments
	sm, # stochastic model, *with* shared library entry
	parMap=log10ParMap
)
objFunc <- makeObjective(ex,s)
```

Test evaluation of the objective function:

```{r test-evaluation}
d <- objFunc(log10(values(m$Parameter)))
print(d)
```

It returns a vector of distance values (one per simulation
experiment). But, when supplied with several parameters, it returns
several columns:

```{r test-matrix-evaluation}
D <- objFunc(t(rprior(7)))
print(D)
```

# Run ABCSMC

We launch the particle filter from the prior (cold start), and pick
resonable values for `delta` (the ABC distance threshold).

The data has measurement errors (standard error). The distance
function takes the standard error into account, so the distances
returned are relative, and expected to be close to 1. The function
`ABCSMC` accepts two values for `delta`, an initial guess that is very
permissive and a final value that triggers the SMC algorithm to
terminate.

```{r abcsmc}
t_initial <- Sys.time()
options(mc.cores = parallel::detectCores())
X <- rprior(5000)    # this determines the sample size
colnames(X) <- names(par_val)
posterior <- ABCSMC(
	objFunc,
	startPar=t(X),
	Sigma=cov(X),
	delta=c(initial=3,final=0.5),
	dprior=dprior
)
t_final <- Sys.time()

cat(difftime(t_final,t_initial,units="min")," minutes")
```

# Visualize the sampled parameters

The parameter-values in `ABCMCMCoutput$draws` are approximate samples
from the posterior distribution, given the experiments.

The sampled three dimensional parameters for the AKAR4 model can be
visualized via pairs of scatter plots, or two dimensional binning:

```{r pairs, out.width="100%", fig.width=12, fig.height=12, dpi=100}
Z <- posterior$draws
if (requireNamespace("rgl")){
	plot3d(
		Z[,1],Z[,2],Z[,3],
		xlab=names(par_val)[1],
		ylab=names(par_val)[2],
		zlab=names(par_val)[3],
		type='p',
		col=rgb(0,0,1,0.5)
	)
	rglwidget()
} else if (requireNamespace("hexbin")){
	hexbin::hexplom(posterior$draws, xbins = 24)
} else {
	pairs(posterior$draws, pch=20)
}
```

## Marhinals

The marginal distributions of each parameter in the parameter vector
can be visualized via histograms as follows.

```{r marginals}
for(i in seq_along(par_val)){
	hist(
		posterior$draws[,i],
		breaks = 20,
		main=names(par_val)[i],
		xlab = "Value in log scale"
	)
}
```

## State Trajectories

```{r trajectories, fig.height=16, out.width="100%"}
t_initial <- Sys.time()
y <- s(t(Z)) # using the stochastic simulator one final time
t_final <- Sys.time()

cat(difftime(t_final,t_initial,units="min")," minutes")

par(mfrow=c(3,1))
for (i in seq_along(ex)){
	tm <- ex[[i]]$outputTimes
	plot(     # the data (errorbars)
		as.errors(tm),
		ex[[i]]$data,
		xlab="time",
		ylab="AKAR4p",
		main=names(ex)[i]
	)
	matplot(  # the simulation
		tm,
		drop(y[[i]]$func),
		type='s',
		col=rgb(50,50,200,1,max=255),
		lty=1,
		lwd=2,
		add=TRUE
	)
}
```
