---
title: "Uncertainty Quantification"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

# Introduction to UQ

Uncertainty Quantification and global Sensitivity Analysis are both
common tasks in *systems biology*. Uncertainty quantification aims to
establish the amount of knowledge we have (or indeed lack) about the
possible values of model parameters. The knowledge is expressed in the
form of probability distributions over the parameters, even though the
parameters are not *random variables*. In this case the probability is
related to the randomness within the observed of data: the measurement
noise. The noise has direct implications for the possible parameters:
parameters with high probability density values can easily explain the
data, while unlikely parameters require the random noise to have had
uncharacteristically large values for the given measurements

This package aims to characterize the posterior probability
distribution $p(\theta|D)$, where $\theta$ is a vector that directly
maps to the model parameters and is used as a Markov chain variable,
e.g.: $\rho = exp(\theta)$, where $\rho$ are the real model
parameters.

This target distribution quantifies the amount of knowledge we have
about $\theta$, and indirectly also about the internal parameters of
the model. In the field of systems biology, the internal parameters $\rho$
are typically quantities like reaction rate coefficients
$k_{\{f,b\}}$, dissociation constants (equilibrium constants) $K_{D}$,
Hill exponents $m$, and other parameters that relate to gene expression,
enzyme-substrate interaction, or other biochemical processes.

