---
title: "Uncertainty Quantification and global Sensitivity Analysis on AKAR4 (deterministic)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
library(SBtabVFGEN)
library(rgsl)
```

# Load Model and Data

First we load the model and data, for detailed explinations see previous examples.

```{r, eval=FALSE}
f  <- uqsa::uqsa_example("AKAR4")
sb <- SBtabVFGEN::sbtab_from_tsv(f)
m  <- SBtabVFGEN::sbtab_to_vfgen(sb, cla=FALSE)
ex <- SBtabVFGEN::sbtab.data(sb)
C <- uqsa::generateCode(m)
cat(C,sep="\n",file="./AKAR4_gvf.c")
modelName <- checkModel("AKAR4",modelFile="./AKAR4_gvf.c")
```

```{r, eval=TRUE, include=FALSE}
f  <- uqsa::uqsa_example("AKAR4")
sb <- SBtabVFGEN::sbtab_from_tsv(f)
m  <- SBtabVFGEN::sbtab_to_vfgen(sb, cla=FALSE)
ex <- SBtabVFGEN::sbtab.data(sb)
C <- uqsa::generateCode(m)
cat(C,sep="\n",file="./AKAR4_gvf.c")
modelName <- checkModel("AKAR4",modelFile="./AKAR4_gvf.c")
```

## Create simulator

We construct a simulator function and and test it on default parameter values.

```{r test}
s <- simcf(ex,modelName,parMap=log10ParMap) # or simulator.c
p <- log10(sb$Parameter[["!DefaultValue"]])
names(p) <- rownames(sb$Parameter)
y <- s(p) # here the simulation happens
```

## Uncertainty quantification: Sample the posterior distribution

For details on how to sample from the posterior distribution, see e.g. the deterministic [AKAR4](sampleAKAR4.html) example.
First we construct the prior:

```{r prior}
sd <- rep(2,length(p))
dprior <- dNormalPrior(p,sd)
rprior <- rNormalPrior(p,sd)
llf <- logLikelihoodFunc(ex)
metropolis_hastings <- mcmcUpdate(s,ex,logLikelihood=llf,dprior=dprior)
MH_MCMC <- mcmc(metropolis_hastings)
```

We obtain a high quality sample by combining several smaller samples using different random starting locations, removing the burn-in phase of each chain and adjust the stepsize at least once (for details see the [example](sampleAKAR4.html)). This will take a minute or so on a laptop.


```{r betterSample}
A <- function(a) {
    return(0.5 + a^4/(0.25^4 + a^4))
}

set.seed(137)
h <- 5e-2
N <- 3e4

nCores <- parallel::detectCores()
bigSample <- parallel::mclapply(
    seq(nCores),
    \(i) {
        x <- mcmcInit(
            1.0,
            parMCMC=t(rprior(1)),
            simulate=s,
            logLikelihood=llf,
            dprior=dprior)
        ## adjust acceptance rate to 25% via step-size h
        for (i in seq(30)){
            z <- MH_MCMC(x,200,h)
            x <- attr(z,"lastPoint")
            h <- h*A(attr(z,"acceptanceRate"))
        }
        return(MH_MCMC(x,N,h))
    },
    mc.cores=nCores
)

S_ <- Reduce(\(a,b) {rbind(a,b)},bigSample,init=NULL)
L_ <- Reduce(\(a,b) {c(a,attr(b,"logLikelihood"))},bigSample,init=NULL)
colnames(S_) <- names(p)

```


```{r pairs-2, fig.width=12, fig.height=12, dpi=100}
if (require(hexbin)){
    hexbin::hexplom(S_)
} else {
    pairs(S_)
}
```


## Sensitivity analysis on the posterior distribution

First we need to (re-)create the simulation result corresponding to the posterior parameter sample, now we set up the model without conservation laws so that we get output trajectories for all compounds. This will take a minute or so on a laptop.

```{r simulation}
s <- simcf(ex,modelName,parMap=log10ParMap) # or simulator.c
y <- s(t(S_[,]))
```

Next we calculate the first order sensitivity indexes. Here for a specific experimental setting and time point:

```{r globalSensitivity}
E <- 2                      # Experiment idx
T <- 60                     # Time idx
fM <- y[[E]]$state[,T,]
SIappr <- globalSensitivity(S_, t(fM), nBins = "Sturges")
```

and plot:

```{r barplot-1}
cols=rainbow(3)
par(mfrow = c(1, 1))
fM <- y[[E]]$state[,T,]
barplot(t(SIappr),
        col=cols,
        border="white",
        space=0.04,
        cex.axis=0.7,
        legend.text=sb$Parameter[,1])

```

Or for all three experiments:

Next we calculate the first order sensitivity indexes. Here for a specific timepoint and all three experimental conditions.

```{r barplot-2}
T <- 200                    #Time idx
par(mfrow = c(1, 3))
for (E in 3:1){
fM <- y[[E]]$state[,T,]
SIappr <-globalSensitivity(S_, t(fM) , nBins = "Sturges")
lgd <- list(sb$Parameter[,1], NULL,NULL)
barplot(t(SIappr),
        col=cols,
        cex.names = 0.7,
        las=2,
        border="white",
        space=0.04,
        cex.axis=1,
        main=names(ex)[E],
        legend.text=lgd[[E]])
}
```

Now we investigate a number of times instead of different species:

```{r barplot-3}
E <- 3
timePts <- seq(2,40,by=4)  #Avoid first time point
nStates=dim(y[[E]]$state)[1]
cols=rainbow(3)
par(mfrow = c(1, nStates))
lgd <- rep(list(NULL),nStates)
lgd[[nStates-1]] <- sb$Parameter[,1]
for (Cm in 1:nStates) {
  fM <- y[[E]]$state[Cm,timePts,]
  SIappr <-globalSensitivity(S_, t(fM) , nBins = "Sturges")
  names(SIappr) <- timePts
  barplot(t(SIappr),
          col=cols,
          names.arg = timePts,
          border="white",
          cex.axis=1,
          main=dimnames(y[[E]]$state)[[1]][Cm],
          xlab="Time",
          ylab="SI",
          legend.text=lgd[[Cm]],
          args.legend = list(x = "topright", inset=c(-0.1, 0), cex=0.7)
          )
}
mtext(names(ex[E]), side=3, line = - 1.5,outer=TRUE)
```
