---
title: "Uncertainty Quantification and global Sensitivity Analysis on AKAR4 (deterministic)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uqsa)
```

# Load Model and Data

First we load the model and data, for detailed explanations see
previous examples ([Simulate AKAR4 deterministically](./simAKAR4.html)).

```{r load}
f  <- uqsa_example("AKAR4")
m <- model_from_tsv(f)
o <- as_ode(m)
C <- generateCode(o)
modelName <- write_and_compile(C)
ex <- experiments(m,o)
```

## Create simulator

We construct a simulator function and and test it on default parameter values.

```{r test}
s <- simfi(ex,modelName,parMap=log10ParMap,omit=2) # or simulator.c
p <- log10(values(m$Parameter))
y <- s(p)
```

## Uncertainty quantification: Sample the posterior distribution

For details on how to sample from the posterior distribution, see, e.g., the [deterministic AKAR4 example](sampleAKAR4.html).
First we construct the prior. For this article, we override the information in the file and pick an iid normal prior with $\sigma=2$ for all parameters:

```{r prior}
sd <- rep(2,length(p))        # sigma = 2
dprior <- dNormalPrior(p,sd)
rprior <- rNormalPrior(p,sd)
mh <- mcmc(                   # metropolis hastings
	metropolis_update(        # random walk
		s,
		dprior=dprior,
		Sigma=diag(1/sd^2)
	)
)
```

We obtain a high quality sample by following the steps described in details in the [deterministic AKAR4 example](sampleAKAR4.html)). This will take a minute or so on a laptop.


```{r betterSample}
set.seed(137)
x <- mcmc_init(beta=1.0,p,s,ll,dprior)
h <- tune_step_size(mh,x)
N <- 3e4

nCores <- parallel::detectCores()

bigSample <- parallel::mclapply(
	seq(nCores),
	function(i) {
		return(mh(x,N,h))
	},
	mc.cores=nCores
)

S_ <- Reduce(\(a,b) {rbind(a,b)},bigSample,init=NULL)
L_ <- Reduce(\(a,b) {c(a,attr(b,"logLikelihood"))},bigSample,init=NULL)
colnames(S_) <- names(p)
```

We should estimate the auto-correlation of this sample, to reduce the
size effectively, for plotting and post-processing:

```{r auto-correlation}
if (require(hadron)){
	res <- uwerr(data=L_)
	tau <- ceiling(res$tauint + res$dtauint)
} else {
	A <- acf(L_,lag.max=1000)
	i <- which(A$acf < exp(-3))[1]
	tau <- ceiling(sum(A$acf[seq(i)]))
}

print(tau)

if (1 < tau && tau < length(L_)){
	N <- length(L_)
	S_ <- S_[seq(1,N,by=tau),]
	L_ <- L_[seq(1,N,by=tau)]
	N <- length(L_)
}
```


Pairs plot (either with hexagonal binning or normal)

```{r pairs-2, fig.width=12, fig.height=12, dpi=100}
if (require(hexbin)){
	hexbin::hexplom(S_)    # binning
} else {
	pairs(
		S_,                # the sample
		pch=16,            # filled circle
		ces=4,             # so that points overlap
		col=rgb(0,0,1,0.1) # transparent blue
	)
}
```

## Sensitivity analysis on the posterior distribution

First we need to (re-)create the simulation result corresponding to the posterior parameter sample. This will take a minute or so on a laptop.

```{r simulation}
y <- s(t(S_))
```

Next we calculate the _first order sensitivity indexes_. Here for compute them for a specific experimental setting and time point:

```{r globalSensitivity}
E <- 2                      # Experiment idx
T <- 60                     # Time idx
fM <- y[[E]]$state[,T,]
SIappr <- gsa_binning(S_, t(fM))
```

and plot the result:

```{r barplot-1}
cols=rainbow(3)
par(mfrow = c(1, 1))
fM <- y[[E]]$state[,T,]
barplot(
	t(SIappr),
	col=cols,
	border="white",
	space=0.04,
	cex.axis=0.7,
	legend.text=names(p)
)

```

Here for compute the first order sensitivity indexes for all three experiments, looking at the different compounds in the system (on the x axis):

```{r barplot-2}
T <- 200                    #Time idx
par(mfrow = c(1,3))
for (E in seq(3,1,-1)){
fM <- y[[E]]$state[,T,]
SIappr <- gsa_binning(S_, t(fM))
lgd <- list(names(p), NULL,NULL)
barplot(
	t(SIappr),
	col = cols,
	cex.names = 0.7,
	las = 2,
	border = "white",
	space = 0.04,
	cex.axis = 1,
	main = names(ex)[E],
	legend.text = lgd[[E]])
}
```

Now we investigate the first order sensitivity indexes at different time points (time is on the x axis) for each of the compounds and one experiment.

```{r barplot-3}
E <- 3
timePts <- seq(2,40,by=4)  #Avoid first time point
nStates=dim(y[[E]]$state)[1]
cols=rainbow(3)
par(mfrow = c(1, nStates))
lgd <- rep(list(NULL),nStates)
lgd[[nStates-1]] <- names(p)
for (Cm in seq(nStates)) {
	fM <- y[[E]]$state[Cm,timePts,]
	SIappr <-gsa_binning(S_, t(fM))
	names(SIappr) <- timePts
	barplot(
		t(SIappr),
		col=cols,
		names.arg = timePts,
		border="white",
		cex.axis=1,
		main=dimnames(y[[E]]$state)[[1]][Cm],
		xlab="Time",
		ylab="SI",
		legend.text=lgd[[Cm]],
		args.legend = list(x = "topright", inset=c(-0.1, 0), cex=0.7)
	)
}
mtext(names(ex[E]), side=3, line = - 1.5, outer=TRUE)
```
