---
title: "Build and simulate your own Model"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=FALSE, cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r, setup, eval=TRUE, include=FALSE}
library(uqsa)
Location <- dirname(uqsa_example('AKAR4cl')[1])
```

In this article we will assume that you have a reaction network model that you want to
work on and that you have written the model in the SBtab format. It may be a
Google spreadsheet, an open document spreadsheet (`.ods`), an MS Excel
file, a file for the "Numbers" application on an Apple device or a
`.gnumeric` file.

Firstly, you need to *build* your model, i.e., translate your reaction network model from the SBtab format into machine code that can be used for simulations in UQSA.

## Build your model

There are three steps we need to make to build the model:

|   | Steps                             | file (out) | tl;dr                |
|--:|:----------------------------------|:----------:|:--------------------:|
| 1 | convert SBtab to ODE              | [optional] | biology => math      |
| 2 | convert the ODE to C code         | .c         | math => code         |
| 3 | convert C code to shared library  | .so        | code => machine code |
---

The _optional_ file formats for ODEs are .vf, .mod, .ode, and plain text (.txt).

We generate C code in R for the SBtab model using the computer algebra package [Ryacas](https://cran.r-project.org/package=Ryacas), which interfaces between R and [yacas](http://www.yacas.org/) -- both need
to be installed.

These three steps can be performed with the R commands below. [SBtabVFGEN](https://github.com/icpm-kth/SBtabVFGEN) refers to our external package.

#### Convert SBtab to ODE
The following two commands are unrelated to code generation. They just find and load the biological model.

```{r, label="load the file", eval=FALSE}
 f <- uqsa::uqsa_example("AKAR4cl")    # paths to the TSV files
sb <- SBtabVFGEN::sbtab_from_tsv(f)   # loads the TSV files
```

```{r, eval=TRUE, include=FALSE}
 f <- uqsa::uqsa_example("AKAR4cl")    # paths to the TSV files
sb <- SBtabVFGEN::sbtab_from_tsv(f)   # loads the TSV files
```

We convert the model into an ordinary differential equation (and create intermediate files, various formats).

```{r, eval=TRUE, include=FALSE}
odeModel <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

```{r, eval=FALSE}
odeModel <- SBtabVFGEN::sbtab_to_vfgen(sb)
```

#### Convert the ODE to C code
The function `generateCode` below accepts the return value of
`SBtabVFGEN::sbtab_to_vfgen()` (a list of ode properties) and returns
a character vector of C code:

```{r, eval=TRUE}
C <- uqsa::generateCode(odeModel)
```

This returns the generated code as a character vector, it can be
viewed (and written to a file) using the `cat` function.

The print the first few lines of the generated code (25 lines in the example), you can run the following command. You can use the `file=` argument of `cat` to print it into a file.:

```{r, eval=FALSE}
cat(head(C,25),sep="\n")
```

The following command saves the C source code in the file "AKAR4cl_gvf.c".

```{r, eval=TRUE}
cat(C, sep = "\n", file = "AKAR4cl_gvf.c")
```

The file just created contains the following functions:
```{sh, eval=TRUE, label=sh}
grep AKAR4cl_ ./AKAR4cl_gvf.c
```

#### Convert C code to shared library 
The following command compiles the model (if not yet compiled).
```{r, eval=TRUE}
modelName <- checkModel("AKAR4cl","AKAR4cl_gvf.c")
```

#### (Optional) Generate code for deSolve (ODE solver in R):
It is also possible to create R code, intended for use with [deSolve](https://cran.r-project.org/package=deSolve).
Combining all previous function calls into one (writing to file immediately).

```{r, label="generate R code", eval=TRUE}
rFile <- paste0(comment(odeModel),".R")
cat(uqsa::generateRCode(odeModel),sep="\n",file=rFile)
source(rFile)

# example call:
p <- AKAR4cl_default(0.0)
y0 <- AKAR4cl_init(0.0,p)

print(y0)
```
A test-simulation with deSolve:

```{r simulation, eval=TRUE}
if (require(deSolve)){
    t_ <- seq(0,50,length.out=128)
    y_ <- ode(y0,t_,AKAR4cl_vf,p)
    plot(t_,y_[,1],type='l',main="AKAR4cl",xlab="time")
}
```



# Simulate

We now show how to simulate the model the same conditions as in the experiments (e.g., same initial conditions).

```{r, eval=TRUE}
# import the experimental data
ex <- SBtabVFGEN::sbtab.data(sb,odeModel$conservationLaws)

## we create a closure called "sim" a function that internally remembers the "experiments" (ex)
sim <- simulator.c(ex,modelName)                   # ex holds the instructions for the solver
par <- as.matrix(sb$Parameter[["!DefaultValue"]])  # one column - but can be several sets (as columns)

## sim can be called with any parameter vector - or several and will always simulate the same set of experiments
sr <- sim(par)                                     # simulation results (one per experiment)
stopifnot(length(sr) == length(ex))

i <- seq(70)
t <- ex[[1]]$outputTimes[i]
y <- ex[[1]]$outputValues[[1]][i]
dy <- ex[[1]]$errorValues[[1]][i]

z <- sr[[1]]$func[1,i,1]                           # third dimension is the par column (just one here)

plot(t,z,bty='n',type='l', ylim=c(90,200))
points(t,y)
arrows(t,y,t,y+dy,angle=90, length=0.01)
arrows(t,y,t,y-dy,angle=90, length=0.01)
legend("bottomright",legend=c("simulation","data"), lty=c(1,1), pch=c(NA,1))
```

The simulation result `sr` is a list with several components, including:

* `state`, state variable trajectories 
* `func`, output functions 

The output functions usually correspond to the data (in some way), but
may need some sort of normalization. In this example, the relationship
is direct.

Both `state` and `func` have three indices: 

```{r, state, eval=FALSE}
sr[[l]]$state[i,j,k]        # is a number
```

This corresponds to experiment `l`, state variable `i`, time-point
`j`, and parameter column `k` (relevant when `par` is a matrix
with multiple columns).

---

The previously created `modelName` variable has a comment attribute
(that you can also set yourself, there is no magic there). This
attribute indicates the location of the shared library used for
simulations. With this, ytou don't have to re-make it every time - as long as the
model doesn't change (i.e., `checkModel` does not need to be re-called):

```{r, eval=FALSE}
comment(modelName) <- "./AKAR4cl.so" # this is sufficient if the .so file already exists
```

# Simulations of Several Parameter Sets

The following commands allow you to simulate multiple parameters (saved in the matrix `P`) at the same time.
```{r, sim, eval=TRUE}
t <- as.numeric(ex[[1]]$outputTimes)
np <- NROW(par)
REPS <- 50
P <- matrix(runif(np*REPS,min=0,max=as.numeric(par)*2),np,REPS)
dim(P)


stm <- Sys.time()
sr <- sim(P)
etm <- Sys.time()
difftime(etm,stm)
```

Let's plot all trajectories in one picture:

```{r Plots, eval=TRUE}
T <- rep(c(t,NA),REPS)                                  # the NA value will break the line
Z <- as.numeric(sr[[1]]$func[1,c(seq_along(t),NA),])    # at the end, so it doesn't loop
plot(T,Z,type='l')
```

The new result is (just like the old one) an array with three indexes:

```{r, func, eval=FALSE}
sr[[l]]$func[i,j,k] # a number
```

- `i` output function
- `j` time index
- `k` selects the trajectories that belong to `P[,k]`

---

Clean-up (delete temporary files):

```{r cleanup, eval=TRUE}
gf <- sprintf("AKAR4cl%s",c(".vf",".tar.gz",".zip",".so"))
print(gf)
file.remove(gf)
```

# Other Formats

The SBtab content can be a normal spreadsheet.

If it is a Google spreadsheet (which can be good for collaborative
work), you should write it until fairly satisfied with all components
and then export it as an `.xlsx`, or `.ods` file. The gnumeric application can
automatically convert both `.xlsx` and `.ods` to a set of `.tsv` files if
you want the benefits of version control (which works on text
files). Both `.ods` and `.xlsx` work:

- `SBtabVFGEN::sbtab_from_ods  ` for ODS files
- `SBtabVFGEN::sbtab_from_excel` for XLSX files

Our scripts find the tables by name:

- TSV files: `TableName` attribute is the name of the table
- ODS files: the Sheet's name is the name of the table, `TableName`
  attribute is not parsed
- Excel files: the Sheet's name (same as ODS)


#### Tab Separated Values

For more information on how to solve common tasks regarding tsv files,
see the [tsv topic](./tsv.html). We obtain a list of all tsv files in
the current directory, and then import the contents, like this:

```{r, tsv, eval=FALSE}
modelFiles <- dir(pattern='[.]tsv$')
SBtab <- sbtab_from_tsv(modelFiles)
```

#### Open Document Spreadsheet

These files can be created with Libre Office, Apple's *numbers*
program, gnumeric, and any web-hosted spreadsheet application (like
Google spreadsheets).

```{r, ods, eval=FALSE}
modelFiles <- "DemoModel.ods"
SBtab <- sbtab_from_ods(modelFiles)
```

#### Excel

These can be created with MS Excel, and loaded like this:

```{r, xlsx, eval=FALSE}
modelFiles <- "DemoModel.xlsx"
SBtab <- sbtab_from_excel(modelFiles)
```

# Parameters

The R code generated by `ode.sh` includes a list with the same
functions, but with generic function names:

- `model$vf()` vector field
- `model$jac()` Jacobian
- `model$init()` initial values
- `model$par()` default parameters in linear scale 
    + regardless of what the scale was in the SBtab file
	+ this function returns the long parameter vector (Parameter and Input)

Observe that `model$par()` returns the ODE model parameters, which can
be more than the biological parameters, the components are:

- biological parameters (`SBtab$Parameter`)
- original input parameters (`SBtab$Input`)
- derived input parameters, from conservationlaws (conserved constants)

The inputs are allowed to be different for each experiment (by
definition), the biological parameters are the same for all
experiments simulated in one go. Therefore, we only supply the
biological parameters (the first `n`) to the simulator and it retrieves
the others from the `experiments` variable, by concatenation:

Usually, the number of leading parameters
(sb$Parameter[["!DefaultValue"]]) does not change. This is all the
simulator needs as the inputs are already in `ex`.

