---
title: "Build your own Model"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval=FALSE, cache=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r, setup, eval=TRUE, include=FALSE}
library(uqsa)
Location <- dirname(uqsa_example('AKAR4cl')[1])
```

In this article we will assume that you have a model that you want to
work on and that you have written the model in SBtab form. It may be a
google spreadsheet, an open document spreadsheet (`ods`), MS Excel
file, a file for the `numbers` application on an Apple device or a
`.gnumeric` file.

There are three steps we need to make:

|   | Steps                             | file (out) | tl;dr                |
|--:|:----------------------------------|:----------:|:--------------------:|
| 1 | convert SBtab to ODE              | .vf        | biology => math      |
| 2 | convert the ODE to sources        | .c         | math => code         |
| 3 | convert sources to shared library | .so        | code => machine code |

---

[RPN-derivative](https://github.com/icpm-kth/RPN-derivative)
calculates derivatives or offloads that work on maxima or yacas,
install them if you want to use the `--maxima` or `--yacas` options.


# Load the TSV files Convert and Compile the Model

```{r, eval=FALSE}
library(uqsa)
library(SBtabVFGEN)

# find the tsv files with "?dir", in the directory of your model
f <- dir(Location,pattern='tsv$',full.names=TRUE)
# you don't need this 'Location' variable, just be in the directory where the TSV files are.
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
cl <- SBtabVFGEN::sbtab_to_vfgen(sb)                 # conservation laws, if any
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # includes the data

print(names(sb))

system2("ode", args=c("-C", "--maxima" ,"AKAR4cl.tar.gz"), stdout="AKAR4cl_gvf.c")
stopifnot(file.exists("AKAR4cl_gvf.c"))
modelName <- checkModel("AKAR4cl","AKAR4cl_gvf.c") # compiles the model if not yet compiled
```

This should produce this output:

```{r, eval=TRUE}
library(uqsa)
library(SBtabVFGEN)

# find the tsv files with "?dir", in the directory of your model
f <- dir(Location,pattern='tsv$',full.names=TRUE)
# you don't need this 'Location' variable, just be in the directory where the TSV files are.
sb <- SBtabVFGEN::sbtab_from_tsv(f)                  # a list of data.frames
cl <- SBtabVFGEN::sbtab_to_vfgen(sb)                 # conservation laws, if any
ex <- SBtabVFGEN::sbtab.data(sb,cl)                  # includes the data


system2("ode", args=c("-C", "--maxima" ,"AKAR4cl.tar.gz"), stdout="AKAR4cl_gvf.c")
stopifnot(file.exists("AKAR4cl_gvf.c"))
modelName <- checkModel("AKAR4cl","AKAR4cl_gvf.c") # compiles the model if not yet compiled
```

---

We have created the c source code, with these functions:

```{sh, eval=TRUE}
grep AKAR4cl_ ./AKAR4cl_gvf.c
```

# Simulate

```{r, eval=TRUE}
## we create a closure called "sim" a function that internally remembers the "experiments" (ex)
sim <- simulator.c(ex,modelName)                   # ex holds the instructions for the solver
par <- as.matrix(sb$Parameter[["!DefaultValue"]])  # one column - but can be several sets (as columns)

## sim can be called with any parameter vector - or several and will always simulate the same set of experiments
sr <- sim(par)                                     # simulation results (one per experiment)
stopifnot(length(sr) == length(ex))

i <- seq(70)
t <- ex[[1]]$outputTimes[i]
y <- ex[[1]]$outputValues[[1]][i]
dy <- ex[[1]]$errorValues[[1]][i]

z <- sr[[1]]$func[1,i,1]                           # third dimension is the par column (just one here)

plot(t,z,bty='n',type='l', ylim=c(90,200))
points(t,y)
arrows(t,y,t,y+dy,angle=90, length=0.01)
arrows(t,y,t,y-dy,angle=90, length=0.01)
legend("bottomright",legend=c("simulation","data"), lty=c(1,1), pch=c(NA,1))
```

The simulation result `sr` is a list with several components, at least 2:
- `state` state variable trajectories 
- `func` outout functions 

The output functions usually correspond to the data (in some way), but
may need some sort of normalization. In this case the relationship
is direct.

Both `state` and `func` have three indexes: 

```{r, state, eval=FALSE}
sr[[l]]$state[i,j,k]        # is a number
```

This corresponds to experiment `l`, state variable `i`, time-point
`j`, and parameter column `k` (again, relevant when `par` is a matrix
with multiple columns).

---

The previously created `modelName` variable has a comment attribute
(that you can also set yourself, there is no magic there), this
attribute inidicates the location of the shared library used for
simulations. You don't have to re-make it evry time - as long as the
model doesn't change (`checkModel` does not need to be called):

```{r, eval=FALSE}
comment(modelName) <- "./AKAR4cl.so" # this is sufficient if the .so file already exists
```

# Simulations of Several Parameter Sets

```{r, sim, eval=TRUE}
t <- as.numeric(ex[[1]]$outputTimes)
np <- NROW(par)
REPS <- 50
P <- matrix(runif(np*REPS,min=0,max=as.numeric(par)*2),np,REPS)
dim(P)


stm <- Sys.time()
sr <- sim(P)
etm <- Sys.time()
difftime(etm,stm)
```

let's plot all trajectories in one picture:

```{r Plots, eval=TRUE}
T <- rep(c(t,NA),REPS)                                  # the NA value will break the line
Z <- as.numeric(sr[[1]]$func[1,c(seq_along(t),NA),])    # at the end, so it doesn't loop
plot(T,Z,type='l')
```

The new result is (just like the old one) an array with three indexes:

```{r, func, eval=FALSE}
sr[[l]]$func[i,j,k] # a number
```

- `i` output function
- `j` time index
- `k` selects the trajectories that belong to `P[,k]`

---

cleanup:

```{r cleanup, eval=TRUE}
gf <- sprintf("AKAR4cl%s",c(".vf",".tar.gz",".zip",".so"))
print(gf)
file.remove(gf)
```

# Other Formats

The SBtab content can be a normal spreadsheet.

If it is a google spreadsheet (which can be good for collaborative
work), you should write it until failry satisfied with all components
and then export it as an `.xlsx`, or `.ods` file. The gnumeric application can
automatically convert both `.xlsx` and `.ods` to a set of `.tsv` files if
you want the benefits of version control (which works on text
files). But, `.ods` and `.xlsx` both work:

- `SBtabVFGEN::sbtab_from_ods  ` for ODS files
- `SBtabVFGEN::sbtab_from_excel` for XLSX files

Our scripts find the tables by name:

- TSV files: `TableName` attribute is the name of the table
- ODS files: the Sheet's name is the name of the table, `TableName`
  attribute is not parsed
- Excel files: the Sheet's name (same as ODS)


## Tab Separated Values

For more information on how to solve common tasks regarding tsv files,
see the [tsv topic](./tsv.html). We obtain a list of all tsv files in
the current directory, and then import the contents, like this:

```{r, tsv, eval=FALSE}
modelFiles <- dir(pattern='[.]tsv$')
SBtab <- sbtab_from_tsv(modelFiles)
```

## Open Document Spreadsheet

These files can be created with Libre Office, Apple's *numbers*
program, gnumeric, and any web-hosted spreadsheet application (like
google spreadsheets).

```{r, ods, eval=FALSE}
modelFiles <- "DemoModel.ods"
SBtab <- sbtab_from_ods(modelFiles)
```

## Excel

These can be created with MS Excel, and loaded like this:

```{r, xlsx, eval=FALSE}
modelFiles <- "DemoModel.xlsx"
SBtab <- sbtab_from_excel(modelFiles)
```

### Parameters

The R code generated by `ode.sh` includes a list with the same
functions, but with generic function names:

- `model$vf()` vector field
- `model$jac()` Jacobian
- `model$init()` initial values
- `model$par()` default parameters in linear scale 
    + regardless of what the scale was in the SBtab file
	+ this function returns the long parameter vector (Parameter and Input)

Observe that `model$par()` returns the ODE model parameters, which can
be more than the biological parameters, the components are:

- biological parameters (`SBtab$Parameter`)
- origignal input parameters (`SBtab$Input`)
- derived input parameters, from conservationlaws (conserved constants)

The inputs are allowed to be different for each experiment (by
definition), the biological parameters are the same for all
experiments simulated in one go. Therefore, we only supply the
biological parameters (the first `n`) to the simultor and it retrieves
the others from the `experiments` variable, by concatenation:

Usually, the number of leading parameters
(sb$Parameter[["!DefaultValue"]]) does not change. This is all the
simulator needs as the inputs are already in `ex`.

